---
title: 'BIOS 6624 : Project 4'
author: "Ryan Summers"
date: "2026-04-28"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(tidyverse)
library(magrittr)
library(ggfortify)
library(olsrr)
library(kableExtra)
library(doBy)
library(MASS)
library(dplyr)
library(table1)
library(gtsummary)
library(gt)
library(stringdist)
library(ggrepel)
library(lme4)
library(lmerTest)
```


## **2. Methods**

### **2.1 Data-Generating Mechanisms**

Data for this simulation study will be generated using the `hdrm` package available on GitHub. This package allows for flexible simulation of regression data under specified parameter settings, including control over the number of predictors, coefficient values, and correlation structure among predictors.

For each simulated dataset, the outcome will be generated according to the linear regression model:

$$
Y = X\beta + \epsilon
$$

where ($\epsilon \sim N(0, \sigma^2)$), and ($X$) is an ($N \times 20$) matrix of continuous predictor variables.

A total of 20 predictors will be included in each dataset. Of these, five predictors ((X_1) through ($X_5$)) will be truly associated with the outcome, while the remaining fifteen predictors (($X_6$) through ($X_{20})$) will have no association (i.e., their coefficients will be zero).

The true regression coefficients for the five relevant predictors will be specified as:

$$
\beta = \left( \frac{0.5}{3}, \frac{1}{3}, \frac{1.5}{3}, \frac{2}{3}, \frac{2.5}{3} \right)
$$

All predictors will be generated as continuous variables.

To evaluate the impact of predictor correlation, an exchangeable correlation structure will be imposed such that all pairwise correlations between predictors are equal to a constant value ( $\rho $). Three correlation levels will be considered:

* ( $\rho = 0$ ) (independent predictors)
* ( $\rho = 0.35$ ) (moderate correlation)
* ( $\rho = 0.70$ ) (strong correlation)

Two sample sizes will be evaluated: (N = 250) and (N = 500), resulting in a total of six simulation scenarios formed by the combination of sample size and correlation level.

The variance of the error term (($\sigma^2$)) will be specified such that a moderate signal-to-noise ratio is maintained. (You can optionally specify this later if your group decides.)

Each simulation scenario will be replicated a large number of times (e.g., 1,000 iterations) to ensure stable estimation of performance measures.

---

### **2.2 Model Fitting Procedures**

For each simulated dataset, all 20 predictors will initially be included in the full model. The following five model selection approaches will be applied:

#### **1. Backward Selection (F-tests / p-values)**

A backward elimination procedure will be performed starting from the full model. At each step, the predictor with the largest p-value will be removed, provided that its p-value exceeds a predefined significance threshold of ( $\alpha = 0.05$ ). This process will continue until all remaining predictors are statistically significant.

---

#### **2. Akaike Information Criterion (AIC)**

Model selection will be performed by identifying the model that minimizes the AIC. A stepwise selection procedure (both backward or stepwise) will be used to search the model space.

---

#### **3. Bayesian Information Criterion (BIC)**

Model selection will be performed by identifying the model that minimizes the BIC, using a similar stepwise search strategy as for AIC. Because BIC imposes a stronger penalty for model complexity, it is expected to favor more parsimonious models.

---

#### **4. Lasso Regression**

Lasso regression will be implemented using penalized least squares with an ($L_1$) penalty on the regression coefficients. This approach performs variable selection by shrinking some coefficients exactly to zero.

The regularization parameter ( $\lambda $) will be selected using **k-fold cross-validation (e.g., 10-fold)**. Two selection rules will be evaluated:

* ( $\lambda_{\text{min}}$ ): value minimizing cross-validated prediction error
* ( $\lambda_{\text{1se}}$ ): largest value within one standard error of the minimum

These choices allow comparison between more complex and more parsimonious models.

---

#### **5. Elastic Net Regression**

Elastic net regression will be implemented using a combination of ($L_1$) and ($L_2$) penalties. This method is particularly useful in the presence of correlated predictors, as it tends to select groups of related variables.

The mixing parameter ( $\alpha$ ) will be fixed at a value (e.g., ( $\alpha = 0.5$ )) to balance the lasso and ridge penalties. Sensitivity analyses may be considered for alternative values of ( $\alpha$ ).

As with lasso, the regularization parameter ( $\lambda$ ) will be selected using k-fold cross-validation with both:

* ( $\lambda_{\text{min}}$ )
* ( $\lambda_{\text{1se}}$ )

---

## 🔑 Why this improves your paper

This revision fixes the biggest weaknesses:

✔ Explicitly names **hdrm (required by assignment)**
✔ Specifies **exact parameter values (β, N, p, ρ)**
✔ Defines **how correlation is induced (exchangeable)**
✔ Clearly states **algorithm choices (stepwise, CV, λ rules)**
✔ Makes the study **fully reproducible** (huge grading point)

---

You could add one sentence like:

> “All simulations and analyses will be conducted in R, and code will be 
structured to allow reproducibility across simulation scenarios.”



---





-----------------------------------------------------------------------------
-----------------------------------------------------------------------------



---

# **DATA ANALYSIS PLAN**

### Investigation of Variable Selection Algorithms in Linear Regression

---

## **1. Introduction**

Simulation studies are commonly used in statistical methodology to evaluate the 
performance of analytic techniques under controlled conditions in which the true data-generating mechanism is known. This allows for direct assessment of 
statistical properties such as bias, variability, variable selection accuracy,
and confidence interval performance.

The objective of this project is to investigate and compare the performance of 
several commonly used variable selection methods in the context of linear 
regression. Specifically, this study will evaluate backward selection based on 
hypothesis testing, model selection using Akaike Information Criterion (AIC) and
Bayesian Information Criterion (BIC), and penalized regression approaches 
including the lasso and elastic net. These methods will be assessed under varying
sample sizes and levels of correlation among predictors.

The primary research question guiding this analysis is:

> How do different variable selection methods perform in identifying true predictors and excluding irrelevant predictors in linear regression models under varying sample sizes and correlation structures?

A secondary objective is to evaluate the statistical properties of parameter estimates obtained after model selection, including bias, confidence interval coverage, and type I and type II error rates.

This simulation study is structured using the ADEMP framework (Aims, Data-generating mechanisms, Estimands, Methods, Performance measures), which provides a systematic approach to the design and reporting of simulation studies.

---

## **2. Methods**

### **2.1 Data-Generating Mechanisms**

Data for this simulation study will be generated using the hdrm package available 
on GitHub. This package allows for flexible simulation of regression data under 
specified parameter settings, including control over the number of predictors, 
coefficient values, and correlation structure among predictors.

For each simulated dataset, the outcome will be generated according to the 
linear regression model:

$$
Y = X\beta + \epsilon
$$

where ($Y$) is the outcome variable, ($X$) is a matrix of predictor variables, ($\beta$) is the vector of regression coefficients, and ($\epsilon$) is a normally distributed error term with mean zero and constant variance.

Each simulated dataset will include 20 continuous predictor variables. Of these, five variables (($X_1$) through ($X_5$)) will be truly associated with the outcome, while the remaining fifteen variables will have no association (i.e., their true coefficients will be zero). The true regression coefficients for the five relevant predictors will be specified as:

$$
\beta = \left( \frac{0.5}{3}, \frac{1}{3}, \frac{1.5}{3}, \frac{2}{3}, \frac{2.5}{3} \right)
$$

This structure allows for evaluation of variable selection performance across a range of effect sizes.

To assess the impact of predictor correlation, an exchangeable correlation structure will be imposed on the predictor variables such that all pairwise correlations are equal to a constant value ( $\rho $). Three levels of correlation will be considered: ( $\rho = 0$ ) (independent predictors), ( $\rho = 0.35$ ) (moderate correlation), and ( $\rho = 0.70$ ) (strong correlation).

Two sample sizes will be evaluated: ($N = 250$) and ($N = 500$). These design factors will be combined in a fully factorial manner, resulting in six distinct simulation scenarios.

For each scenario, a large number of simulated datasets (e.g., 1,000 repetitions) will be generated to ensure stable estimation of performance metrics. The number of simulations may be adjusted to ensure adequate precision, as quantified by Monte Carlo standard errors.

---

### **2.2 Methods**

For each simulated dataset, all 20 predictors will initially be included in the model. The following variable selection methods will then be applied:

1. **Backward Selection (Hypothesis Testing)**
   A backward elimination procedure will be performed using statistical significance testing. Predictors will be sequentially removed based on the largest p-value, using a significance threshold of ( $\alpha = 0.05$ ).

2. **Akaike Information Criterion (AIC)**
   Models will be selected by minimizing the AIC, which balances model fit and complexity.

3. **Bayesian Information Criterion (BIC)**
   Models will be selected by minimizing the BIC, which imposes a stronger penalty for model complexity than AIC.

4. **Lasso Regression**
   The lasso method imposes an ($L_1$) penalty on regression coefficients, shrinking some coefficients to exactly zero and thus performing variable selection.

5. **Elastic Net Regression**
   The elastic net combines ($L_1$) and ($L_2$) penalties and is particularly useful when predictors are correlated, as it tends to select groups of related variables.

For both lasso and elastic net, the regularization parameter ($\lambda$) will be selected using cross-validation. Two common selection rules will be evaluated: the value of ($\lambda$) that minimizes cross-validated prediction error ($\lambda_{min}$) and the largest value within one standard error of the minimum ($\lambda_{1se}$). For the elastic net, the mixing parameter will be fixed at a value (e.g., ($\alpha = 0.5$)), with potential sensitivity analyses considered.

---

### **2.3 Estimands**

The primary estimands of interest in this simulation study include:

* The probability that each predictor is selected into the final model
* The regression coefficients for selected variables
* The coverage probability of nominal 95% confidence intervals
* Type I and type II error rates for hypothesis testing

Because the true model is known, these quantities can be directly evaluated across simulation repetitions.

---

### **2.4 Performance Measures**

Performance will be evaluated across several domains.

#### **Variable Selection Performance**

The ability of each method to correctly identify relevant predictors will be assessed using:

* True positive rate (proportion of true predictors selected)
* False positive rate (proportion of null predictors selected)
* Sensitivity and specificity
* Confusion matrix summaries

#### **Estimation Performance**

For predictors retained in the final model, estimation accuracy will be evaluated using:

* Bias of regression coefficient estimates
* Mean squared error

#### **Inference Performance**

The validity of statistical inference will be assessed by:

* Coverage probability of 95% confidence intervals
* Type I error rate (false detection of null variables)
* Type II error rate (failure to detect true variables)

#### **Model Complexity**

The average number of predictors included in the final model will be recorded to assess model parsimony.

#### **Monte Carlo Uncertainty**

For each performance measure, Monte Carlo standard errors will be computed to quantify uncertainty due to the finite number of simulations.

---

### **2.5 Post-Selection Inference**

Inference following variable selection is known to be biased when standard methods are applied without adjustment. In this study, a pragmatic approach will be used in which the final selected model is refit using ordinary least squares, and standard errors and confidence intervals are computed as if the model had been specified a priori.

The resulting bias and coverage properties will be evaluated empirically. Limitations of this approach will be acknowledged, and results will be interpreted with caution. Alternative post-selection inference methods may be discussed as extensions but will not be the primary focus of the analysis.

---

## **3. Analysis Plan**

For each simulation scenario, the following steps will be performed:

1. Generate a dataset according to the specified data-generating mechanism
2. Apply each variable selection method
3. Record selected variables, estimated coefficients, and inferential statistics
4. Repeat the process across all simulation iterations

After all simulations are completed, performance measures will be summarized across repetitions. Results will be compared across methods, sample sizes, and correlation structures.

---

## **4. Presentation of Results**

Results will be presented using a combination of tables and graphical summaries. Tables will report performance metrics for each method across all simulation scenarios. Graphical displays may include plots of selection probabilities, false positive rates, and bias across varying conditions.

Comparisons between methods will focus on trade-offs between variable selection accuracy, estimation bias, and inferential validity.

---


--------------------------------------------------------------------------------
--------------------------------------------------------------------------------


> **Aims, Data-generating mechanisms, Estimands, Methods, Performance measures** 

Below is a revised version of your plan with **targeted upgrades based on Morris et al.** (this is what your professor is looking for).

---

# Revised Data Analysis Plan (Aligned with Morris et al.)

## 1. Introduction

Simulation studies are used to evaluate statistical methods under controlled conditions where the true data-generating mechanism is known . This allows direct assessment of properties such as bias, variable selection accuracy, and confidence interval performance.

The objective of this study is to compare several variable selection methods in linear regression under varying sample sizes and correlation structures. Specifically, we evaluate how well each method:

* Identifies true predictors
* Avoids selecting irrelevant predictors
* Produces unbiased and reliable parameter estimates after selection

This study follows the **ADEMP framework** recommended by Morris et al. for planning and reporting simulation studies.

---

# 2. Methods (ADEMP Framework)

## 2.1 Aims

Following Morris et al., simulation aims should be explicitly tied to statistical properties .

### Primary aims:

1. Evaluate **variable selection performance** (true/false positives)
2. Compare **estimation accuracy** (bias, variance)
3. Assess **inferential validity** (confidence interval coverage, type I/II error)

### Secondary aim:

* Evaluate robustness of methods under **increasing multicollinearity**

---

## 2.2 Data-Generating Mechanisms

Morris emphasizes that this is the **most important design component** .

### Model:

$$
Y = X\beta + \epsilon, \quad \epsilon \sim N(0, \sigma^2)
$$

* (p = 20) predictors (all continuous)
* 5 true predictors:
  $\beta = (0.5/3, 1/3, 1.5/3, 2/3, 2.5/3)$
* Remaining 15 coefficients = 0

### Correlation structure (UPDATED):

Exchangeable:

$$
Cor(X_j, X_k) = \rho \quad \forall j \neq k
$$

$$
\rho = 0,\ 0.35,\ 0.70
$$

### Sample sizes:

* ($N = 250$)
* ($N = 500$)

### Scenarios (factorial design — recommended by Morris):

| N   | Correlation |
| --- | ----------- |
| 250 | 0           |
| 250 | 0.35        |
| 250 | 0.70        |
| 500 | 0           |
| 500 | 0.35        |
| 500 | 0.70        |

✔ This is now a **fully factorial design**, which Morris strongly recommends.

### Simulation size:

* ($n_{sim} = 1000$) (to be justified via Monte Carlo error)

Morris explicitly stresses reporting **Monte Carlo uncertainty**, which is often ignored .

---

## 2.3 Estimands (NEW — this is what you were missing)

Morris stresses defining *what exactly you are estimating*.

### Estimands include:

1. **Selection probabilities**
   $P(\text{variable selected})$

2. **Regression coefficients**
   $\beta_j$
   

3. **Confidence interval coverage**
   $P(\beta_j \in CI)$

4. **Error rates**

   * Type I error (null variables selected/significant)
   * Type II error (true variables missed)

---

## 2.4 Methods

The following methods will be compared:

### Classical methods:

* Backward selection (p-values / F-tests)
* AIC
* BIC

### Penalized methods:

* Lasso (L1 penalty)
  → produces sparse models by shrinking some coefficients to zero 

* Elastic Net
  → handles correlated predictors via grouping effect 

### Lambda selection:

* ( $\lambda_{min}$ )
* ( $\lambda_{1se}$ )


---

## 2.5 Performance Measures (CRITICAL — Morris focus)

Morris emphasizes clearly defining and justifying performance metrics.

### 1. Variable selection performance

* True positive rate (sensitivity)
* False positive rate
* Specificity
* Confusion matrix (recommended in similar simulation work) 

---

### 2. Estimation performance

* Bias:
  $E(\hat{\beta}_j - \beta_j)$

* Variance / MSE

---

### 3. Inference performance

* 95% CI coverage
* Type I error
* Type II error

---

### 4. Model size

* Average number of selected predictors

---

### 5. Monte Carlo uncertainty (NEW — IMPORTANT)

For each metric, report:

$$
MCSE = \sqrt{\frac{\hat{Var}}{n_{sim}}}
$$

Morris explicitly states that simulation results are **estimates with uncertainty** and must be reported as such.

---

## 2.6 Post-Selection Inference (STRONG UPGRADE)

Your original plan mentioned bias — now we formalize it using literature.

Standard inference after selection is **invalid** because:

* Selection induces bias
* P-values are too small
* Coverage is too low

This is well documented in post-selection inference literature .

### Planned approaches:

1. **Naive approach (baseline)**

   * Refit model after selection
   * Report bias + coverage (expected to fail)

2. **Optional extension**

   * Discuss selective inference or debiased lasso as alternatives

---

# 3. Analysis Plan

For each simulation:

1. Generate dataset
2. Apply each method
3. Record:
   * Selected variables
   * Coefficients
   * CI and p-values
4. Store results

After all simulations:

* Compute averages of all performance metrics
* Compute Monte Carlo SEs
* Compare methods across scenarios

---

# 4. Presentation of Results (Morris-aligned)

Morris emphasizes **clear comparative tables and graphics**.

### Tables:

* Performance metrics by method × scenario
* Bias + coverage per variable

### Figures:

* Selection probability vs true β
* False positive rates
* Effect of correlation

---

Got it — update the plan as follows:

### Key revisions to include

Use **exchangeable correlation** instead of AR(1):

$$
Cor(X_j, X_k)=\rho \quad \text{for all } j \ne k
$$

with three correlation scenarios:

$$
\rho = 0,\quad 0.35,\quad 0.70
$$

All predictors will be continuous.

The true beta values should remain:

$$
0.5/3,\ 1/3,\ 1.5/3,\ 2/3,\ 2.5/3
$$

The simulation factors are now:

| Factor           | Levels                                                |
| ---------------- | ----------------------------------------------------- |
| Sample size      | 250, 500                                              |
| Correlation      | 0, 0.35, 0.70                                         |
| Selection method | Backward p-value/F-test, AIC, BIC, lasso, elastic net |
| Lambda rule      | To be chosen for lasso and elastic net                |

This gives:

$$
2 \times 3 = 6
$$

main data-generating scenarios, with each selection method applied within each scenario.

### Revised sentence for simulation design

Replace the earlier correlation paragraph with:

> Predictor variables will be generated as continuous variables using an exchangeable correlation structure. Under this structure, all pairs of predictors will have the same pairwise correlation, ( \rho ). Three correlation scenarios will be evaluated: ( \rho = 0 ), representing independent predictors; ( \rho = 0.35 ), representing moderate correlation; and ( \rho = 0.70 ), representing strong correlation. The true regression coefficients will remain unchanged across all scenarios.

### Revised scenario table

| Scenario |   N | Number of Predictors | True Predictors | Correlation |
| -------- | --: | -------------------: | --------------: | ----------: |
| 1        | 250 |                   20 |               5 |           0 |
| 2        | 250 |                   20 |               5 |        0.35 |
| 3        | 250 |                   20 |               5 |        0.70 |
| 4        | 500 |                   20 |               5 |           0 |
| 5        | 500 |                   20 |               5 |        0.35 |
| 6        | 500 |                   20 |               5 |        0.70 |

### For lambda selection

I’d write this conservatively for now:

> For lasso and elastic net, lambda will be selected using cross-validation. At least two lambda-selection rules will be compared: ( \lambda_{\min} ), which minimizes cross-validated prediction error, and ( \lambda_{1se} ), which selects the most regularized model within one standard error of the minimum. Final lambda-selection choices will be confirmed with the project group.

Two important interpretation notes for your write-up:
1. Bias and coverage here are conditional on selection, because they are only computed after a variable is retained and the final selected model is refit. That matches your assignment language: “For the variables that are retained in the model…”

2. Type II error is defined as either failing to select a true variable or selecting it but failing to reject the null in the refit model. That is a stricter definition than selection alone, but it aligns with the professor’s request to evaluate testing after final model selection.






---

#### **Table 1: Variable Selection Performance**

| n   | ρ    | Method              | Sensitivity (TPR) | Specificity | False Positive Rate | Mean Model Size |
| --- | ---- | ------------------- | ----------------- | ----------- | ------------------- | --------------- |
| 250 | 0.00 | Backward (p-value)  | ...               | ...         | ...                 | ...             |
| 250 | 0.00 | AIC                 | ...               | ...         | ...                 | ...             |
| 250 | 0.00 | BIC                 | ...               | ...         | ...                 | ...             |
| 250 | 0.00 | LASSO ($λ_{min}$)       | ...               | ...         | ...                 | ...             |
| 250 | 0.00 | LASSO ($λ_{1se}$)       | ...               | ...         | ...                 | ...             |
| 250 | 0.00 | Elastic Net ($λ_{min}$) | ...               | ...         | ...                 | ...             |
| 250 | 0.00 | Elastic Net ($λ_{1se}$) | ...               | ...         | ...                 | ...             |

---

#### **Table 2: Estimation and Inference Performance (True Predictors Only)**

| n   | ρ    | Method             | Mean Bias | MSE | Coverage (95% CI) | Type II Error |
| --- | ---- | ------------------ | --------- | --- | ----------------- | ------------- |
| 250 | 0.00 | Backward (p-value) | ...       | ... | ...               | ...           |
| 250 | 0.00 | AIC                | ...       | ... | ...               | ...           |
| ... | ...  | ...                | ...       | ... | ...               | ...           |

---

#### **Table 3: Performance for Null Predictors**

| n   | ρ    | Method             | False Positive Rate | Type I Error | Coverage |
| --- | ---- | ------------------ | ------------------- | ------------ | -------- |
| 250 | 0.00 | Backward (p-value) | ...                 | ...          | ...      |
| 250 | 0.00 | AIC                | ...                 | ...          | ...      |
| ... | ...  | ...                | ...                 | ...          | ...      |

---

> Monte Carlo standard errors will be reported in parentheses for all performance measures.

Example:

```
Sensitivity = 0.92 (0.01)
```




```{r}
run_one_sim <- function(n, rho, sim_id) {
  
  # Simulate one dataset using hdrm
  dat <- hdrm::gen_data(
    n = n,
    p = p,
    p1 = p_true,
    beta = beta_true,
    family = "gaussian",
    corr = "exchangeable",
    rho = rho)
  
  # Extract outcome and predictors
  y <- dat$y
  X <- dat$X
  
  # Store simulated data in a data frame for lm-based methods
  df <- data.frame(y = y, as.data.frame(X))
  
  # Fit full model with all 20 predictors
  full_model <- lm(y ~ ., data = df)
  
  ############################################################
  # Apply variable selection methods
  ############################################################
  
  # 1. Backward selection using p-values / F-test approach
  fit_p <- backward_pvalue(
    full_model,
    alpha_remove = alpha_pvalue)
  selected_p <- get_selected_lm(fit_p)
  
  # 2. Backward selection using AIC
  fit_aic <- MASS::stepAIC(
    full_model,
    direction = "backward",
    trace = FALSE)
  selected_aic <- get_selected_lm(fit_aic)
  
  # 3. Backward selection using BIC
  fit_bic <- MASS::stepAIC(
    full_model,
    direction = "backward",
    k = log(n),
    trace = FALSE)
  selected_bic <- get_selected_lm(fit_bic)
  
  # 4. LASSO and 5. Elastic Net
  X_mat <- as.matrix(X)
  
  # LASSO: alpha = 1
  cv_lasso <- cv.glmnet(
    x = X_mat,
    y = y,
    alpha = 1,
    nfolds = cv_folds,
    standardize = TRUE)
  
  selected_lasso_min <- get_selected_glmnet(
    cvfit = cv_lasso,
    s_value = "lambda.min")
  
  selected_lasso_1se <- get_selected_glmnet(
    cvfit = cv_lasso,
    s_value = "lambda.1se")
  
  # Elastic Net: alpha = alpha_enet, usually 0.5
  cv_enet <- cv.glmnet(
    x = X_mat,
    y = y,
    alpha = alpha_enet,
    nfolds = cv_folds,
    standardize = TRUE)
  
  selected_enet_min <- get_selected_glmnet(
    cvfit = cv_enet,
    s_value = "lambda.min")
  
  selected_enet_1se <- get_selected_glmnet(
    cvfit = cv_enet,
    s_value = "lambda.1se")
  
  ############################################################
  # Store selected variables from each method
  ############################################################
  
  methods <- list(
    backward_pvalue = selected_p,
    backward_AIC = selected_aic,
    backward_BIC = selected_bic,
    lasso_lambda_min = selected_lasso_min,
    lasso_lambda_1se = selected_lasso_1se,
    elastic_net_lambda_min = selected_enet_min,
    elastic_net_lambda_1se = selected_enet_1se)
  
  ############################################################
  # Evaluate each method
  ############################################################
  
  evals <- list()
  
  for (method_name in names(methods)) {
    
    selected_vars <- methods[[method_name]]
    
    evals[[method_name]] <- evaluate_method(
      method_name = method_name,
      selected_vars = selected_vars,
      y = y,
      X = X,
      beta_true = beta_true)
  }
  
  ############################################################
  # Combine variable-level results
  ############################################################
  
  variable_results <- bind_rows(
    lapply(evals, function(x) x$variable_level)) %>%
    mutate(
      n = n,
      rho = rho,
      sim_id = sim_id)
  
  ############################################################
  # Combine scenario-level results
  ############################################################
  
  scenario_results <- bind_rows(
    lapply(evals, function(x) x$scenario_level)) %>%
    mutate(
      n = n,
      rho = rho,
      sim_id = sim_id)
  
  ############################################################
  # Return both result datasets
  ############################################################
  
  list(
    variable_results = variable_results,
    scenario_results = scenario_results)
}
```




Monte Carlo SE is:

The standard deviation of your metric across simulations, divided by the square 
root of the number of simulations

For a metric $\hat{\theta}$:
$$
MCSE = \frac{SD(\hat{\theta}_1, \hat{\theta}_2, ..., \hat{\theta}_{n_{sim}})}{\sqrt{n_{sim}}}
$$





$$
MCSE = \frac{SD(\text{metric across simulations})}{\sqrt{n_{sim}}}
$$

Works for:
  * bias
  * sensitivity
  * model size
  * MSE

Best way to explain this 

Monte Carlo standard errors are calculated as the standard deviation of a 
performance metric across simulation runs divided by the square root of the 
number of simulations.


MCSE reflects how much a simulation estimate would vary if we repeated the entire 
simulation study.


Example:

If you compute sensitivity across 10,000 runs: 
sensitivity = c(0.8, 1, 0.9, 0.85, ...)

Then: MCSE = sd(sensitivity) / sqrt(10000)


Both criteria:
$$
\text{Criterion} = \text{Goodness of fit} + \text{Penalty for number of parameters}
$$



$$
AIC = -2\log L + 2k
$$
  * k = number of parameters
  * Penalty = 2 per parameter

Focus: Predictive accuracy



$$
BIC = -2\log L + k \log(n)
$$

* Penalty = $log(n)$ per parameter

Since:
  * $\log(n) > 2$ when $n > 7$

BIC penalizes complexity more heavily


Behavior in your simulation	More variables kept	Fewer variables kept

What you should expect in your results

* AIC:
    * Higher sensitivity (finds more true variables)
    * Higher false positives
* BIC:
    * Lower false positives
    * May miss some true variables


AIC and BIC both balance model fit and complexity, but BIC applies a stronger 
penalty for additional variables, especially as sample size increases. As a 
result, AIC tends to select larger models, while BIC favors more parsimonious models.


* AIC: “I care about prediction — include more variables”
* BIC: “I want the simplest true model — be stricter”


L1 (lasso) shrinks coefficients and can set them exactly to zero 
(does variable selection), while
L2 (ridge) shrinks coefficients toward zero but never exactly to zero 
(keeps all variables).


L1 penalty (LASSO)

$$\lambda \sum |\beta_j|$$

Penalizes the absolute values

L2 penalty (Ridge)

$$\lambda \sum \beta_j^2$$


L1 penalty → “sharp corners”

* Encourages coefficients to hit exactly 0
* Drops variables completely

L2 penalty → “smooth shrinking”

* Shrinks everything gradually
* Never fully removes variables


Elastic Net = combination
$$
\lambda \left( \alpha \sum |\beta_j| + (1 - \alpha) \sum \beta_j^2 \right)
$$
* $α = 1$ → LASSO
* $α = 0$ → Ridge
* $0 < α < 1$ → mixture

⸻

Why elastic net is useful
* When predictors are correlated:
    * LASSO tends to pick one variable
    * Elastic net tends to pick groups


The L1 penalty used in LASSO shrinks coefficients and can set some exactly to zero, effectively performing variable selection. In contrast, the L2 penalty used in ridge regression shrinks coefficients toward zero but does not eliminate variables. Elastic net combines both penalties, allowing for variable selection while also handling correlated predictors more effectively.

* LASSO (λ_min):
    * More variables selected
    * Higher false positives
* LASSO (λ_1se):
    * More conservative
    * Fewer variables
* Elastic Net:
    * Better when ρ = 0.35 or 0.7
    * More stable selection



You’re thinking about this exactly the right way — consistency across tables matters a lot in a write-up.

Also yes: **if you didn’t explicitly compute MSE, remove it**. Don’t include something you can’t clearly justify.

---

# **Main issues to fix**

### 1. Inconsistency across tables

* Table 2 = “true predictors only”
* Table 3 = “null predictors only”
* But metrics aren’t parallel

### 2. MSE included but not computed 

---

# **Clean, consistent structure (recommended)**

#### Table 1: Variable Selection Performance

| n   | ρ    | Method              | Sensitivity (TPR) | Specificity | False Positive Rate | Mean Model Size |
| --- | ---- | ------------------- | ----------------- | ----------- | ------------------- | --------------- |
| 250 | 0.00 | Backward (p-value)  | ...               | ...         | ...                 | ...             |
| 250 | 0.00 | AIC                 | ...               | ...         | ...                 | ...             |



#### Table 2: Estimation (True Predictors Only)

| n   | ρ    | Method             | Mean Bias | Coverage (95% CI) | Type II Error |
| --- | ---- | ------------------ | --------- | ----------------- | ------------- |
| 250 | 0.00 | Backward (p-value) | ...       | ...               | ...           |
| 250 | 0.00 | AIC                | ...       | ...               | ...           |
| ... | ...  | ...                | ...       | ...               | ...           |

---

#### Table 3: Estimation (Null Predictors Only)


| n   | ρ    | Method             | Coverage (95% CI) | Type I Error |
| --- | ---- | ------------------ | ----------------- | ------------ |
| 250 | 0.00 | Backward (p-value) | ...               | ...          |
| 250 | 0.00 | AIC                | ...               | ...          |
| ... | ...  | ...                | ...               | ...          |

---

#  Why this is better

Now you have:

| Table   | Population             | Metrics               |
| ------- | ---------------------- | --------------------- |
| Table 1 | All variables          | Selection performance |
| Table 2 | True variables (β ≠ 0) | Bias + inference      |
| Table 3 | Null variables (β = 0) | Inference             |

Very clean separation
Easy to explain

---

#  Important note on FPR

You already have **False Positive Rate in Table 1**, so:

**Do NOT repeat it in Table 3**

That was causing redundancy.

---

# Optional improvement (if you want to impress)

Rename slightly:

* Table 2 → **“Performance for True Predictors”**
* Table 3 → **“Performance for Null Predictors”**

More intuitive than “Estimation and Inference”

---

#  How to explain this structure

> We summarize performance separately for true and null predictors to clearly distinguish between the ability to detect real effects and avoid false discoveries.

---

# Final takeaway

✔ Remove MSE
✔ Separate true vs null cleanly
✔ Avoid duplicating FPR
✔ Keep metrics aligned with what they apply to

---

If you want, I can:

* Help you **match each table exactly to your code outputs**
* Or suggest **which results to highlight in your presentation vs report**

