---
title: 'Project 4 : Simulation'
author: "Ryan Summers"
date: "2026-05-06"
output: html_document
---

```{r setup, include=FALSE}
library(hdrm)
library(glmnet)
library(MASS)
library(dplyr)
library(tibble)
library(future)
library(furrr)
library(progress)
library(progressr)
library(qs)
library(ggplot2)
library(future.apply)
library(gt)
library(gtsummary)
library(tidyr)
library(here)
```

```{r}
# load objects
scenario_results <- readRDS(file.path(
  "/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs",
  "GitHub/BIOS6624-Class/Project4/DataProcessed/scenario_results.rds"))
variable_results <- readRDS(file.path(
  "/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs",
  "GitHub/BIOS6624-Class/Project4/DataProcessed/variable_results.rds"))
```

```{r}
# # reproducibility checks - results from Rscript run from terminal
# scenario_results <- readRDS(file.path(
#   "/Users/ryan_summers/DataProcessed/scenario_results.rds"))
# variable_results <- readRDS(file.path(
#   "/Users/ryan_summers/DataProcessed/variable_results.rds"))
```


## Simulation Settings

```{r}
set.seed(2522)
n_sim <- 10000
n_values <- c(250, 500)
rho_values <- c(0, 0.35, 0.70)

p <- 20
p_true <- 5
true_vars <- paste0("V", sprintf("%02d", 1:p_true))

beta_true <- c(
  0.5/3,
  1/3,
  1.5/3,
  2/3,
  2.5/3,
  rep(0, 15))

names(beta_true) <- paste0("V", sprintf("%02d", 1:p))

alpha_pvalue <- 0.10 # change this to 0.10 - stated to be similar to AIC @ 0.157
alpha_enet <- 0.5 # perhaps test multiple alphas??? c(0.25, 0.5, 0.75)
cv_folds <- 10
```

### Replication Justification

```{r}
# MCSE for binary performance measure
mcse_bin <- function(p, nsim) {
  sqrt(p * (1 - p) / nsim)
  }

mcse_bin(0.50, 10000)  # worst case
mcse_bin(0.95, 10000)  # nominal coverage
```

> If the true probability were 0.50, the estimated proportion from the sim would
vary by about ±0.005 due to Monte Carlo sampling error.
>
> For coverage probability near 95%, the sim estimate is expected to vary by
about ±0.002.




## Helper Functions

```{r}
backward_pvalue <- function(full_model, alpha_remove = 0.157) {
  current_model <- full_model
  
  repeat {
    pvals <- summary(current_model)$coefficients[-1, 4]
    
    if (length(pvals) == 0) break
    
    max_p <- max(pvals, na.rm = TRUE)
    
    if (max_p <= alpha_remove) break
    
    remove_var <- names(which.max(pvals))
    current_terms <- attr(terms(current_model), "term.labels")
    updated_terms <- setdiff(current_terms, remove_var)
    
    if (length(updated_terms) == 0) {
      current_model <- lm(
        formula = as.formula("y ~ 1"),
        data = model.frame(current_model))
      break
      }
    
    new_formula <- as.formula(
      paste("y ~", paste(updated_terms, collapse = " + ")))
    
    current_model <- lm(new_formula, data = model.frame(current_model))
    }
  
  current_model
  }

get_selected_lm <- function(model) {
  setdiff(names(coef(model)), "(Intercept)")
  }

get_selected_glmnet <- function(cvfit, s_value) {
  b <- coef(cvfit, s = s_value)
  selected <- rownames(b)[as.vector(b != 0)]
  setdiff(selected, "(Intercept)")
  }

post_selection_refit <- function(y, X, selected_vars) {
  df <- data.frame(y = y, X)
  
  if (length(selected_vars) == 0) {
    return(tibble(
      variable = character(),
      estimate = numeric(),
      p_value = numeric(),
      ci_low = numeric(),
      ci_high = numeric()))
    }
  
  form <- as.formula(paste("y ~", paste(selected_vars, collapse = " + ")))
  fit <- lm(form, data = df)
  
  coefs <- summary(fit)$coefficients
  cis <- confint(fit)
  
  tibble(
    variable = rownames(coefs)[-1],
    estimate = coefs[-1, "Estimate"],
    p_value = coefs[-1, "Pr(>|t|)"],
    ci_low = cis[-1, 1],
    ci_high = cis[-1, 2])
  }

evaluate_method <- function(
    method_name, 
    selected_vars, 
    y, 
    X, 
    beta_true, 
    alpha_test = 0.05) {
  
  all_vars <- names(beta_true)
  true_nonzero <- names(beta_true[beta_true != 0])
  true_null <- names(beta_true[beta_true == 0])
  
  refit <- post_selection_refit(y, X, selected_vars)
  
  variable_level <- tibble(variable = all_vars) %>%
    mutate(
      method = method_name,
      # convert selected variable names into a logical indicator for all variables
      # TRUE = selected by method, FALSE = not selected -> used to compute 
      # TP,FP,etc
      selected = variable %in% selected_vars,
      beta_true = beta_true[variable]) %>%
    left_join(refit, by = "variable") %>%
    ## this coverage calculation is saying, "“If we fail to select a true 
    #  variable, it counts as a failure. Per Carter Sevick
    mutate(
      estimate_final = ifelse(selected, estimate, 0),
      bias = estimate_final - beta_true,
      covered = case_when(
        selected ~ (ci_low <= beta_true & ci_high >= beta_true),
        !selected & beta_true == 0 ~ TRUE,
        !selected & beta_true != 0 ~ FALSE),
      
    ## coverage is conditional on selection for the below code
    # mutate(
    #   bias = ifelse(selected, estimate - beta_true, NA_real_),
    #   covered = ifelse(
    #     selected,
    #     # **REVIEW** - not sure if this captures coverage correctly...
    #     ci_low <= beta_true & ci_high >= beta_true, NA),
      
      significant = ifelse(selected, p_value < alpha_test, FALSE),
    # β=0 AND selected AND significant
      type1_error = beta_true == 0 & selected & significant,
    # β≠0 AND not selected OR not significant
      # missed detection can happen by not selecting or finding significance
      type2_error = beta_true != 0 & (!selected | !significant))
  
  ######--- Confusion matrix ---######
  
  # number truly associated variables (beta ≠ 0) that were correctly selected
  TP <- sum(variable_level$selected & variable_level$beta_true != 0)
  # number of null variables (beta = 0) that were incorrectly selected
  FP <- sum(variable_level$selected & variable_level$beta_true == 0)
  # nmumber of truly associated variables (beta ≠ 0) that were NOT selected
  FN <- sum(!variable_level$selected & variable_level$beta_true != 0)
  # of null variables (beta = 0) that were correctly NOT selected
  TN <- sum(!variable_level$selected & variable_level$beta_true == 0)
  
  scenario_level <- tibble(
    method = method_name,
    TP = TP,
    FP = FP,
    FN = FN,
    TN = TN,
    sensitivity = TP / (TP + FN),
    specificity = TN / (TN + FP),
    false_positive_rate = FP / (FP + TN),
    false_negative_rate = FN / (FN + TP),
    model_size = length(selected_vars))
  
  list(
    variable_level = variable_level,
    scenario_level = scenario_level)
  }

mcse_mean <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) <= 1) return(NA_real_)
  sd(x) / sqrt(length(x))
  }

mcse_prop <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_real_)
  p_hat <- mean(x)
  sqrt(p_hat * (1 - p_hat) / length(x))
  }


```

## Run Single Replication Function

```{r}
run_single_rep <- function(n, rho, sim_id) {
  # simulates one dataset using hdrm
  dat <- hdrm::gen_data(
    n = n,
    p = p,
    p1 = p_true, # 5 true predictors
    beta = beta_true,
    family = "gaussian",
    corr = "exchangeable",
    rho = rho)
  
  # extract X and y from dat
  y <- dat$y
  X <- dat$X
  
  # store simulated data into df (optional, for backward selection)
  df <- data.frame(y = y, as.data.frame(X))
  # includes all 20 predictors
  full_model <- lm(y ~ ., data = df)
 
  ## Apply variable selection methods ##
  
  # backward selection using p-value / F-test 
  fit_p <- backward_pvalue(full_model, alpha_remove = alpha_pvalue)
  selected_p <- get_selected_lm(fit_p)
  
  # backward selection using AIC
  fit_aic <- MASS::stepAIC(full_model, direction = "backward", trace = FALSE)
  # chooses the model that minimizes AIC 
  selected_aic <- get_selected_lm(fit_aic)
  
  # ackward selection using BIC
  fit_bic <- MASS::stepAIC(
    full_model,
    direction = "backward",
    k = log(n),
    trace = FALSE)
  # chooses the model that minimizes BIC 
  selected_bic <- get_selected_lm(fit_bic)
  
  ######---Penalized Regression---######
  X_mat <- as.matrix(X)
  
  # LASSO
  cv_lasso <- cv.glmnet(
    x = X_mat,
    y = y,
    alpha = 1,
    nfolds = cv_folds,
    standardize = TRUE)
  
  # Elastic Net
  cv_enet <- cv.glmnet(
    x = X_mat,
    y = y,
    alpha = alpha_enet,
    nfolds = cv_folds,
    standardize = TRUE)
  
  selected_lasso_min <- get_selected_glmnet(cv_lasso, "lambda.min")
  selected_lasso_1se <- get_selected_glmnet(cv_lasso, "lambda.1se")
  
  selected_enet_min <- get_selected_glmnet(cv_enet, "lambda.min")
  selected_enet_1se <- get_selected_glmnet(cv_enet, "lambda.1se")
  
  # store selected variables from each method
  methods <- list(
    backward_pvalue = selected_p,
    backward_AIC = selected_aic,
    backward_BIC = selected_bic,
    lasso_lambda_min = selected_lasso_min,
    lasso_lambda_1se = selected_lasso_1se,
    elastic_net_lambda_min = selected_enet_min,
    elastic_net_lambda_1se = selected_enet_1se)
  
  ######---Evaluation of each method---######
  evals <- list()
  
  for (method_name in names(methods)) {
    selected_vars <- methods[[method_name]]
    evals[[method_name]] <- evaluate_method(
      method_name = method_name,
      selected_vars = selected_vars,
      y = y,
      X = X,
      beta_true = beta_true)
    }
  
  variable_results <- bind_rows(
    lapply(evals, function(x) x$variable_level)) %>%
    mutate(n = n, rho = rho, sim_id = sim_id)
  
  scenario_results <- bind_rows(
    lapply(evals, function(x) x$scenario_level)) %>%
    mutate(n = n, rho = rho, sim_id = sim_id)
  
  list(
    variable_results = variable_results, 
    scenario_results = scenario_results)
  }
  
```

## Main simulation loop w/ Parallelization

```{r}
t0 <- Sys.time()

handlers(global = TRUE)
handlers("progress") # or "txtprogressbar"

sim_grid <- expand.grid(
  n = n_values,
  rho = rho_values,
  sim_id = 1:n_sim)

# 18 cores but R crashes with 15 or more
plan(multisession, workers = parallel::detectCores() - 6)

with_progress({
  # progress bar
  prog <- progressor(along = seq_len(nrow(sim_grid)))
  # for each row of sim_grid, run one simulation replication.
  all_results <- future_lapply(
    seq_len(nrow(sim_grid)), # nrow(sim_grid)
    function(i) {
      prog()
      run_single_rep(
        n = as.numeric(sim_grid$n[i]), #as.numeric(sim_grid$n[i]),
        rho = as.numeric(sim_grid$rho[i]), # as.numeric(sim_grid$rho[i]),
        sim_id = as.integer(sim_grid$sim_id[i])) #as.integer(sim_grid$sim_id[i]))
      },
    future.seed = 2522) # review this******
  })

# simple run
with_progress({
  # progress bar
  prog <- progressor(along = seq_len(nrow(sim_grid)))
  # for each row of sim_grid, run one simulation replication.
  all_results <- future_lapply(
    seq_len(100), # nrow(sim_grid)
    function(i) {
      prog()
      run_single_rep(
        n = as.numeric(100), 
        rho = as.numeric(0.75), 
        sim_id = as.integer(1)) 
    },
    future.seed = 2522) 
})

plan(sequential)

variable_results <- bind_rows(
  lapply(all_results, function(x) x$variable_results))

scenario_results <- bind_rows(
  lapply(all_results, function(x) x$scenario_results))

# save - rather large file so use seriallizatin
qsave(
  all_results, 
  file.path(
    "/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs",
    "GitHub/BIOS6624-Class/Project4/DataProcessed/all_results.qs"), 
  preset = "fast", 
  nthreads = 8)

saveRDS(variable_results, "variable_results.rds")
saveRDS(scenario_results, "scenario_results.rds")

t1 <- Sys.time()

message("Runtime: ", 
        round(as.numeric(difftime(t1, t0, units = "mins")), 2), 
        " minutes")
```


```{r}
# mean bias under Null and True vars
variable_results %>%
  mutate(var_type = ifelse(beta_true == 0, "Null", "True")) %>%
  group_by(var_type) %>%
  summarise(mean_bias = mean(bias, na.rm = TRUE) , .groups = "drop")

# Null bias relationship (in aggregate to just get a feel)
variable_results %>%
  filter(beta_true == 0) %>%
  summarise(
    total_null = n(),
    zero_bias_count = sum(bias == 0, na.rm = TRUE),
    prop_zero_bias = mean(bias == 0, na.rm = TRUE)) # proportion selected too
```


## Results


### Method & Variable Summaries

```{r}
# Method-level summary: selection accuracy and model size
method_summary <- scenario_results %>%
  group_by(n, rho, method) %>%
  summarise(
    # mean number of TPs across simulations
    mean_TP = mean(TP, na.rm = TRUE),
    # Monte Carlo SE (MCSE) of mean TP - uses variability across sim runs
    mcse_TP = mcse_mean(TP),
    # mean number of FPs
    mean_FP = mean(FP, na.rm = TRUE),
    # MCSE of mean FP
    mcse_FP = mcse_mean(FP),
    # avg TPR (sensitivity) across ims - prop of true vars correctly selected
    mean_sensitivity = mean(sensitivity, na.rm = TRUE),
    # MCSE of sens
    mcse_sensitivity = mcse_mean(sensitivity),
    #avg specificity (TNR) across sims - prop of null vars correctly not selected 
    mean_specificity = mean(specificity, na.rm = TRUE),
    # MCSE of spec
    mcse_specificity = mcse_mean(specificity),
    # avg FPR across sims - prop of null vars incorrectly selected
    mean_false_positive_rate = mean(false_positive_rate, na.rm = TRUE),
    mcse_false_positive_rate = mcse_mean(false_positive_rate),
    # avg number of variables selected
    mean_model_size = mean(model_size, na.rm = TRUE),
    # MCSE of size
    mcse_model_size = mcse_mean(model_size),
    .groups = "drop")

# Variable-level summary: selection probability, bias, coverage, type I/II
variable_summary <- variable_results %>%
  group_by(n, rho, method, variable, beta_true) %>%
  summarise(
    # prob that a given variable is selected by the method across sim runs
    selection_probability = mean(selected),
    mcse_selection_probability = mcse_prop(selected),
    
    mean_bias = mean(bias, na.rm = TRUE),
    mcse_bias = mcse_mean(bias),
    
    mse = mean(bias^2, na.rm = TRUE),
    mcse_mse = mcse_mean(bias^2),
    
    coverage = mean(covered, na.rm = TRUE),
    mcse_coverage = mcse_prop(covered),
    # Type I error: only defined for the 15 null variables
    # Pr(reject H0 | beta = 0)
    type1_error_rate = ifelse(
      unique(beta_true) == 0,
      mean(type1_error), # median??
      NA_real_),
    mcse_type1_error = ifelse(
      unique(beta_true) == 0,
      mcse_prop(type1_error),
      NA_real_),
    
    # Type II error: only defined for the 5 true variables
    # Pr(fail to reject H0 | beta != 0)
    type2_error_rate = ifelse(
      unique(beta_true) != 0,
      mean(type2_error),
      NA_real_),
    mcse_type2_error = ifelse(
      unique(beta_true) != 0,
      mcse_prop(type2_error),
      NA_real_),
    .groups = "drop")

```


### Selection Performance

```{r label = "Table used in final report (Table 1)"}
selection_table_wide <- method_summary %>%
  mutate(
    Method = recode(
      method,
      backward_pvalue = "Backward (p-value)",
      backward_AIC = "AIC",
      backward_BIC = "BIC",
      lasso_lambda_min = "LASSO (λₘᵢₙ)",
      lasso_lambda_1se = "LASSO (λ₁ₛₑ)",
      elastic_net_lambda_min = "ElNet (λₘᵢₙ)",
      elastic_net_lambda_1se = "ElNet (λ₁ₛₑ)"),
    rho_label = case_when(
      rho == 0 ~ "rho0",
      rho == 0.35 ~ "rho035",
      rho == 0.7 ~ "rho07"
    )) %>%
  select(Method, rho_label, n, TPR = mean_sensitivity,
         FPR = mean_false_positive_rate, Size = mean_model_size) %>%
  pivot_wider(
    names_from = c(rho_label, n),
    values_from = c(TPR, FPR, Size),
    names_glue = "{rho_label}_n{n}_{.value}") %>%
  select(
    Method,
    rho0_n250_TPR, rho0_n250_FPR, rho0_n250_Size,
    rho0_n500_TPR, rho0_n500_FPR, rho0_n500_Size,
    rho035_n250_TPR, rho035_n250_FPR, rho035_n250_Size,
    rho035_n500_TPR, rho035_n500_FPR, rho035_n500_Size,
    rho07_n250_TPR, rho07_n250_FPR, rho07_n250_Size,
    rho07_n500_TPR, rho07_n500_FPR, rho07_n500_Size)

rho0_cols <- c(
  "rho0_n250_TPR", "rho0_n250_FPR", "rho0_n250_Size",
  "rho0_n500_TPR", "rho0_n500_FPR", "rho0_n500_Size")

rho035_cols <- c(
  "rho035_n250_TPR", "rho035_n250_FPR", "rho035_n250_Size",
  "rho035_n500_TPR", "rho035_n500_FPR", "rho035_n500_Size")

rho07_cols <- c(
  "rho07_n250_TPR", "rho07_n250_FPR", "rho07_n250_Size",
  "rho07_n500_TPR", "rho07_n500_FPR", "rho07_n500_Size")

selection_table_wide %>%
  gt(rowname_col = "Method") %>%
  tab_header(
    title = "Variable Selection Performance") %>%
  
  # add sample-size spanners FIRST
  tab_spanner(label = "n = 250",id = "rho0_n250",
      columns = c(rho0_n250_TPR, rho0_n250_FPR, rho0_n250_Size)) %>%
  tab_spanner(label = "n = 500",id = "rho0_n500",
      columns = c(rho0_n500_TPR, rho0_n500_FPR, rho0_n500_Size)) %>%
  tab_spanner(label = "n = 250",id = "rho035_n250",
      columns = c(rho035_n250_TPR, rho035_n250_FPR, rho035_n250_Size)) %>%
  tab_spanner(label = "n = 500",id = "rho035_n500",
      columns = c(rho035_n500_TPR, rho035_n500_FPR, rho035_n500_Size)) %>%
  tab_spanner(label = "n = 250", id = "rho07_n250",
      columns = c(rho07_n250_TPR, rho07_n250_FPR, rho07_n250_Size)) %>%
  tab_spanner(
    label = "n = 500",
    id = "rho07_n500",
    columns = c(rho07_n500_TPR, rho07_n500_FPR, rho07_n500_Size)) %>%
  
  # correlation spanners SECOND so they sit on top
  tab_spanner(label = "ρ = 0",id = "rho0",
              columns = all_of(rho0_cols)) %>%
  tab_spanner(label = "ρ = 0.35",id = "rho035",
              columns = all_of(rho035_cols)) %>%
  tab_spanner(label = "ρ = 0.7",id = "rho07",
              columns = all_of(rho07_cols)) %>%
  cols_label(
    .list = setNames(
      rep(c("TPR", "FPR", "Size"), 6),
      names(selection_table_wide)[-1]
    )) %>%
  fmt_number(columns = contains("TPR") | contains("FPR"),decimals = 2) %>%
  fmt_number(columns = contains("Size"),decimals = 1) %>% 
  tab_style(
  style = cell_borders(
    sides = "left",
    color = "gray33",
    weight = px(3)),
  locations = cells_body(
    columns = c(
      rho0_n250_TPR,
      rho0_n500_TPR,
      rho035_n250_TPR,
      rho035_n500_TPR,
      rho07_n250_TPR,
      rho07_n500_TPR
      ))) 
```

For supplement section
```{r label = "Figure used in final report (Supplementary Figure 1)"}
selection_heatmap_data <- variable_summary %>%
  filter(beta_true != 0) %>%
  mutate(
    Method = recode(
      method,
      backward_pvalue = "Backward (p-value)",
      backward_AIC = "AIC",
      backward_BIC = "BIC",
      lasso_lambda_min = "LASSO (min)",
      lasso_lambda_1se = "LASSO (1se)",
      elastic_net_lambda_min = "ElNet (min)",
      elastic_net_lambda_1se = "ElNet (1se)"))

ggplot(
  selection_heatmap_data,
  aes(
    x = variable,
    y = Method,
    fill = selection_probability
  )) +
  geom_tile(color = "white") +
  geom_text(
    aes(
      label = round(selection_probability, 3),
      color = selection_probability < 0.75),
    size = 5) +
  facet_grid(n ~ rho) +
  scale_fill_gradientn(
    colours = c(
      "#B2182B",
      "#EF8A62",
      "#FEE0D2",
      "white",
      "#D9F0D3",
      "#1B7837" ),
    values = scales::rescale(
      c(0.20, 0.50, 0.75, 0.90, 0.97, 1.00),
      from = c(0.20, 1.00)),
    limits = c(0.20, 1.00),
    name = "Selection\nProbability") +
  scale_color_manual(
    values = c(
      "TRUE" = "white",
      "FALSE" = "black"),
    guide = "none") +
  labs(
    x = "True Predictor",
    y = "Method",
    title = "Selection Probability for True Predictors") +
  theme_bw(base_size=13) +
  theme(
    strip.background = element_rect(fill = "white"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom")
```


### Estimation

#### Bias

```{r label = "Figure used in final report (Figure 1)"}

bias_heatmap_data <- variable_summary %>%
  filter(beta_true != 0) %>%
  mutate(
    Method = recode(
      method,
      backward_pvalue = "Backward p-value",
      backward_AIC = "AIC",
      backward_BIC = "BIC",
      lasso_lambda_min = "LASSO (min)",
      lasso_lambda_1se = "LASSO (1se)",
      elastic_net_lambda_min = "ElNet (min)",
      elastic_net_lambda_1se = "ElNet (1se)"),
    scenario = paste0("n=", n, ", ρ=", rho))

ggplot(
  bias_heatmap_data,
  aes(x = variable, y = Method, fill = mean_bias)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(mean_bias, 2)),size = 5) +
  facet_grid(n ~ rho) +
  scale_fill_gradient2(
    breaks = c(-0.09, 0, 0.02),
    low = "#4575B4",
    mid = "white",
    high = "#D73027",
    midpoint = 0,
    name = "Mean Bias") +
  labs(
    x = "True Predictor",
    y = "Method",
    title = "Mean Bias for True Predictors") +
  theme_bw(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "white"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom")
```



#### Type I & II

```{r}
# Type I error plotting
type1_summary <- variable_results %>%
  filter(beta_true == 0) %>%
  group_by(n, rho, method) %>%
  summarise(
    type1_error_rate = mean(type1_error, na.rm = TRUE), 
    mcse_type1_error = mcse_prop(type1_error),
    .groups = "drop")

ggplot(type1_summary,
       aes(x = method, y = type1_error_rate, fill = method)) +
  geom_col() +
  facet_grid(n ~ rho) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "black") +
  scale_x_discrete(labels = c(
    backward_AIC = "AIC",
    backward_BIC = "BIC",
    backward_pvalue = "p-value",
    lasso_lambda_min = expression("LASSO (" * lambda[min] * ")"),
    lasso_lambda_1se = expression("LASSO (" * lambda[1*se] * ")"),
    elastic_net_lambda_min = expression("ElNet (" * lambda[min] * ")"),
    elastic_net_lambda_1se = expression("ElNet (" * lambda[1*se] * ")"))) +
  theme_bw() +
  labs(
    x = "Method",
    y = "Type I Error Rate",
    title = "Type I Error Rate Among Null Variables") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none")
```


```{r}
# Type I boxplots
ggplot(variable_summary %>% filter(beta_true == 0),
       aes(x = method, y = type1_error_rate, fill = method)) +
  geom_boxplot() +
  facet_grid(n ~ rho) +
  scale_x_discrete(labels = c(
    backward_AIC = "AIC",
    backward_BIC = "BIC",
    backward_pvalue = "p-value",
    lasso_lambda_min = expression("LASSO (" * lambda[min] * ")"),
    lasso_lambda_1se = expression("LASSO (" * lambda[1*se] * ")"),
    elastic_net_lambda_min = expression("ElNet (" * lambda[min] * ")"),
    elastic_net_lambda_1se = expression("ElNet (" * lambda[1*se] * ")"))) +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```


**Tables:**

```{r}
error_table <- variable_summary %>%
  group_by(n, rho, method) %>%
  summarise(
    `Type I Error (Null)` = mean(type1_error_rate[beta_true == 0], 
                                 na.rm = TRUE),
    `Type II Error (True)` = mean(type2_error_rate[beta_true != 0], 
                                  na.rm = TRUE),
    .groups = "drop") %>%
  mutate(
    Method = recode(
      method,
      backward_pvalue = "Backward (p-value)",
      backward_AIC = "AIC",
      backward_BIC = "BIC",
      lasso_lambda_min = "LASSO (min)",
      lasso_lambda_1se = "LASSO (1se)",
      elastic_net_lambda_min = "ELNET (min)",
      elastic_net_lambda_1se = "ELNET (1se)"
      )) %>%
  select(n, rho, Method, `Type I Error (Null)`, `Type II Error (True)`) %>% 
  gt() %>%
  tab_header(title = "Table: Type I & II Error Rates") %>%
  fmt_number(
    columns = c(`Type I Error (Null)`, `Type II Error (True)`),
    decimals = 3)

error_table_wide <- variable_summary %>%
  group_by(n, rho, method) %>%
  summarise(
    type1 = mean(type1_error_rate[beta_true == 0], na.rm = TRUE),
    type2 = mean(type2_error_rate[beta_true != 0], na.rm = TRUE),
    .groups = "drop") %>%
  mutate(
    Method = recode(
      method,
      backward_pvalue = "Backward (p-value)",
      backward_AIC = "AIC",
      backward_BIC = "BIC",
      lasso_lambda_min = "LASSO (λ<sub>min</sub>)", #(λₘᵢₙ) ; (λ₁ₛₑ)"
      lasso_lambda_1se = "LASSO (λ<sub>1se</sub>)",
      elastic_net_lambda_min = "ElNet (λ<sub>min</sub>)",
      elastic_net_lambda_1se = "ElNet (λ<sub>1se</sub>)"),
    Correlation = case_when(
      rho == 0.0 ~ "ρ = 0 (Independent)",
      rho == 0.35 ~ "ρ = 0.35 (Moderate)",
      rho == 0.7 ~ "ρ = 0.7 (Strong)")) %>% 
  select(n, Correlation, Method, type1, type2) %>%
  pivot_wider(
    names_from = n,
    values_from = c(type1, type2),
    names_glue = "{.value}_{n}") %>%
  arrange(Correlation, Method)
```


```{r}
# Type I & II as spanners
error_table_wide %>%
  gt(
    rowname_col = "Method",
    groupname_col = "Correlation") %>%
  fmt_markdown(columns = Method) %>%
  tab_header(
    title = "Type I & II Error Rates") %>%
  # First level: Error type
  tab_spanner(
    label = "Type I (Null)",
    columns = c(type1_250, type1_500)) %>%
  tab_spanner(
    label = "Type II (True)",
    columns = c(type2_250, type2_500)) %>%
  # Column labels underneath
  cols_label(
    type1_250 = "n = 250",
    type1_500 = "n = 500",
    type2_250 = "n = 250",
    type2_500 = "n = 500") %>%
  fmt_number(
    columns = starts_with("type"),
    decimals = 3)
```

```{r label = "Table used in final report (Table 2)"}
error_table <- variable_summary %>%
  group_by(n, rho, method) %>%
  summarise(
    type1 = mean(type1_error_rate[beta_true == 0], na.rm = TRUE),
    type2 = mean(type2_error_rate[beta_true != 0], na.rm = TRUE),
    .groups = "drop") %>%
  mutate(
    Method = recode(
      method,
      backward_pvalue = "Backward (p-value)",
      backward_AIC = "AIC",
      backward_BIC = "BIC",
      lasso_lambda_min = "LASSO (λₘᵢₙ)",
      lasso_lambda_1se = "LASSO (λ₁ₛₑ)",
      elastic_net_lambda_min = "ElNet (λₘᵢₙ)",
      elastic_net_lambda_1se = "ElNet (λ₁ₛₑ)"),
    rho_label = case_when(
      rho == 0 ~ "rho0",
      rho == 0.35 ~ "rho035",
      rho == 0.7 ~ "rho07"
    )) %>%
  select(Method, rho_label, n, type1, type2) %>%
  pivot_wider(
    names_from = c(rho_label, n),
    values_from = c(type1, type2),
    names_glue = "{rho_label}_n{n}_{.value}") %>%
  select(
    Method,
    rho0_n250_type1, rho0_n250_type2,
    rho0_n500_type1, rho0_n500_type2,
    rho035_n250_type1, rho035_n250_type2,
    rho035_n500_type1, rho035_n500_type2,
    rho07_n250_type1, rho07_n250_type2,
    rho07_n500_type1, rho07_n500_type2)


error_table %>%
  gt(rowname_col = "Method") %>%
  tab_header(title = "Type I and Type II Error Rates") %>%
  tab_spanner(label = "n = 250",id = "rho0_n250",
    columns = c(rho0_n250_type1, rho0_n250_type2)) %>%
  tab_spanner(label = "n = 500",id = "rho0_n500",
    columns = c(rho0_n500_type1, rho0_n500_type2)) %>%
  tab_spanner(label = "n = 250",id = "rho035_n250",
    columns = c(rho035_n250_type1, rho035_n250_type2)) %>%
  tab_spanner(label = "n = 500",id = "rho035_n500",
    columns = c(rho035_n500_type1, rho035_n500_type2)) %>%
  tab_spanner(label = "n = 250",id = "rho07_n250",
    columns = c(rho07_n250_type1, rho07_n250_type2)) %>%
  tab_spanner(label = "n = 500",id = "rho07_n500",
    columns = c(rho07_n500_type1, rho07_n500_type2)) %>%
  tab_spanner(label = "ρ = 0",id = "rho0",
    columns = c(
      rho0_n250_type1, rho0_n250_type2,
      rho0_n500_type1, rho0_n500_type2
    )) %>%
  tab_spanner(label = "ρ = 0.35",id = "rho035",
    columns = c(
      rho035_n250_type1, rho035_n250_type2,
      rho035_n500_type1, rho035_n500_type2
    )) %>%
  tab_spanner(label = "ρ = 0.7",id = "rho07",
    columns = c(
      rho07_n250_type1, rho07_n250_type2,
      rho07_n500_type1, rho07_n500_type2
    )) %>%
  cols_label(
    rho0_n250_type1 = "Type I",
    rho0_n250_type2 = "Type II",
    rho0_n500_type1 = "Type I",
    rho0_n500_type2 = "Type II",
    rho035_n250_type1 = "Type I",
    rho035_n250_type2 = "Type II",
    rho035_n500_type1 = "Type I",
    rho035_n500_type2 = "Type II",
    rho07_n250_type1 = "Type I",
    rho07_n250_type2 = "Type II",
    rho07_n500_type1 = "Type I",
    rho07_n500_type2 = "Type II") %>%
  fmt_number(columns = starts_with("rho"),decimals = 3) %>% 
  tab_style(
  style = cell_borders(
    sides = "left",
    color = "gray33",
    weight = px(3)),
  locations = cells_body(
    columns = c(
      rho0_n250_type1,
      rho0_n500_type1,
      rho035_n250_type1,
      rho035_n500_type1,
      rho07_n250_type1,
      rho07_n500_type1
      ))) 

```


#### Coverage

```{r label = "Figure used in final report (Figure 3)"}
variable_summary %>%
  filter(beta_true != 0) %>%
  mutate(
    Method = recode(
      method,
      backward_pvalue = "Backward p-value",
      backward_AIC = "AIC",
      backward_BIC = "BIC",
      lasso_lambda_min = "LASSO (min)",
      lasso_lambda_1se = "LASSO (1se)",
      elastic_net_lambda_min = "ElNet (min)",
      elastic_net_lambda_1se = "ElNet (1se)"),
    scenario = paste0("n=", n, ", ρ=", rho)) %>% 

ggplot(aes(x = variable, y = Method, fill = coverage)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(coverage, 3), color=coverage<0.8), size = 5)+
  scale_color_manual(
    values = c("TRUE" = "white","FALSE" = "black"),guide = "none")+
  facet_grid(n ~ rho) +
  # scale_fill_gradient2(
  #   low = "#D73027",
  #   midpoint = 0.95,
  #   high = "#009E73",
  #   name = "Coverage") +
  scale_fill_gradientn(
  colours = c(
    "#67001F",  
    "#B2182B",
    "#F4A582",
    "#FEEBE2",  
    "white",
    "#D9F0D3",
    "#1B7837"),
  values = scales::rescale(
    c(0.20, 0.60, 0.85, 0.93, 0.95, 0.97, 1.00), #c(0.20, 0.50, 0.90, 0.95, 0.97, 1.00)
    from = c(0.20, 1.00)),
  limits = c(0.20, 1.00),
  name = "Coverage") +
  labs(
    x = "True Predictor",
    y = "Method",
    title = "Coverage of 95% CIs for True Predictors") +
  theme_bw(base_size=14) +
  theme(
    strip.background = element_rect(fill = "white"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom")
```


```{r}

true_coverage_wide <- variable_summary %>%
  filter(beta_true != 0) %>%
  mutate(
    Method = recode(
      method,
      backward_pvalue = "Backward (p-value)",
      backward_AIC = "AIC",
      backward_BIC = "BIC",
      lasso_lambda_min = "LASSO (λₘᵢₙ)",
      lasso_lambda_1se = "LASSO (λ₁ₛₑ)",
      elastic_net_lambda_min = "ElNet (λₘᵢₙ)",
      elastic_net_lambda_1se = "ElNet (λ₁ₛₑ)"),
    Correlation = case_when(
      rho == 0 ~ "ρ = 0 (Independent)",
      rho == 0.35 ~ "ρ = 0.35 (Moderate)",
      rho == 0.7 ~ "ρ = 0.7 (Strong)"
    )) %>%
  select(Correlation, Method, n, variable, coverage) %>%
  pivot_wider(
    names_from = c(n, variable),
    values_from = coverage,
    names_glue = "n{n}_{variable}") %>%
  arrange(Correlation, Method)

true_coverage_wide %>%
  gt(rowname_col = "Method",groupname_col = "Correlation") %>%
  tab_header(
    title = "Coverage of 95% Confidence Intervals for True Predictors") %>%
  tab_spanner(
    label = "n = 250",
    columns = c(n250_V01, n250_V02, n250_V03, n250_V04, n250_V05)) %>%
  tab_spanner(
    label = "n = 500",
    columns = c(n500_V01, n500_V02, n500_V03, n500_V04, n500_V05)) %>%
  cols_label(
    n250_V01 = "V01",n250_V02 = "V02",n250_V03 = "V03",n250_V04 = "V04",
    n250_V05 = "V05",
    n500_V01 = "V01",n500_V02 = "V02",n500_V03 = "V03",n500_V04 = "V04",
    n500_V05 = "V05") %>%
  fmt_number(columns = starts_with("n"),decimals = 3) %>% 
  tab_style(
    style = cell_borders(
      sides = "left",
      color = "gray33",
      weight = px(3)),
    locations = cells_body(
      columns = n500_V01
    )) %>% 
  # highlight cells that didn't meet coverage criteria
  data_color(columns = starts_with("n"), fn = function(x) {
    ifelse(x <= 0.94, "#FDE0DD", "white")
    })
  
```

### MCSE Summaries

```{r}
# summarize MCSEs to determine meaningful differences in non-coverage
variable_summary %>%
  summarise(
    min_mcse = min(mcse_coverage, na.rm = TRUE),
    mean_mcse = mean(mcse_coverage, na.rm = TRUE),
    median_mcse = median(mcse_coverage, na.rm = TRUE),
    max_mcse = max(mcse_coverage, na.rm = TRUE))

variable_summary %>%
  mutate(
    deviation = 0.95 - coverage, 
    z_mcse = deviation / mcse_coverage) %>% 
  arrange(desc(z_mcse)) %>% 
  select(n,rho, method, variable, deviation, z_mcse)
```

```{r}
mcse_table_wide <- method_summary %>%
  mutate(
    Method = recode(
      method,
      backward_pvalue = "Backward (p-value)",
      backward_AIC = "AIC",
      backward_BIC = "BIC",
      lasso_lambda_min = "LASSO (min)",
      lasso_lambda_1se = "LASSO (1se)",
      elastic_net_lambda_min = "ElNet (min)",
      elastic_net_lambda_1se = "ElNet (1se)"),
    rho_label = case_when(
      rho == 0 ~ "rho0",
      rho == 0.35 ~ "rho035",
      rho == 0.7 ~ "rho07"
    )) %>%
  select(
    Method,
    rho_label,
    n,
    TPR = mcse_sensitivity,
    FPR = mcse_false_positive_rate,
    Size = mcse_model_size) %>%
  pivot_wider(
    names_from = c(rho_label, n),
    values_from = c(TPR, FPR, Size),
    names_glue = "{rho_label}_n{n}_{.value}") %>%
  select(
    Method,
    rho0_n250_TPR, rho0_n250_FPR, rho0_n250_Size,
    rho0_n500_TPR, rho0_n500_FPR, rho0_n500_Size,
    rho035_n250_TPR, rho035_n250_FPR, rho035_n250_Size,
    rho035_n500_TPR, rho035_n500_FPR, rho035_n500_Size,
    rho07_n250_TPR, rho07_n250_FPR, rho07_n250_Size,
    rho07_n500_TPR, rho07_n500_FPR, rho07_n500_Size)

mcse_table_wide %>%
  gt(rowname_col = "Method") %>%
  tab_header(
    title = "Monte Carlo Standard Errors for Selection Performance") %>%
  
  tab_spanner(label = "n = 250", id = "rho0_n250",
          columns = c(rho0_n250_TPR, rho0_n250_FPR, rho0_n250_Size)) %>%
  tab_spanner(label = "n = 500", id = "rho0_n500",
          columns = c(rho0_n500_TPR, rho0_n500_FPR, rho0_n500_Size)) %>%
  
  tab_spanner(label = "n = 250", id = "rho035_n250",
          columns = c(rho035_n250_TPR, rho035_n250_FPR, rho035_n250_Size)) %>%
  tab_spanner(label = "n = 500", id = "rho035_n500",
          columns = c(rho035_n500_TPR, rho035_n500_FPR, rho035_n500_Size)) %>%
  
  tab_spanner(label = "n = 250", id = "rho07_n250",
          columns = c(rho07_n250_TPR, rho07_n250_FPR, rho07_n250_Size)) %>%
  tab_spanner(label = "n = 500", id = "rho07_n500",
          columns = c(rho07_n500_TPR, rho07_n500_FPR, rho07_n500_Size)) %>%
  
  tab_spanner(label = "ρ = 0", id = "rho0",
              columns = c(
                rho0_n250_TPR, rho0_n250_FPR, rho0_n250_Size,
                rho0_n500_TPR, rho0_n500_FPR, rho0_n500_Size
              )) %>%
  tab_spanner(label = "ρ = 0.35", id = "rho035",
              columns = c(
                rho035_n250_TPR, rho035_n250_FPR, rho035_n250_Size,
                rho035_n500_TPR, rho035_n500_FPR, rho035_n500_Size
              )) %>%
  tab_spanner(label = "ρ = 0.7", id = "rho07",
              columns = c(
                rho07_n250_TPR, rho07_n250_FPR, rho07_n250_Size,
                rho07_n500_TPR, rho07_n500_FPR, rho07_n500_Size
              )) %>%
  
  cols_label(
    rho0_n250_TPR = "TPR",  rho0_n250_FPR = "FPR",  rho0_n250_Size = "Size",
    rho0_n500_TPR = "TPR",  rho0_n500_FPR = "FPR",  rho0_n500_Size = "Size",
    
    rho035_n250_TPR = "TPR", rho035_n250_FPR = "FPR", rho035_n250_Size = "Size",
    rho035_n500_TPR = "TPR", rho035_n500_FPR = "FPR", rho035_n500_Size = "Size",
    
    rho07_n250_TPR = "TPR", rho07_n250_FPR = "FPR", rho07_n250_Size = "Size",
    rho07_n500_TPR = "TPR", rho07_n500_FPR = "FPR", rho07_n500_Size = "Size") %>%
  
  fmt_number(columns = contains("TPR") | contains("FPR"),decimals = 4) %>%
  fmt_number(columns = contains("Size"),decimals = 4) %>%
  
  tab_style(
    style = cell_borders(
      sides = "left",
      color = "gray33",
      weight = px(3)),
    locations = cells_body(
      columns = c(
        rho0_n250_TPR,
        rho0_n500_TPR,
        rho035_n250_TPR,
        rho035_n500_TPR,
        rho07_n250_TPR,
        rho07_n500_TPR
      ))) %>%
  tab_style(
    style = cell_borders(
      sides = "left",
      color = "gray33",
      weight = px(3)),
    locations = cells_column_labels(
      columns = c(
        rho0_n250_TPR,
        rho0_n500_TPR,
        rho035_n250_TPR,
        rho035_n500_TPR,
        rho07_n250_TPR,
        rho07_n500_TPR
      ))) %>% 
  tab_style(style = cell_text(size = px(18)),locations = cells_body())
```

```{r label = "Figure used in final report (Supplementary Figure 2)"}
mcse_heatmap_data <- method_summary %>%
  mutate(
    Method = recode(
      method,
      backward_pvalue = "Backward (p-value)",
      backward_AIC = "AIC",
      backward_BIC = "BIC",
      lasso_lambda_min = "LASSO (min)",
      lasso_lambda_1se = "LASSO (1se)",
      elastic_net_lambda_min = "ElNet (min)",
      elastic_net_lambda_1se = "ElNet (1se)"
    )) %>%
  select(n,rho,Method,
         `TPR` = mcse_sensitivity,
         `FPR` = mcse_false_positive_rate,
         `Model Size` = mcse_model_size) %>%
  pivot_longer(
    cols = c(`TPR`, `FPR`, `Model Size`),
    names_to = "metric",
    values_to = "mcse")

ggplot(
  mcse_heatmap_data,
  aes(x = factor(rho), y = Method, fill = mcse)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(mcse, 4),color = mcse > 0.02),size = 5) +
  facet_grid(metric ~ n) +
  scale_fill_gradient(
    low = "white",
    high = "#B2182B",
    name = "MCSE") +
  scale_color_manual(
    values = c("TRUE" = "white","FALSE" = "black"),guide = "none") +
  labs(
    x = "Correlation (ρ)",
    y = "Method",
    title = "Monte Carlo Standard Errors for Selection Performance") +
  theme_bw(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "white"),
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1))
```




```{r label = "Figure used in final report (Supplementary Figure 3)"}
mcse_inference_table <- variable_summary %>%
  group_by(n, rho, method) %>%
  summarise(
    mcse_bias = mean(mcse_bias[beta_true != 0], na.rm = TRUE),
    mcse_coverage = mean(mcse_coverage[beta_true != 0], 
                         na.rm = TRUE),
    mcse_type1 = mean(mcse_type1_error[beta_true == 0], 
                      na.rm = TRUE),
    mcse_type2 = mean(mcse_type2_error[beta_true != 0], 
                      na.rm = TRUE),
    .groups = "drop")

mcse_inference_heatmap <- mcse_inference_table %>%
  mutate(
    Method = recode(
      method,
      backward_pvalue = "Backward (p-value)",
      backward_AIC = "AIC",
      backward_BIC = "BIC",
      lasso_lambda_min = "LASSO (min)",
      lasso_lambda_1se = "LASSO (1se)",
      elastic_net_lambda_min = "ElNet (min)",
      elastic_net_lambda_1se = "ElNet (1se)"
    )) %>%
  pivot_longer(
    cols = c(
      mcse_bias,
      mcse_coverage,
      mcse_type2,
      mcse_type1),
    names_to = "metric",
    values_to = "mcse") %>%
  mutate(
    metric = recode(
      metric,
      mcse_bias = "Bias",
      mcse_coverage = "Coverage",
      mcse_type2 = "Type II",
      mcse_type1 = "Type I"
    ))

ggplot(
  mcse_inference_heatmap,
  aes(x = factor(rho), y = Method, fill = mcse)) +
  geom_tile(color = "white") +
  geom_text(
    aes(label = round(mcse, 4),color = mcse > 0.002),size = 5) +
  facet_grid(metric ~ n) +
  scale_fill_gradient(
    low = "white",
    high = "#B2182B",
    name = "MCSE") +
  scale_color_manual(
    values = c("TRUE" = "white","FALSE" = "black"),guide = "none") +
  labs(
    x = "Correlation (ρ)",
    y = "Method",
    title = "Monte Carlo Standard Errors for Inferential Performance") +
  theme_bw(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "white"),
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1))
```



## RAM

```{r}
data.frame(
  size_MB = round(sapply(
    ls(), function(x) object.size(get(x)) / 1024^2), 1)) %>% 
  arrange(desc(size_MB))
```

