---
title: 'BIOS 6624: Project 2'
author: "Ryan Summers"
date: "2026-03-13"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(magrittr)
library(powertools)
library(dplyr)
library(purrr)
library(tibble)
library(tidyr)
library(InteractionPoweR)
```

## Load Data

```{r}
prelim_df <- read_csv("/Users/ryan_summers/GitHub/BIOS6624-Class/Project2/DataRaw/PrelimData.csv")
```
## Exploratory Data Analysis (EDA)

Calculate correlations between predictors and outcomes
```{r}
cat(
"Pearson correlation between predictor inflammatory marker IL-6 and outcome 
  cognitive measure CVLT is", round(cor(prelim_df$IL_6, prelim_df$CVLT_CNG3),3),
  "in the 95% confidence range of",
  round(cor.test(prelim_df$IL_6,prelim_df$CVLT_CNG3)$conf.int[1],3), "to",
  round(cor.test(prelim_df$IL_6, prelim_df$CVLT_CNG3)$conf.int[2],3),
  "\n\n")

cat(
"Pearson correlation between predictor inflammatory marker IL-6 and outcome
  cortical thickness is", round(cor(prelim_df$IL_6, prelim_df$CORT_CNG3),3)
  ,"in the 95% confidence range of", 
  round(cor.test(prelim_df$IL_6, prelim_df$CORT_CNG3)$conf.int[1],3), "to",
  round(cor.test(prelim_df$IL_6, prelim_df$CORT_CNG3)$conf.int[2],3),
  "\n\n")

cat(
  "Pearson correlation between predictor inflammatory marker MCP-1 and outcome
  cognitive measure CVLT is", round(cor(prelim_df$MCP_1, prelim_df$CVLT_CNG3),3),
  "in the 95% confidence range of", 
  round(cor.test(prelim_df$MCP_1, prelim_df$CVLT_CNG3)$conf.int[1],3), "to",
  round(cor.test(prelim_df$MCP_1, prelim_df$CVLT_CNG3)$conf.int[2],3),
  "\n\n")


cat(
  "Pearson correlation between predictor inflammatory marker MCP-1 and outcome
  cortical thickness is", round(cor(prelim_df$MCP_1, prelim_df$CORT_CNG3),3),
  "in the 95% confidence range of", 
  round(cor.test(prelim_df$MCP_1, prelim_df$CORT_CNG3)$conf.int[1],3), "to",
  round(cor.test(prelim_df$MCP_1, prelim_df$CORT_CNG3)$conf.int[2],3),
  "\n\n")


cat(
  "Pearson correlation between predictor inflammatory marker MCP-1 and predictor
  IL-6 is", round(cor(prelim_df$MCP_1, prelim_df$IL_6),3),
  "in the 95% confidence range of", 
  round(cor.test(prelim_df$MCP_1, prelim_df$IL_6)$conf.int[1],3), "to",
  round(cor.test(prelim_df$MCP_1, prelim_df$IL_6)$conf.int[2],3),
  "\n\n")


cat(
  "Pearson correlation between outcome cognitive and cortical measures is", 
  round(cor(prelim_df$CVLT_CNG3, prelim_df$CORT_CNG3),3),
  "in the 95% confidence range of", 
  round(cor.test(prelim_df$CVLT_CNG3, prelim_df$CORT_CNG3)$conf.int[1],3), "to",
  round(cor.test(prelim_df$CVLT_CNG3, prelim_df$CORT_CNG3)$conf.int[2],3),
  "\n\n")

```

```{r}
cor.test(prelim_df$IL_6, prelim_df$CVLT_CNG3)$conf.int[1]

test <- cor.test(prelim_df$IL_6, prelim_df$CVLT_CNG3)
test$conf.int

```
Sample interpretation:
Preliminary data suggest correlations between inflammatory marker IL-6 and outcome changes from baseline to year-1 in CVLT are in the 95% confidence range of -0.566 to 0.112 with a point estimate of -0.259, which informed the range of correlations evaluated in the power analysis.



```{r}
# convert df to long
prelim_long <- prelim_df %>% 
  pivot_longer(
    cols = c(CVLT_CNG3, CORT_CNG3, IL_6, MCP_1),
    names_to = "variable", values_to = "value")

ggplot(prelim_long, aes(y = value, fill = variable)) +
  geom_boxplot(width = 0.9) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title = "Distributions from Prelim Data",
    x = "",
    y = NULL) +
  theme_bw() +
  theme(legend.position = "none")

```

## Aim 1 Power Analysis

```{r}
# function get corr estimate and 95% CI
cor_ci <- function(data, x, y) {
  ct <- cor.test(data[[x]], data[[y]])
  tibble(
    predictor = x,
    outcome   = y,
    r         = unname(ct$estimate),
    ci_low    = ct$conf.int[1],
    ci_high   = ct$conf.int[2],
    p_value   = ct$p.value,
    n         = sum(complete.cases(data[[x]], data[[y]])))
  }

# correlations for pilot effect-size range
cor_results <- bind_rows(
  cor_ci(prelim_df, "IL_6", "CVLT_CNG3"),
  cor_ci(prelim_df, "IL_6", "CORT_CNG3"),
  cor_ci(prelim_df, "MCP_1", "CVLT_CNG3"),
  cor_ci(prelim_df, "MCP_1", "CORT_CNG3"),
  cor_ci(prelim_df, "IL_6", "MCP_1"),  # predictor-predictor
  cor_ci(prelim_df, "CVLT_CNG3", "CORT_CNG3"))       # outcome-outcome

```


```{r}

alpha_grid <- c(0.025, 0.01, 0.001)
ss_grid    <- c(150, 175, 200, 225, 250, 275)

power_results <- cor_results %>%
  mutate(
    r_low  = pmax(r - 0.3, -0.99),
    r_high = pmin(r + 0.3,  0.99),
    r_grid = map2(r_low, r_high, ~ seq(.x, .y, by = 0.01)),
    grid_df = map(
      r_grid,
      ~ tidyr::crossing(
          r = .x,
          alpha = alpha_grid,
          ss = ss_grid)),
    power_table = map(
      grid_df,
      ~ mutate(
          .x,
          power = pmap_dbl(
            list(r, alpha, ss),
            \(rr, aa, nn)
              powertools::corr.1samp(
                N = nn,
                rho0 = 0,
                rhoA = rr,
                alpha = aa,
                sides = 2
              )))))

power_results_long <- power_results %>%
  dplyr::select(predictor, outcome, power_table) %>%
  tidyr::unnest(power_table)

pred_vars <- c("IL_6", "MCP_1")
out_vars  <- c("CVLT_CNG3", "CORT_CNG3")

plot_dat <- power_results_long %>%
  filter(
    predictor %in% pred_vars,
    outcome %in% out_vars) %>%
  mutate(
    alpha = factor(
      alpha,
      levels = c(0.025, 0.01, 0.001),
      labels = c("0.025", "0.01", "0.001")),
    ss = factor(
      ss,
      levels = c(150, 175, 200, 225, 250, 275),
      labels = c("150", "175", "200", 
                 "225", "250", "275"))) %>%
  arrange(ss, alpha, predictor, outcome, r)

```

```{r}
min_r_80 <- plot_dat %>%
  filter(power >= 0.80) %>%
  group_by(predictor, outcome, alpha, ss) %>%
  slice_max(r, n = 1, with_ties = F) %>%   # closest to 0 since r is negative
  ungroup() %>%
  mutate(
    r_label = sprintf("r = %.2f", r))
```


```{r}
ggplot(min_r_80, aes(x = ss, y = r, color = alpha, group = alpha)) +
  geom_line(linewidth = 0.7, alpha = 0.7) +
  geom_point(size = 0.8, alpha = 0.7) +
  geom_text(aes(label = sprintf("%.2f", r)),
            vjust = -0.6, size = 3, show.legend = FALSE) +
  facet_grid(outcome ~ predictor,
             labeller = labeller(
      predictor = c(IL_6 = "IL-6", MCP_1 = "MCP-1"))) +
  scale_color_manual(
    values = c(
      "0.025" = "steelblue",
      "0.01"  = "#8e63ce",
      "0.001" = "#f46d43")) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  labs(
    x = "sample size",
    y = "Correlation",
    color = expression(alpha),
    title = "Aim 1 minimum detectable correlations for 80% power") +
  theme_bw()+
    theme(
    panel.spacing = unit(1.5, "lines"),
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
min_r_80 <- power_results_long %>%
  filter(predictor %in% pred_vars,
         outcome %in% out_vars,
         power >= 0.80) %>%
  group_by(predictor, outcome, alpha) %>%
  slice_min(order_by = abs(r), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    r_label = paste0("r = ", round(r, 2)))

min_r_80 <- min_r_80 %>%
  mutate(
    alpha = factor(alpha,
                   levels = c(0.025, 0.01, 0.001),
                   labels = c("α = 0.025", 
                              "α = 0.01","α = 0.001")))

```
How large the correlation between an inflammatory marker and outcome decline must be to achieve 80% power under different significance thresholds (α) for sample size 175. Power decreases
toward zero because weaker associations are harder to detect.




```{r, fig.width=10, fig.height=8}
min_r_80_labels <- min_r_80 %>%
  mutate(
    label_x = case_when(
      outcome == "CORT_CNG3" ~ r - 0.075,
      outcome == "CVLT_CNG3" ~ r + 0.08),
    label_y = case_when(
      outcome == "CORT_CNG3" ~ 0.1,
      outcome == "CVLT_CNG3" ~ 0.1))

ggplot(plot_dat, aes(x = r, y = power, color = outcome)) +
  geom_line(linewidth = 0.7, alpha = 0.5) +
  geom_point(size = 0.8, alpha = 0.5) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey40") +
  geom_vline(
    data = min_r_80,
    aes(xintercept = r, color = outcome),
    linetype = "dotted",
    linewidth = 0.8,
    show.legend = F) +
  geom_text(
    data = min_r_80_labels,
    aes(x = label_x, y = label_y, 
        label = r_label, color = outcome),
    angle = 0,
    hjust = 0.5,
    vjust = 1,
    size = 3.5,
    show.legend = F) +
  facet_grid(
    ss ~ predictor,
    labeller = labeller(
      predictor = c(IL_6 = "IL-6", MCP_1 = "MCP-1"))) +
  scale_color_manual(
    values = c("CORT_CNG3" = "steelblue", "CVLT_CNG3" = "firebrick"),
    labels = c("CORT_CNG3" = "CORTΔ", "CVLT_CNG3" = "CVLTΔ")) +
  scale_x_continuous(breaks = seq(-1, 1, by = 0.1)) +
  labs(
    x = "Assumed correlation (pilot estimate ±0.30)",
    y = "Power",
    color = "Outcome",
    title = "Power curves for inflammatory marker–outcome associations",
    subtitle = "Dotted lines show the minimum correlation reaching 80% power") +
  coord_cartesian(ylim = c(0, 1)) +
  theme_bw() +
  theme(
    panel.spacing = unit(1.5, "lines"),
    legend.position = "bottom")
```


```{r}
plot_dat_single_alpha <- plot_dat %>%
  dplyr::filter(alpha == "0.01",
                ss %in% c("150", "175", "200", "275"))

min_r_80_single_alpha <- min_r_80 %>%
  dplyr::filter(alpha == "0.01",
                ss %in% c("150", "175", "200", "275"))

min_r_80_labels_single_alpha <- min_r_80_single_alpha %>%
  mutate(
    label_x = case_when(
      outcome == "CORT_CNG3" ~ r - 0.075,
      outcome == "CVLT_CNG3" ~ r + 0.08
    ),
    label_y = 0.10
  )

ggplot(plot_dat_single_alpha, aes(x = r, y = power, color = outcome)) +
  geom_line(linewidth = 0.7, alpha = 0.7) +
  geom_point(size = 0.8, alpha = 0.7) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey40") +
    geom_vline(
    data = min_r_80_single_alpha,
    aes(xintercept = r, color = outcome),
    linetype = "dotted",
    linewidth = 0.8,
    show.legend = F) +
  geom_text(
    data = min_r_80_labels_single_alpha,
    aes(x = label_x, y = label_y, 
        label = r_label, color = outcome),
    angle = 0,
    hjust = 0.5,
    vjust = 1,
    size = 3.5,
    show.legend = F) +
  facet_grid(ss ~ predictor,
             labeller = labeller(
               predictor = c(IL_6 = "IL-6", MCP_1 = "MCP-1")
             )) +
  scale_color_manual(
    values = c("CORT_CNG3" = "steelblue", "CVLT_CNG3" = "firebrick"),
    labels = c("CORT_CNG3" = "CORTΔ", "CVLT_CNG3" = "CVLTΔ")
  ) +
  scale_x_continuous(breaks = seq(-1, 1, by = 0.1)) +
  labs(
    x = "Assumed correlation (pilot estimate ±0.30)",
    y = "Power",
    color = "Outcome",
    title = "Power curves for inflammatory marker–outcome associations"
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  theme_bw() +
  theme(
    panel.spacing = unit(1.5, "lines"),
    legend.position = "bottom"
  )
```


Example interpretation:
Preliminary data were used to estimate correlations between inflammatory markers and outcome change scores. Because outcomes were defined as baseline minus year-1 values, greater inflammation is expected to be associated with greater decline, corresponding to negative correlations. Power analyses therefore evaluated a range of plausible negative correlations informed by the preliminary estimates.


## Aim 2 Power Analysis

Does AD deposition modify the association between inflammation and 1-year decline?
* Specifically, patients with greater amyloid accumulation and elevated inflammation will show the greatest decline in memory and AD-signature cortical thickness.

For Aim 2, power calculations will evaluate the incremental variance explained by the inflammation × AD pathology interaction in covariate-adjusted linear regression models. Because the primary hypothesis concerns the added contribution of the interaction term beyond the main effects and covariates, power will be assessed using a partial multiple-regression F-test framework.

For Aim 2, amyloid deposition will be operationalized as amyloid positive versus amyloid negative based on established Florbetapir PET cutoffs (global SUVR threshold 1.10). This approach is consistent with the grant’s stated alternative analytic strategy and provides a more feasible and interpretable framework for power analysis than modeling an interaction between two continuous variables. We will test whether the association between inflammation and 1-year change in outcomes differs by amyloid status by fitting change-score regression models including inflammation, amyloid group, and their interaction term.

For Aim 2, power calculations will evaluate the incremental effect of the inflammation × amyloid-status interaction in linear regression models for 1-year change outcomes. Because amyloid will be treated as a dichotomous PET-based variable, this provides a more feasible approximation of the proposed moderation hypothesis than a continuous-by-continuous interaction.

For Aim 2, power will be evaluated for the added inflammation × amyloid-status interaction term using mlrF.partial, with the reduced model including inflammation, amyloid group, age, sex, CDR, and APOE genotype, and the full model additionally including the interaction term.


```{r}
## number of predictors in full model:
## inflammation + factor(amyloid) + interaction + covariates
## covariates: age, sex, CDR, APOE = 4 
## total full-model predictors = 1 + 1 + 1 + 4 = 7
p <- 6   # covariates (4) + inflammation + amyloid
q <- 1   # amyloid*inflammation - testing 1 interaction term


pc_grid    <- seq(0.02, 0.4, by = 0.01)
alpha_grid <- c(0.025, 0.01, 0.001)
ss_grid    <- c(150, 175, 200, 225, 250, 275)

aim2_power <- tidyr::crossing(
  pc = pc_grid,
  alpha = alpha_grid,
  ss = ss_grid) %>%
  mutate(
    power = pmap_dbl(
      list(pc, alpha, ss),
      \(ppc, aa, nn)
        powertools::mlrF.partial(
          N = nn,
          p = p,
          q = q,
          pc = ppc,
          alpha = aa
        ))) %>%
  mutate(
    alpha = factor(
      alpha,
      levels = c(0.025, 0.01, 0.001),
      labels = c("0.025", "0.01", "0.001")),
    ss = factor(
      ss,
      levels = c(150, 175, 200, 225, 250, 275),
      labels = c("N = 150", "N = 175 (expected)", "N = 200", 
                 "N = 225", "N = 250", "N = 275")))
```

```{r}
aim2_min_pc <- aim2_power %>%
  filter(power >= 0.80) %>%
  group_by(alpha, ss) %>%
  slice_min(pc, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    pc_label = sprintf("r = %.2f", pc))
```

```{r}

ggplot(aim2_min_pc, aes(x = ss, y = pc, color = alpha, group = alpha)) +
  geom_line(linewidth = 0.7, alpha = 0.7) +
  geom_point(size =1, alpha=0.7) +
  geom_text(aes(label = sprintf("%.2f", pc)),
            vjust = -0.6, size = 3, show.legend = FALSE) +
  labs(
    x = "Final analytic sample size",
    y = "Minimum detectable partial correlation for 80% power",
    color = expression(alpha),
    title = "Aim 2 minimum detectable interaction effects") +
  theme_bw()+
    theme(
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    panel.spacing = unit(1.5, "lines"),
    legend.position = "bottom") + guides(color = guide_legend(reverse = TRUE))
```

```{r}
aim2_min_pc <- aim2_power %>%
  filter(power >= 0.80) %>%
  group_by(ss, alpha) %>%
  slice_min(pc, n = 1, with_ties = FALSE) %>%
  ungroup()

aim2_min_pc_labels <- aim2_min_pc %>%
  group_by(ss) %>%
  arrange(desc(alpha)) %>%   # controls vertical order
  mutate(
    label_x = max(aim2_power$pc) - 0.005,
    label_y = 0.05 + row_number()*0.11,
    label = sprintf("r = %.2f", pc)) %>%
  ungroup()

ggplot(aim2_power, aes(x = pc, y = power, color = alpha, group = alpha)) +
  geom_line(linewidth = 0.75) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey40") +
  geom_vline(
    data = aim2_min_pc,
    aes(xintercept = pc, color = alpha),
    linetype = "dotted",
    show.legend = FALSE) +
  geom_text(
    data = aim2_min_pc_labels,
    aes(x = label_x, y = label_y, label = label, color = alpha),
    size = 4,
    hjust = 1,
    show.legend = F) +
  scale_color_manual(
  values = c(
    "0.025" = "steelblue",
    "0.01"  = "#8e63ce",
    "0.001" = "#f46d43")) +
  facet_wrap(~ ss, ncol=2) +
  labs(
    x = "Assumed partial correlation for inflammation × amyloid interaction",
    y = "Power",
    color = expression(alpha),
    title = "Aim 2 power curves stratified by final sample size") +
  theme_bw()+
      theme(
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    panel.spacing = unit(1.5, "lines"),
    legend.position = "bottom") 
```


```{r}

pc_vals <- c(0.05, 0.10, 0.15, 0.20, 0.25, 0.30)
ss_grid <- seq(140, 500, by = 5)

aim2_power <- tidyr::crossing(
  pc = pc_vals,
  alpha = alpha_grid,
  ss = ss_grid) %>%
  dplyr::mutate(
    power = purrr::pmap_dbl(
      list(pc, alpha, ss),
      \(ppc, aa, nn)
        powertools::mlrF.partial(
          N = nn,
          p = p,
          q = q,
          pc = ppc,
          alpha = aa)),
    alpha = factor(
      alpha,
      levels = c(0.025, 0.01, 0.001),
      labels = c("0.025", "0.01", "0.001")),
    pc = factor(
      pc,
      levels = pc_vals,
      labels = paste0("r = ", sprintf("%.2f", pc_vals))))
```

```{r}
ggplot(aim2_power, aes(x = ss, y = power, color = alpha, group = alpha)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 175, linetype = "dotted", color = "black") +
  facet_wrap(~ pc, ncol = 3) +
  scale_x_continuous(breaks = seq(150, 500, by = 50)) +
  coord_cartesian(ylim = c(0, 1.02)) +
  labs(
    x = "Final analytic sample size",
    y = "Power",
    color = expression(alpha),
    title = "Aim 2 power by assumed interaction effect",
    subtitle = "Vertical dotted line marks the proposed final sample size (N = 175)"
  ) +
  theme_bw() +
  theme(
    panel.spacing = unit(1.2, "lines"),
    legend.position = "bottom"
  )
```

Power is shown as a function of final analytic sample size for several plausible interaction effect sizes; the vertical line marks the proposed final sample size of 175 participants.

```{r}
# sensitivity analysis using other package
aim2_ipower <- InteractionPoweR::power_interaction(
  n.iter = 1000,
  N = c(155, 175, 195, 215),
  r.x1.y = 0.15, # corr btw inflammation and outcome
  r.x2.y = 0.10, # corr btw amyloid and outcome
  r.x1x2.y = seq(0.05, 0.30, by = 0.01), # corr btw interaction and outcome
  r.x1.x2 = 0.10, # corr btw two predictors
  k.x1 = 0, # number of categories for x1
  k.x2 = 2, # number of categories for x2 (amyloid will be binary)
  k.y = 0, # number of categories for outcome (continuous outcome)
  alpha = c(0.025, 0.01, 0.001),
  detailed_results = FALSE) %>%
  rename(power = pwr)


```


```{r}
aim2_ipower_min <- aim2_ipower %>%
  filter(power >= 0.80) %>%
  group_by(N, alpha) %>%
  slice_min(r.x1x2.y, n = 1, with_ties = FALSE) %>%
  ungroup()

aim2_ipower_min_labels <- aim2_ipower_min %>%
  group_by(N) %>%
  arrange(desc(alpha)) %>%
  mutate(
    label_x = max(aim2_ipower$r.x1x2.y) - 0.005,
    label_y = 0.05 + row_number() * 0.06,
    label = sprintf("α = %s: r = %.2f", alpha, r.x1x2.y)
  ) %>%
  ungroup()

ggplot(aim2_ipower,
       aes(x = r.x1x2.y, y = power, color = factor(alpha), group = alpha)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey40") +
  geom_vline(
    data = aim2_ipower_min,
    aes(xintercept = r.x1x2.y, color = factor(alpha)),
    linetype = "dotted",
    show.legend = FALSE) +
  geom_text(
    data = aim2_ipower_min_labels,
    aes(x = label_x, y = label_y, label = label, color = factor(alpha)),
    hjust = 1,
    size = 3.5,
    show.legend = FALSE) +
  facet_wrap(~ N, ncol = 2) +
  labs(
    x = "Assumed interaction effect correlation",
    y = "Power",
    color = expression(alpha),
    title = "Aim 2 interaction power by sample size") +
  theme_bw()
```




```{r}
aim2_min_pc <- aim2_power %>%
  filter(power >= 0.80) %>%
  group_by(alpha) %>%
  slice_min(pc, n = 1, with_ties = FALSE) %>%
  ungroup()

aim2_min_pc
```



```{r,fig.width=10, fig.height=8}

aim2_min_pc_labels <- aim2_min_pc %>%
  mutate(
    label_x = case_when(
      alpha == 0.05  ~ pc - 0.014,
      alpha == 0.025 ~ pc + 0.014,
      alpha == 0.01  ~ pc + 0.017,
      alpha == 0.001  ~ pc + 0.017,
      TRUE ~ pc),
    label_y = case_when(
      alpha == 0.05  ~ 0.01,
      alpha == 0.025 ~ 0.01,
      alpha == 0.01  ~ 0.01,
      alpha == 0.001  ~ 0.01,
      TRUE ~ 0.05))

ggplot(aim2_power, aes(x = pc, y = power, color = factor(alpha))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  geom_point(size = 0.8, alpha = 0.5) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey40") +
  geom_vline(
    data = aim2_min_pc,
    aes(xintercept = pc, color = factor(alpha)),
    linetype = "dotted",
    show.legend = F) +
  geom_text(
    data = aim2_min_pc_labels,
    aes(x = label_x, y = label_y,
        label = sprintf("r = %.2f", pc),
        color = factor(alpha)),
    angle = 0,
    size = 4,
    vjust = 0,
    show.legend = F) +
scale_color_manual(
  values = c(
    "0.05"  = "#1f78b4",
    "0.025" = "#8e63ce",
    "0.01"  = "#f46d43",
    "0.001" = "#b2182b"),
  labels = c(
    "0.05"  = "0.05",
    "0.025" = "0.025",
    "0.01"  = "0.01",
    "0.001" = "0.001")) +
  labs(
    x = "Assumed partial correlation for inflammation × amyloid interaction",
    y = "Power",
    color = expression(alpha),
    title = "Aim 2 power for the interaction term") +
  theme_bw() +
  theme(
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    panel.spacing = unit(1.5, "lines"),
    legend.position = "bottom") + guides(color = guide_legend(reverse = TRUE))

```

This plot shows how large the inflammation × amyloid interaction must be to detect it with adequate power given sample size (N ≈ 175)
* This is the strength of the inflammation × amyloid interaction after controlling for other predictors.
* As α becomes more strict, the curve shifts to the right, meaning you need a larger interaction effect to achieve the same power.

Where each curve crosses this line tells you:
What interaction effect size is detectable with 80% power.

Power analyses suggest that the proposed sample size is likely adequate for detecting moderate interaction effects in Aim 2, but may be underpowered for small interaction effects. Thus, the current design appears more suitable for detecting relatively strong inflammation × amyloid effects than subtle moderation effects.


A useful way to think about it is to square them:
	*	0.21^2 $\approx$ 0.044
	*	0.31^2 $\approx$ 0.096

So your Aim 2 power plot is saying you’d need the interaction to explain roughly 4% to 10% of residual outcome variance after accounting for the other predictors in the model. For an interaction, that is not small. In many biomedical regression settings, interaction effects are weaker than main effects, often closer to the small range.


### Summary of Analyses

| Aim | Hypothesis | Primary Outcome                                                                | Primary Exposure                                     | Covariates                                              |
| -- | ---------- | ------------------------------------------------------------------------------ | ---------------------------------------------------- | ------------------------------------------------------- |
| 1  | H1a        | 1-yr Δ in CVLT, Benson Figure, MST, and Store Recall (continuous cognitive measures); <br>1-yr Δ in AD-signature cortical thickness (continuous) | Baseline inflammatory markers (continuous cytokines: IL-6, TNF-α, continuous chemokines:MCP-1, Eotaxin-1) | Age, sex, APOE genotype, baseline CDR, baseline outcome |
| 1  | H1b        | Same outcomes                                                                  | Change in inflammatory markers (baseline → year-1)   | Age, sex, APOE genotype, baseline CDR, baseline outcome |
| 2  | H2         | Same outcomes                                                                  | Inflammation × baseline amyloid interaction          | Age, sex, APOE genotype, baseline CDR, baseline outcome |



For Aim 1, pilot data provided preliminary estimates of the association between inflammatory markers and outcomes; therefore, power calculations are summarized as the minimum detectable correlations required to achieve 80% power across candidate sample sizes. In contrast, because pilot data were not available for the inflammation × amyloid interaction in Aim 2, power is displayed across a range of plausible interaction effect sizes.

Those are much smaller than the 0.21–0.31 range your mlrF.partial results suggested is needed for 80% power at N = 175, which would support the conclusion that Aim 2 is likely underpowered for realistic small interaction effects. That is also broadly consistent with the literature the tutorial cites, which emphasizes that interaction effects are often substantially smaller than main effects.

Because pilot data supported moderate main effects of inflammatory markers but did not permit direct estimation of the inflammation × amyloid interaction, Aim 2 power was evaluated across an attenuated range of plausible partial correlations for the interaction term and across several feasible final sample sizes centered on the proposed N = 175.



