---
title: 'BIOS 6624: Project 2'
author: "Ryan Summers"
date: "2026-03-13"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(magrittr)
library(powertools)
library(dplyr)
library(purrr)
library(tibble)
library(tidyr)
library(InteractionPoweR)
library(broom)
library(gt)
```

## Load Data

```{r}
prelim_df <- read_csv("/Users/ryan_summers/GitHub/BIOS6624-Class/Project2/DataRaw/PrelimData.csv")
```
## Exploratory Data Analysis (EDA)

Calculate correlations between predictors and outcomes
```{r}
cat(
"Pearson correlation between predictor inflammatory marker IL-6 and outcome 
  cognitive measure CVLT is", round(cor(prelim_df$IL_6, prelim_df$CVLT_CNG3),3),
  "in the 95% confidence range of",
  round(cor.test(prelim_df$IL_6,prelim_df$CVLT_CNG3)$conf.int[1],3), "to",
  round(cor.test(prelim_df$IL_6, prelim_df$CVLT_CNG3)$conf.int[2],3),
  "\n\n")

cat(
"Pearson correlation between predictor inflammatory marker IL-6 and outcome
  cortical thickness is", round(cor(prelim_df$IL_6, prelim_df$CORT_CNG3),3)
  ,"in the 95% confidence range of", 
  round(cor.test(prelim_df$IL_6, prelim_df$CORT_CNG3)$conf.int[1],3), "to",
  round(cor.test(prelim_df$IL_6, prelim_df$CORT_CNG3)$conf.int[2],3),
  "\n\n")

cat(
  "Pearson correlation between predictor inflammatory marker MCP-1 and outcome
  cognitive measure CVLT is", round(cor(prelim_df$MCP_1, prelim_df$CVLT_CNG3),3),
  "in the 95% confidence range of", 
  round(cor.test(prelim_df$MCP_1, prelim_df$CVLT_CNG3)$conf.int[1],3), "to",
  round(cor.test(prelim_df$MCP_1, prelim_df$CVLT_CNG3)$conf.int[2],3),
  "\n\n")


cat(
  "Pearson correlation between predictor inflammatory marker MCP-1 and outcome
  cortical thickness is", round(cor(prelim_df$MCP_1, prelim_df$CORT_CNG3),3),
  "in the 95% confidence range of", 
  round(cor.test(prelim_df$MCP_1, prelim_df$CORT_CNG3)$conf.int[1],3), "to",
  round(cor.test(prelim_df$MCP_1, prelim_df$CORT_CNG3)$conf.int[2],3),
  "\n\n")


cat(
  "Pearson correlation between predictor inflammatory marker MCP-1 and predictor
  IL-6 is", round(cor(prelim_df$MCP_1, prelim_df$IL_6),3),
  "in the 95% confidence range of", 
  round(cor.test(prelim_df$MCP_1, prelim_df$IL_6)$conf.int[1],3), "to",
  round(cor.test(prelim_df$MCP_1, prelim_df$IL_6)$conf.int[2],3),
  "\n\n")


cat(
  "Pearson correlation between outcome cognitive and cortical measures is", 
  round(cor(prelim_df$CVLT_CNG3, prelim_df$CORT_CNG3),3),
  "in the 95% confidence range of", 
  round(cor.test(prelim_df$CVLT_CNG3, prelim_df$CORT_CNG3)$conf.int[1],3), "to",
  round(cor.test(prelim_df$CVLT_CNG3, prelim_df$CORT_CNG3)$conf.int[2],3),
  "\n\n")

```

```{r}
cor.test(prelim_df$IL_6, prelim_df$CVLT_CNG3)$conf.int[1]

test <- cor.test(prelim_df$IL_6, prelim_df$CVLT_CNG3)
test$conf.int

```
Sample interpretation:
Preliminary data suggest correlations between inflammatory marker IL-6 and outcome changes from baseline to year-1 in CVLT are in the 95% confidence range of -0.566 to 0.112 with a point estimate of -0.259, which informed the range of correlations evaluated in the power analysis.

```{r}
# sample models
il6.cvlt.model <- lm(CVLT_CNG3~IL_6, data=prelim_df)
tidy(il6.cvlt.model)

il6.cort.model <- lm(CORT_CNG3~IL_6, data=prelim_df)
tidy(il6.cort.model)

mcp.cvlt.model <- lm(CVLT_CNG3~MCP_1, data=prelim_df)
tidy(mcp.cvlt.model)

mcp.cort.model <- lm(CORT_CNG3~MCP_1, data=prelim_df)
tidy(mcp.cort.model)

# both predictors in model
cvlt.model <- lm(CVLT_CNG3~MCP_1+IL_6, data=prelim_df)
tidy(cvlt.model)

cort..model <- lm(CORT_CNG3~MCP_1+IL_6, data=prelim_df)
tidy(cort..model)
```


```{r}
# convert df to long
prelim_long <- prelim_df %>% 
  pivot_longer(
    cols = c(CVLT_CNG3, CORT_CNG3, IL_6, MCP_1),
    names_to = "variable", values_to = "value")

ggplot(prelim_long, aes(y = value, fill = variable)) +
  geom_boxplot(width = 0.9) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title = "Distributions from Prelim Data",
    x = "",
    y = NULL) +
  theme_bw() +
  theme(legend.position = "none")

```

## Aim 1 Power Analysis

```{r}
# function get corr estimate and 95% CI
cor_ci <- function(data, x, y) {
  ct <- cor.test(data[[x]], data[[y]])
  tibble(
    predictor = x,
    outcome   = y,
    r         = unname(ct$estimate),
    ci_low    = ct$conf.int[1],
    ci_high   = ct$conf.int[2],
    p_value   = ct$p.value,
    n         = sum(complete.cases(data[[x]], data[[y]])))
  }

# correlations for pilot effect-size range
cor_results <- bind_rows(
  cor_ci(prelim_df, "IL_6", "CVLT_CNG3"),
  cor_ci(prelim_df, "IL_6", "CORT_CNG3"),
  cor_ci(prelim_df, "MCP_1", "CVLT_CNG3"),
  cor_ci(prelim_df, "MCP_1", "CORT_CNG3"),
  cor_ci(prelim_df, "IL_6", "MCP_1"),  # predictor-predictor
  cor_ci(prelim_df, "CVLT_CNG3", "CORT_CNG3"))       # outcome-outcome

```


How large the correlation between an inflammatory marker and outcome decline must be to achieve 80% power under different significance thresholds (α) for sample size 175. Power decreases
toward zero because weaker associations are harder to detect.

```{r}

pred_vars <- c("IL_6", "MCP_1")
out_vars  <- c("CVLT_CNG3", "CORT_CNG3")

# filter to only predictor and outcome correlation calculations
aim1_pairs <- cor_results %>%
  dplyr::filter(
    predictor %in% pred_vars,
    outcome %in% out_vars)

# primary and conservative alpha
alpha_grid <- c(0.01, 0.005, 0.0025)

# initialize results object
aim1_pwr_results <- data.frame()

# loop over each pilot predictor-outcome pair
for (i in seq_len(nrow(aim1_pairs))) {
  
  pred_i <- aim1_pairs$predictor[i]
  out_i  <- aim1_pairs$outcome[i]
  r_hat  <- aim1_pairs$r[i]
  
  # define correlation range based on pilot data
  r_low  <- max(r_hat - 0.30, -0.99)
  r_high <- min(r_hat + 0.30,  0.99)
  r_grid <- seq(r_low, r_high, by = 0.01)
  
  for (a in alpha_grid) {
    for (r in r_grid) {
      # power function
      pwr <- powertools::corr.1samp(
        N = 175, # target sample size
        rho0 = 0,
        rhoA = r,
        alpha = a,
        sides = 2)
      # converts to long format
      aim1_pwr_results <- rbind(
        aim1_pwr_results,
        data.frame(
          predictor = pred_i,
          outcome   = out_i,
          pilot_r   = r_hat,
          r         = r,
          alpha     = a,
          power     = pwr))
    }
  }
  }

```

```{r}
# plotting dataset
aim1_plot_dat <- aim1_pwr_results %>%
  mutate(
    alpha = factor(
      alpha,
      levels = c(0.01, 0.005, 0.0025),
      labels = c("α = 0.01",
        "α = 0.005", "α = 0.0025"))) %>%
  arrange(alpha, predictor, outcome, r)

# min detectable correlation for 80% power
aim1_min_r_80 <- aim1_plot_dat %>%
  dplyr::filter(power >= 0.80) %>%
  group_by(predictor, outcome, alpha) %>%
  slice_min(order_by = abs(r), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(r_label = paste0("r = ", round(r, 2)))

# label positions
aim1_min_r_80_labels <- aim1_min_r_80 %>%
  dplyr::mutate(
    label_x = dplyr::case_when(
      outcome == "CORT_CNG3" ~ r - 0.075,
      outcome == "CVLT_CNG3" ~ r + 0.065),
    label_y = 0.10)

# plot
ggplot(aim1_plot_dat, aes(x = r, y = power, color = outcome)) +
  geom_line(linewidth = 0.7, alpha = 0.5) +
  geom_point(size = 0.8, alpha = 0.5) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey40") +
  geom_vline(
    data = aim1_min_r_80,
    aes(xintercept = r, color = outcome),
    linetype = "dotted",
    linewidth = 0.8,
    show.legend = F) +
  geom_text(
    data = aim1_min_r_80_labels,
    aes(x = label_x, y = label_y, label = r_label, color = outcome),
    size = 3.5,
    show.legend = F) +
  facet_grid(
    alpha ~ predictor,
    labeller = labeller(
      predictor = c(IL_6 = "IL-6 (CORTΔ r=-0.6; CVLTΔ r=-0.26)", 
                    MCP_1 = "MCP-1 (CORTΔ r=-0.69; CVLTΔ r=-0.32)"))) +
  scale_color_manual(
    values = c("CORT_CNG3" = "steelblue", "CVLT_CNG3" = "firebrick"),
    labels = c("CORT_CNG3" = "CORTΔ", "CVLT_CNG3" = "CVLTΔ")) +
  scale_x_continuous(breaks = seq(-1, 1, by = 0.1)) +
  labs(
    x = "Assumed correlation (pilot estimate ±0.30)",
    y = "Power",
    color = "Outcome",
    title = "Power curves for inflammatory marker–outcome associations at N = 175",
    subtitle = "Dotted lines show the minimum correlation reaching 80% power") +
  coord_cartesian(ylim = c(0, 1)) +
  theme_bw(base_size = 12) +
  theme(
    panel.spacing = unit(1.5, "lines"),
    legend.position = "bottom")
```



```{r}
r_grid <- seq(-0.5, -0.2, by = 0.01)
alpha_grid <- c(0.05/4, 0.05/10, 0.05/20)

aim1_power_full <- data.frame()

for (a in alpha_grid) {
  for (r in r_grid) {
    
    pwr <- corr.1samp(
      N = 175,
      rho0 = 0,
      rhoA = r,
      alpha = a,
      sides = 2)
    
    aim1_power_full <- rbind(
      aim1_power_full,
      data.frame(r = r, alpha = a, power = pwr))
  }
  }
```

```{r}
min_r_80 <- aim1_power_full %>%
  dplyr::filter(power >= 0.80) %>%
  group_by(alpha) %>%
  slice_min(order_by = abs(r), n = 1, with_ties = FALSE) %>%
  ungroup()
```

```{r}
r_vals <- c(-0.20, -0.25, -0.30, -0.35, -0.40, -0.45, -0.50)

aim1_power_table <- aim1_power_full %>%
  dplyr::filter(round(r, 2) %in% r_vals)

min_r_80_row <- min_r_80 %>%
  dplyr::mutate(
    alpha = dplyr::case_when(
      alpha == 0.0125 ~ "α = 0.05/4",
      alpha == 0.005 ~ "α = 0.05/10",
      alpha == 0.0025 ~ "α = 0.05/20"),
    value = round(abs(r)*100, 1)) %>%
  dplyr::select(alpha, value) %>%
  pivot_wider(
    names_from = alpha,
    values_from = value) %>%
  mutate(correlation = "Min |r| for 80% power") %>%
  dplyr::select(correlation, `α = 0.05/4`, `α = 0.05/10`, `α = 0.05/20`)

aim1_power_table_wide <- aim1_power_table %>%
  mutate(
    correlation = sprintf("%.2f", abs(r)),
    alpha = case_when(
      alpha == 0.0125 ~ "α = 0.05/4",
      alpha == 0.005  ~ "α = 0.05/10",
      alpha == 0.0025 ~ "α = 0.05/20"),
    power = round(power * 100, 1)) %>%
  dplyr::select(correlation, alpha, power) %>%
  pivot_wider(
    names_from = alpha,
    values_from = power)

final_table <- dplyr::bind_rows(
  aim1_power_table_wide,
  min_r_80_row)

final_table
```

```{r label="Table used in final report (Table 1)"}

gt_power_table <- final_table %>%
  gt(rowname_col = "correlation") %>%
  tab_header(
    title = md("**Aim 1 Power at N = 175**"),
    subtitle = md("Power (%) across assumed correlations and 
                  significance thresholds")) %>%
  tab_stubhead(label = md("**Assumed |r|**")) %>%  # Add this line
  cols_label(
    correlation = "Assumed |r|",
    `α = 0.05/4`   = md("**α = 0.05/4**"),
    `α = 0.05/10`  = md("**α = 0.05/10**"),
    `α = 0.05/20` = md("**α = 0.05/20**")) %>%
  
   cols_width(
    stub() ~ px(150),  # This controls the stub column width
    everything() ~ px(87)  # Optional: control other column widths
  ) %>%
  
  fmt_missing(
    columns = everything(),
    missing_text = "") %>%
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_stub(rows = correlation == "Min |r| for 80% power")) %>%
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_body(
      rows = correlation == "Min |r| for 80% power",
      columns = c(`α = 0.05/4`, `α = 0.05/10`, `α = 0.05/20`))) %>%
  
  # Add these tab_style calls for background highlighting
  tab_style(
    style = list(
      cell_borders(
        sides = c("top", "bottom"),
        color = "gray60",
        weight = px(1),
        style = "solid"),
      cell_fill(color = "lightgreen")),
    locations = cells_body(
      rows = correlation == "Min |r| for 80% power",
      columns = everything())) %>%
  tab_style(
    style = list(
      cell_borders(
        sides = c("top", "bottom"),
        color = "gray60",
        weight = px(1),
        style = "solid"),
      cell_fill(color = "lightgreen")),
    locations = cells_stub(
      rows = correlation == "Min |r| for 80% power")) %>%
  
  tab_spanner(label = md("**Power (%)**"),
              columns = c(`α = 0.05/4`, `α = 0.05/10`, `α = 0.05/20`)) %>% 
  tab_source_note(
    source_note = md(
      "Minimum detectable correlation corresponds to the smallest absolute 
      correlation achieving at least 80% power.")) %>%
  tab_options(
    table.width = pct(65),
    table.border.left.style = "none",
    table.border.right.style = "none",
    stub.border.style = "none",
    table.font.size = px(12),
    heading.title.font.size = px(14),
    heading.subtitle.font.size = px(11),
    data_row.padding = px(4),
    source_notes.font.size = px(10))

gt_power_table
```

## Aim 2 Power Analysis

```{r}
## number of predictors in full model:
## inflammation + factor(amyloid) + interaction + covariates
## covariates: age, sex, CDR, APOE = 4 
## total full-model predictors = 1 + 1 + 1 + 4 = 7
p <- 6   # covariates (4) + inflammation(1) + amyloid (1) - per model
q <- 1   # amyloid*inflammation - one interaction term - per model


## plausible interaction partial correlations
## literature states that interactions are typically much smaller than main effects
  ## 1/3 of main effects in some studies
pc_grid <- seq(0.03, 0.4, by = 0.01)

# initialize aim 2 results object
aim2_pwr_results <- data.frame()

for (a in alpha_grid) {
  for (pc in pc_grid) {
    
    pwr <- powertools::mlrF.partial(
      N = 175,
      p = p,
      q = q,
      pc = pc, # pc only accepts positive correlations
      alpha = a)
    
    aim2_pwr_results <- rbind(
      aim2_pwr_results,
      data.frame(
        pc = pc,
        alpha = a,
        power = pwr))
  }
  }

```


```{r}
# plotting data
aim2_plot_dat <- aim2_pwr_results %>%
  mutate(
    alpha = factor(
      alpha,
      levels = c(0.05, 0.01, 0.005, 0.0025),
      labels = c("α = 0.05", "α = 0.01",
        "α = 0.005", "α = 0.0025"))) %>%
  arrange(alpha, pc)

# minimum detectable interaction for 80% power
aim2_min_pc_80 <- aim2_plot_dat %>%
  filter(power >= 0.80) %>%
  group_by(alpha) %>%
  slice_min(order_by = pc, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    pc_label = paste0("r = ", round(pc, 2)),
    label_x = pc + 0.01,
    label_y = 0.08)

# label positions
aim2_min_pc_80_labels <- aim2_min_pc_80 %>%
  mutate(
    pc_label = paste0("r = ", round(pc, 2)),
    label_x = case_when(
      alpha == "α = 0.05"  ~ pc - 0.015,
      alpha == "α = 0.01"  ~ pc - 0.015,
      alpha == "α = 0.005"  ~ pc + 0.005,
      alpha == "α = 0.0025" ~ pc + 0.015),
    label_y = case_when(
      alpha == "α = 0.05"  ~ 0.08,
      alpha == "α = 0.01"  ~ 0.08,
      alpha == "α = 0.005"  ~ 0.12,
      alpha == "α = 0.0025" ~ 0.08))
```


Change correlation direction to match Aim1 to be consistent
```{r}
aim2_plot_dat2 <- aim2_pwr_results %>%
  mutate(
    pc_plot = -pc,
    alpha = factor(
      alpha,
      levels = c(0.05, 0.01, 0.005, 0.0025),
      labels = c("0.05", "0.01",
        "0.005", "0.0025"))) 

aim2_min_pc_802 <- aim2_plot_dat2 %>%
  filter(power >= 0.80) %>%
  group_by(alpha) %>%
  slice_min(order_by = pc, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    pc_label = paste0("r = ", round(pc_plot, 2)),
    label_x = pc_plot - 0.015,
    label_y = 0.08)


aim2_min_pc_80_labels2 <- aim2_min_pc_802 %>%
  mutate(
    pc_label = paste0("r = ", round(-pc, 2)),
    label_x = case_when(
      alpha == "0.05"  ~ pc_plot + 0.017,
      alpha == "0.01"  ~ pc_plot + 0.017,
      alpha == "0.005"  ~ pc_plot - 0.015,
      alpha == "0.0025" ~ pc_plot - 0.017),
    label_y = case_when(
      alpha == "0.05"  ~ 0.08,
      alpha == "0.01"  ~ 0.08,
      alpha == "0.005"  ~ 0.12,
      alpha == "0.0025" ~ 0.08))

ggplot(aim2_plot_dat2, aes(x = pc_plot, y = power, color = alpha)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.2) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey40") +
  geom_vline(
    data = aim2_min_pc_802,
    aes(xintercept = pc_plot, color = alpha),
    linetype = "dotted",
    linewidth = 0.8,
    show.legend = F) +
  geom_text(
    data = aim2_min_pc_80_labels2,
    aes(x = label_x, y = label_y, label = pc_label, color = alpha),
    size = 3.5,
    show.legend = F) +
  scale_color_manual(
    values = c("0.01" = "steelblue", "0.005" = "#d2980d",
       "0.0025" = "#de425b")) +
  labs(
    x = "Assumed partial correlation for inflammation × amyloid interaction",
    y = "Power",
    color = expression(alpha),
    title = "Aim 2 power for the interaction effect at N = 175",
    subtitle = "Negative values shown to reflect the hypothesized direction") +
  coord_cartesian(ylim = c(0, 1)) +
  theme_bw() +
  theme(legend.position = "bottom")
  

```

```{r}
sidak <- function(alpha, m) {
  1- (1-alpha)^(1/m)}
```

```{r}
sidak(alpha = 0.05, m = 20)
0.05/20
```


```{r}
N_total <- 175
alpha <- 0.05/4  # assuming 4 independent tests

## low amyloid group: small plausible correlations
#rho_low_grid <- c(-0.10, -0.20, -0.30)
rho_low_grid <- c(0.10, 0.20, 0.30)

## plausible high-amyloid correlations
#rho_high_grid <- seq(-0.70, -0.10, by = 0.01)
rho_high_grid <- seq(0.20, 0.8, by = 0.01)

## ratio = n_high / n_low
nratio_grid <- c(1, 0.75, 0.5, 0.25)

aim2_corr2samp_results <- data.frame()

for (rho_low in rho_low_grid) {
  for (rho_high in rho_high_grid) {
    for (nr in nratio_grid) {

      ## ensure analysis only considers cases where
      ## |rho_high| > |rho_low|
      if (abs(rho_high) <= abs(rho_low)) next

      ## group sizes
      n_low  <- floor(N_total / (1 + nr))
      n_high <- N_total - n_low

      pwr_obj <- corr.2samp(
        n1      = n_low,
        n.ratio = n_high / n_low,
        rho1    = rho_low,
        rho2    = rho_high,
        alpha   = alpha,
        sides   = 2)

      power_val <- if (is.list(pwr_obj) && "power" %in% names(pwr_obj)) {
        pwr_obj$power
        } else {
          as.numeric(pwr_obj)
          }

      aim2_corr2samp_results <- rbind(
        aim2_corr2samp_results,
        data.frame(
          alpha    = alpha,
          rho_low  = rho_low,
          rho_high = rho_high,
          n_low    = n_low,
          n_high   = n_high,
          n_ratio  = n_high / n_low,
          power    = power_val))
    }
  }
  }
```


```{r label="Figure used in final report (Figure 1)"}
# better labels for facet grid
# aim2_corr2samp_results$pos_label <- paste0(
# "Amyloid+ : ", aim2_corr2samp_results$n_high)

aim2_corr2samp_results$pos_label <- paste0(aim2_corr2samp_results$n_high)

aim2_corr2samp_results$neg_label <- paste0(
"Amyloid Neg- : ", aim2_corr2samp_results$n_low)

# create labels for rho for each facet grid
annot_df <- aim2_corr2samp_results %>%
  group_by(rho_low, pos_label) %>%
  filter(power >= 0.80) %>%
  slice_min(order_by = rho_high, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(label = paste0("r = ", sprintf("%.2f", rho_high)))

annot_df <- aim2_corr2samp_results %>%
  group_by(rho_low, pos_label, n_high) %>%
  summarise(
    rho_star = if (any(power >= 0.80)) 
      min(rho_high[power >= 0.80]) else NA_real_, # change to max if using neg
    .groups = "drop") %>%
  mutate(
    label = ifelse(
      is.na(rho_star),
      "min not reached",
      paste0("r = ", sprintf("%.2f", rho_star))),
    rho_low = factor(rho_low,levels = c(0.1, 0.2, 0.3))) %>%
  group_by(rho_low) %>%
  arrange(desc(n_high)) %>%
  mutate(
    x = 0.2,
    y = seq(0.30, 0.50, length.out = n())) %>%
  ungroup()

ggplot(aim2_corr2samp_results,
       aes(x = rho_high, y = power,
           color = pos_label,
           group = pos_label)) +
  geom_line(linewidth = 0.8, alpha = 0.75) +
  geom_hline(yintercept = 0.80, linetype = "dashed") +
  geom_text(data = annot_df,
    aes(x = x, y = y, label = label, color = pos_label),
    inherit.aes = FALSE, hjust = 0, size = 3.5, show.legend = FALSE) +
  facet_wrap(~ rho_low, nrow = 1,
    labeller = labeller(
      rho_low = function(x) paste0("Amyloid Neg- (r = ", x, ")"))) +
  scale_x_continuous("Correlation in Amyloid + group",
                     breaks = seq(0.2, 1, by = 0.1))  +
  scale_y_continuous("Power", limits = c(0, 1)) +
  scale_color_discrete(name = "Amyloid Pos+ (N)") +
    labs(
    title = "Aim 2 Power for the Interaction Effect at N = 175",
    subtitle = expression(paste("Text labels show min correlation required to achieve 80% power at ", alpha, "=0.05/4 in Amyloid +"))) +
  theme_bw() +
  theme(legend.position = "bottom")
```



```{r}
threshold_table <- aim2_corr2samp_results %>%
  arrange(alpha, n_ratio, abs(rho_high)) %>%
  group_by(alpha, n_ratio) %>%
  filter(power >= 0.80) %>%
  slice(1) %>%
  ungroup()

threshold_table
```

```{r}
# play around with different scenarios 
corr.2samp(
  n1= 87, 
  n.ratio = 88 / 87, 
  rho1 = 0.075, # when neg amyloid group neg. corr incr -> pwr incr and vice-versa
  rho2 = 0.5, # weaker high amyloid corr lowers pwr
  alpha = 0.025,
  sides = 2)

corr.2samp(
  n1= 87, 
  n.ratio = 88 / 87, 
  rho1 = 0.2, 
  rho2 = -0.4, 
  alpha = 0.01,
  sides = 2)

corr.2samp(
  n1= 87, 
  n.ratio = 88 / 87, 
  rho1 = 0.3, 
  rho2 = -0.4,
  alpha = 0.01, 
  sides = 2)
```

If the results aren't as correlated as we expected, then we have to treat each test
as independent

### Sesnitivity Analysis

Using a different package
```{r, eval=FALSE}
aim2_ipower <- InteractionPoweR::power_interaction(
  n.iter = 1000,
  N = 175, # target
  r.x1.y = 0.10, # r btw inflammation and outcome
  r.x2.y = 0.00, # r btw amyloid and outcome - **purely speculative**
  r.x1x2.y = seq(0.05, 0.30, by = 0.01), # r btw interaction and outcome
  r.x1.x2 = 0.05, # r btw amyloid and inflammation
  k.x1 = 0, # number of categories for x1
  k.x2 = 2, # number of categories for x2 (amyloid will be binary)
  k.y = 0, # number of categories for outcome (continuous outcome)
  alpha = c(0.01, 0.005, 0.0025),
  cl = parallel::detectCores() - 5,
  detailed_results = T) %>%
  rename(power = pwr)

```


```{r, eval=FALSE}
aim2_ipower_min <- aim2_ipower %>%
  filter(power >= 0.80) %>%
  group_by(alpha) %>%
  slice_min(r.x1x2.y, n = 1, with_ties = FALSE) %>%
  ungroup()

aim2_ipower_min_labels <- aim2_ipower_min %>%
  arrange(desc(alpha)) %>%
  mutate(
    label_x = max(aim2_ipower$r.x1x2.y) - 0.005,
    label_y = 0.05 + row_number() * 0.06,
    label = sprintf("r = %.2f", r.x1x2.y)) %>%
  ungroup()

ggplot(aim2_ipower,
       aes(x = r.x1x2.y, y = power, color = factor(alpha), group = alpha)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey40") +
  geom_vline(
    data = aim2_ipower_min,
    aes(xintercept = r.x1x2.y, color = factor(alpha)),
    linetype = "dotted",
    show.legend = FALSE) +
  geom_text(
    data = aim2_ipower_min_labels,
    aes(x = label_x, y = label_y, label = label, color = factor(alpha)),
    hjust = 1,
    size = 3.5,
    show.legend = FALSE) +
  facet_wrap(~ alpha, ncol = 2) +
  labs(
    x = "Assumed interaction effect correlation",
    y = "Power",
    color = expression(alpha),
    title = "Aim 2 interaction power by sample size") +
  theme_bw()+
  theme(legend.position = "bottom")
```


### Summary of Analyses (Used in final report)

| Aim | Hypothesis | Primary Outcome                                                                | Primary Exposure                                     | Covariates                                              |
| -- | ---------- | ------------------------------------------------------------------------------ | ---------------------------------------------------- | ------------------------------------------------------- |
| 1  | H1a        | 1-yr Δ in CVLT, Benson Figure, MST, and Store Recall (continuous cognitive measures); <br>1-yr Δ in AD-signature cortical thickness (continuous) | Baseline inflammatory markers (continuous cytokines: IL-6, TNF-α, continuous chemokines:MCP-1, Eotaxin-1) | Baseline age, sex, & baseline outcome |
| 1  | H1b        | Same outcomes                                                                  | Change in inflammatory markers (baseline → year-1)   | Baseline age, sex, baseline outcome, and baseline inflammatory marker level |
| 2  | H2         | Same outcomes                                                                  | baseline Inflammation × baseline amyloid interaction          | Same as H1b |

licity burden is likely between the extreme cases of complete independence and complete redundancy.

