---
title: 'BIOS 6624: Project 3'
author: "Ryan Summers"
date: "2026-04-01"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(magrittr)
library(dplyr)
library(purrr)
library(tibble)
library(tidyr)
library(broom)
library(gt)
library(gtsummary)
library(survival)
library(survminer)

```

## Load Data

```{r}
fmhst_raw <- read_csv("/Users/ryan_summers/GitHub/BIOS6624-Class/Project3/DataRaw/frmgham2.csv")
```

## Data Management & Cleanup

Remove subjects who had event (stroke) at baseline and Filter to 10yrs cutoff.
Calculate censored indicator if event doesn't happen within 10yrs and calculate
time to censoring variable.

```{r}
cutoff_days <- 10*365.25

fmhst_filt <- fmhst_raw %>% 
  # remove subjects who had event prior to baseline assessment
  filter(PERIOD == 1, PREVSTRK == 0) %>%   # baseline cohort
  mutate(
    # time to event or being censored truncated at 10 years
    TIMESTRKorCENS_10y = pmin(TIMESTRK, cutoff_days),

    # event indicator within 10 years
    STROKE_10y = if_else(STROKE == 1 & TIMESTRK <= cutoff_days, 1, 0),
    
    ## all censored subjects
    ## **TIMESTRK includes censored participants if STROKE==0**
    CENSORED_10y = if_else(STROKE_10y == 0, 1, 0),
    
    # convert variables to factors
    SEX = factor(SEX, levels = c(1,2), labels = c("Men", "Women")),
    DIABETES = factor(DIABETES, levels = c(0,1), labels = c("No", "Yes")),
    PREVCHD = factor(PREVCHD, levels = c(0,1), labels = c("No", "Yes")),
    BPMEDS = factor(BPMEDS, levels = c(0,1), labels = c("No", "Yes")),
    CURSMOKE = factor(CURSMOKE, levels = c(0,1), labels = c("No", "Yes")),
    STROKE_10y = factor(STROKE_10y, levels = c(0,1), labels = c("No-Censored", "Yes")),
    STROKE_bin = ifelse(STROKE_10y == "Yes", 1, 0),
    CENSORED_10y = factor(CENSORED_10y, levels = c(0, 1), labels = c("No", "Yes")),
    TIMESTRKorCENS_10y_yrs = TIMESTRKorCENS_10y/365.25,
    )
```

We restricted the primary analysis to baseline risk factors measured at period 1 and 
modeled incident stroke over the first 10 years of follow-up. This approach aligns 
predictor measurement more closely with the risk window of interest. We additionally 
examined how major risk factors changed across the three examination periods to assess whether a time-varying covariate analysis might be informative. If risk factors change materially over time, 
a baseline-only model may understate or misclassify the relationship between current 
risk-factor status and stroke hazard. However, time-varying covariate analyses in 
observational data may also be harder to interpret because updated covariates can be
influenced by treatment changes and evolving disease status.

## Exploratory Data Analysis (EDA)

### Table1

```{r label = "Used in final report as Table 1"}
tbl1 <- fmhst_filt %>%
  tbl_summary(
    by = SEX,
    include = c(
      AGE, SYSBP, DIABETES, TOTCHOL, PREVCHD, BPMEDS, CURSMOKE,
      BMI, STROKE_10y),
    type = list(
      all_continuous() ~ "continuous2",
      all_categorical() ~ "categorical"),
    statistic = list(
      all_continuous() ~ c(
        "{mean} ({sd})",
        "{median} ({p25}, {p75})",
        "{min} - {max}"),
      all_categorical() ~ "{n} ({p}%)"),
    digits = list(AGE ~ 0, SYSBP ~ 0),
    missing = "ifany",
    missing_text = "NA",
    missing_stat = "{N_miss} ({p_miss}%)",
    label = list(
      AGE ~ "Age",
      SYSBP ~ "Systolic BP",
      DIABETES ~ "Diabetes",
      TOTCHOL ~ "Total Cholesterol (mg/dL)",
      PREVCHD ~ "Prevalent Coronary Heart Disease",
      BPMEDS ~ "BP Meds",
      CURSMOKE ~ "Smoker",
      STROKE_10y ~ "Event (Stroke)")) %>%
  add_overall() %>% bold_labels() %>% italicize_levels() %>% add_n() %>%
  as_gt() %>%
  gt::fmt_markdown(columns = everything()) %>%
  tab_header(
    title = "Characteristics of the Framingham Heart Study Cohort") %>% 
  tab_options(
    table.width = pct(70),
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

tbl1
```

```{r}
# number of participants
cat("Number of participants in entire cohort", 
   length(unique(fmhst_raw$RANDID)), "\n\n")

cat("Number of participants in 10yr subset cohort", 
   length(unique(fmhst_filt$RANDID)), "\n\n")
```


```{r}
hist(fmhst_filt$TOTCHOL)
```


STROKE indicator is 0 but there's a time to stroke value????
```{r}
fmhst_filt %>% 
  filter(STROKE==0 & TIMESTRK <= 10*365.25) %>% 
  select(c(1,24,29,), 36:43)
```

```{r}
# crude incidence
fmhst_filt %>%
  group_by(SEX) %>%
  summarise(
    events = sum(STROKE_10y == "Yes"),
    total = n(),
    risk = round( (events / total)*100,1 ))
```

### Kaplan-Meier Curve
```{r}
km.fit <- survfit(Surv(TIMESTRKorCENS_10y, STROKE_10y == "Yes") ~ BPMEDS, 
                  data = fmhst_filt)

ggsurvplot(km.fit,
           conf.int = F,
           risk.table = T,
           ylim = c(0.97,1),
           xlab = "Time (Days)",
           ylab = "Stroke-free probability",
           tables.height = 0.2,
           tables.theme = theme_cleantable(),
            # Color palettes. Use custom color: c("#E7B800", "#2E9FDF"),
           # or brewer color (e.g.: "Dark2"), or ggsci color (e.g.: "jco")
           palette = c("#E7B800", "#2E9FDF"),
           ggtheme = theme_bw() )
           
```

```{r}

# create grouped versions of variables
fmhst_plot <- fmhst_filt %>%
  mutate(
    SBP_cat = if_else(SYSBP >= 140, "BP≥140", "BP<140"),
    AGE_cat = if_else(AGE >= 60, "Age≥60", "Age<60"),
    TOTCHOL_cat = if_else(TOTCHOL >= 240, "Chol≥240", "Chol<240"),
    BMI_cat = if_else(BMI >= 30, "BMI≥30", "BMI<30")
    ) %>%
  select(TIMESTRKorCENS_10y, STROKE_10y, SBP_cat, DIABETES, PREVCHD, AGE_cat) %>%
  pivot_longer(
    cols = -c(TIMESTRKorCENS_10y, STROKE_10y),
    names_to = "variable",
    values_to = "group") %>% 
    mutate(
    variable = factor(variable,
      levels = c("AGE_cat", "SBP_cat", "DIABETES", "PREVCHD"),
      labels = c("Age", "Systolic BP", "Diabetes", "Prior CHD")
    ))

# fit survival object
fit <- survfit(Surv(TIMESTRKorCENS_10y, STROKE_10y == "Yes") ~ group + variable, 
  data = fmhst_plot)

# plot
p <- ggsurvplot_facet(
  fit,
  data = fmhst_plot,
  facet.by = "variable",
  conf.int = FALSE,
  ylim = c(0.97,1),
  xlab = "Time (Days)",
  ylab = "Stroke-free probability",
  ggtheme = theme_bw())

p + theme(legend.position = "bottom")

```

```{r}

# create grouped versions of variables
fmhst_plot2 <- fmhst_filt %>%
  mutate(
    SBP_cat = if_else(SYSBP >= 140, "BP≥140", "BP<140"),
    AGE_cat = if_else(AGE >= 60, "Age≥60", "Age<60"),
    TOTCHOL_cat = if_else(TOTCHOL >= 240, "Chol≥240", "Chol<240"),
    BMI_cat = if_else(BMI >= 30, "BMI≥30", "BMI<30")
    ) %>%
  select(TIMESTRKorCENS_10y, STROKE_10y, TOTCHOL_cat, BMI_cat, BPMEDS, CURSMOKE) %>%
  pivot_longer(
    cols = -c(TIMESTRKorCENS_10y, STROKE_10y),
    names_to = "variable",
    values_to = "group") %>% 
    mutate(
    variable = factor(variable,
      levels = c("TOTCHOL_cat", "BMI_cat", "BPMEDS", "CURSMOKE"),
      labels = c("Total Chol", "BMI", "BP Meds", "Smoker")
    ))

# fit survival object
fit2 <- survfit(Surv(TIMESTRKorCENS_10y, STROKE_10y == "Yes") ~ group + variable, 
  data = fmhst_plot2)

# plot
p2 <- ggsurvplot_facet(
  fit2,
  data = fmhst_plot2,
  facet.by = "variable",
  conf.int = FALSE,
  ylim = c(0.97,1),
  xlab = "Time (Days)",
  ylab = "Stroke-free probability",
  ggtheme = theme_bw())

p2 + theme(legend.position = "bottom")

```



### Distributions of Time to Event by Sex

```{r}
# time-to-event variables and labels
# time_vars <- c(
#   TIMEHYP  = "Days Baseline-Inc Hypertension",
#   TIMEAP   = "Days Baseline-Inc Angina",
#   TIMEMI   = "Days Baseline-Inc Hosp MI",
#   TIMEMIFC = "Days Baseline-Inc MI-Fatal CHD",
#   TIMESTRK = "Days Baseline-Inc Stroke",
#   TIMECHD  = "Days Baseline-Inc Any CHD",
#   TIMECVD  = "Days Baseline-Inc CVD",
#   TIMEDTH  = "Days Baseline-Death")

time_vars <- c(TIMESTRKorCENS_10y = "Follow-up time (days)")

# build summary table
time_tbl <- fmhst_filt %>%
  group_by(SEX) %>%
  summarise(
    `Time to Event` = "Follow-up time (days)",
    N = sum(!is.na(TIMESTRKorCENS_10y)),
    NMiss = sum(is.na(TIMESTRKorCENS_10y)),
    Mean = mean(TIMESTRKorCENS_10y, na.rm = TRUE),
    Std = sd(TIMESTRKorCENS_10y, na.rm = TRUE),
    Min = min(TIMESTRKorCENS_10y, na.rm = TRUE),
    P25 = quantile(TIMESTRKorCENS_10y, 0.25, na.rm = TRUE),
    Median = median(TIMESTRKorCENS_10y, na.rm = TRUE),
    P75 = quantile(TIMESTRKorCENS_10y, 0.75, na.rm = TRUE),
    Max = max(TIMESTRKorCENS_10y, na.rm = TRUE),
    .groups = "drop") %>%
  # indent labels within sex groups
  mutate(SEX = if_else(duplicated(SEX), "", as.character(SEX))) %>%
  gt(groupname_col = "SEX") %>%
  tab_header(title = "Distributions of Time to Event by Sex") %>%
  cols_label(
    `Time to Event` = "Time to Event",
    N = "N",
    NMiss = "NMiss",
    Mean = "Mean",
    Std = "Std",
    Min = "Min",
    P25 = "P25",
    Median = "Median",
    P75 = "P75",
    Max = "Max") %>%
  fmt_number(
    columns = c(Mean, Std, Min, P25, Median, P75, Max),
    decimals = 0) %>%
  cols_align(
    align = "left",
    columns = c(`Time to Event`)) %>%
  cols_align(
    align = "center",
    columns = c(N, NMiss, Mean, Std, Min, P25, Median, P75, Max)) %>%
  tab_options(
    table.font.size = px(16),
    heading.align = "left",
    data_row.padding = px(6),
    row_group.font.weight = "normal",
    column_labels.font.weight = "normal",
    table.border.top.width = px(1),
    table.border.bottom.width = px(1))

time_tbl
```


## Primary Analysis
```{r}
cox.male.model <- coxph(
  Surv(TIMESTRKorCENS_10y, STROKE_bin) ~ AGE + SYSBP + DIABETES +  
  TOTCHOL + BMI + CURSMOKE + BPMEDS + PREVCHD, 
                   data = subset(fmhst_filt, SEX == "Men"))
summary(cox.male.model)
```

Proportional assumption - low p-values indicates variable has time dependency
"the proportionality assumption does not check linearity - the Cox PH model is 
semi-parametric and thus makes no assumption as to the form of the hazard. 
The proportionality assumption is that the hazard rate of an individual is 
relatively constant in time"
  * semi-parametric = don't have to choose probability dist'n for survival time
```{r}
cox.zph(cox.male.model)
plot(cox.zph(cox.male.model))
```


```{r}
# # different arbitrary risk profiles
# risk_profiles <- data.frame(
#   profile_name = c("Low Risk", "Moderate Risk", "High Risk"),
#   AGE = c(55, 65, 75),
#   DIABETES = c(0, 0, 1),
#   SYSBP = c(120, 140, 160),
#   CURSMOKE = c(0, 1, 1),
#   BMI = c(25, 28, 32))

# baseline survival at 10 years 
baseline_surv <- survfit(cox.male.model)
summary(baseline_surv, times = 10*365.25)

# survival to probability of stroke
#risk_profiles$prob_stroke_10yr <- 1 - summary_baseline$surv
```




## Secondary Analysis

"how much change there is in the risk factors over in the 3 periods of data collection"
```{r}
# # change across periods
# rf_change <- fmhst_filt %>%
#   group_by(SEX) %>%
#   summarise(
#     # SYSBP
#     SYSBP_p1_med = median(SYSBP_PERIOD1, na.rm = TRUE),
#     SYSBP_p2_med = median(SYSBP_PERIOD2, na.rm = TRUE),
#     SYSBP_p3_med = median(SYSBP_PERIOD3, na.rm = TRUE),
#     SYSBP_change_abs = SYSBP_p3_med - SYSBP_p1_med,
#     SYSBP_change_pct = (SYSBP_change_abs / SYSBP_p1_med) * 100,
#     
#     # BMI
#   )
```













