---
title: 'BIOS 6624: Project 3'
author: "Ryan Summers"
date: "2026-04-01"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(magrittr)
library(dplyr)
library(purrr)
library(tibble)
library(tidyr)
library(broom)
library(gt)
library(gtsummary)
library(survival)
library(survminer)
library(olsrr)
library(ncvreg)
library(StepReg)
```

## Load Data

```{r}
fmhst_raw <- read_csv("/Users/ryan_summers/Library/Mobile Documents/com~apple~CloudDocs/GitHub/BIOS6624-Class/Project3/DataRaw/frmgham2.csv")
```

## Data Management & Cleanup

Remove subjects who had event (stroke) at baseline and Filter to 10yrs cutoff.
Calculate censored indicator if event doesn't happen within 10yrs and calculate
time to censoring variable.

```{r}
cutoff_days <- 10*365.25

fmhst_filt <- fmhst_raw %>% 
  # remove subjects who had event prior to baseline assessment
  filter(PERIOD == 1, PREVSTRK == 0) %>%   # baseline cohort
  mutate(
    # time to event or being censored truncated at 10 years
    TIMESTRKorCENS_10y = pmin(TIMESTRK, cutoff_days),

    # event indicator within 10 years
    STROKE_10y = if_else(STROKE == 1 & TIMESTRK <= cutoff_days, 1, 0),
    
    ## all censored subjects
    ## **TIMESTRK includes censored participants if STROKE==0**
    CENSORED_10y = if_else(STROKE_10y == 0, 1, 0),
    
    # convert variables to factors
    SEX = factor(SEX, levels = c(1,2), labels = c("Men", "Women")),
    DIABETES = factor(DIABETES, levels = c(0,1), labels = c("No", "Yes")),
    PREVCHD = factor(PREVCHD, levels = c(0,1), labels = c("No", "Yes")),
    BPMEDS = factor(BPMEDS, levels = c(0,1), labels = c("No", "Yes")),
    CURSMOKE = factor(CURSMOKE, levels = c(0,1), labels = c("No", "Yes")),
    STROKE_10y = factor(STROKE_10y, levels = c(0,1), labels = c("No-Censored", "Yes")),
    STROKE_bin = ifelse(STROKE_10y == "Yes", 1, 0),
    CENSORED_10y = factor(CENSORED_10y, levels = c(0, 1), labels = c("No", "Yes")),
    TIMESTRKorCENS_10y_yrs = TIMESTRKorCENS_10y/365.25,
    )
```

We restricted the primary analysis to baseline risk factors measured at period 1 and 
modeled incident stroke over the first 10 years of follow-up. This approach aligns 
predictor measurement more closely with the risk window of interest. We additionally 
examined how major risk factors changed across the three examination periods to assess whether a time-varying covariate analysis might be informative. If risk factors change materially over time, 
a baseline-only model may understate or misclassify the relationship between current 
risk-factor status and stroke hazard. However, time-varying covariate analyses in 
observational data may also be harder to interpret because updated covariates can be
influenced by treatment changes and evolving disease status.

## Exploratory Data Analysis (EDA)

### Table1

```{r label = "Used in final report as Table 1"}
tbl1 <- fmhst_filt %>%
  tbl_summary(
    by = SEX,
    include = c(
      AGE, SYSBP, DIABETES, TOTCHOL, PREVCHD, BPMEDS, CURSMOKE,
      BMI, STROKE_10y),
    type = list(
      all_continuous() ~ "continuous2",
      all_categorical() ~ "categorical"),
    statistic = list(
      all_continuous() ~ c(
        "{mean} ({sd})",
        "{median} ({p25}, {p75})",
        "{min} - {max}"),
      all_categorical() ~ "{n} ({p}%)"),
    digits = list(AGE ~ 0, SYSBP ~ 0),
    missing = "ifany",
    missing_text = "NA",
    missing_stat = "{N_miss} ({p_miss}%)",
    label = list(
      AGE ~ "Age",
      SYSBP ~ "Systolic BP",
      DIABETES ~ "Diabetes",
      TOTCHOL ~ "Total Cholesterol (mg/dL)",
      PREVCHD ~ "Prevalent Coronary Heart Disease",
      BPMEDS ~ "BP Meds",
      CURSMOKE ~ "Smoker",
      STROKE_10y ~ "Event (Stroke)")) %>%
  add_overall() %>% bold_labels() %>% italicize_levels() %>% add_n() %>%
  as_gt() %>%
  gt::fmt_markdown(columns = everything()) %>%
  tab_header(
    title = "Characteristics of the Framingham Heart Study Cohort") %>% 
  tab_options(
    table.width = pct(70),
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

tbl1
```

```{r}
# number of participants
cat("Number of participants in entire cohort", 
   length(unique(fmhst_raw$RANDID)), "\n\n")

cat("Number of participants in 10yr subset cohort", 
   length(unique(fmhst_filt$RANDID)), "\n\n")
```


```{r}
hist(fmhst_filt$TOTCHOL)
```



```{r}
# crude incidence
fmhst_filt %>%
  group_by(SEX) %>%
  summarise(
    events = sum(STROKE_10y == "Yes"),
    total = n(),
    risk = round( (events / total)*100,1 ))
```

Correlations
```{r}
library(ggcorrplot)
library(rcompanion)

corr_vars <- c("AGE", "SYSBP", "TOTCHOL", "BMI")

cor_matrix <- cor(
  fmhst_filt[, corr_vars],
  use = "pairwise.complete.obs",
  method = "spearman")

ggcorrplot(cor_matrix,
           type = "lower",
           lab = T)

tab <- table(fmhst_filt$DIABETES, fmhst_filt$PREVCHD)
cramerV(tab)
```



### Kaplan-Meier Curve

```{r}
make_stroke_plot <- function(data, sex, vars, title = NULL) {
  plot_data <- data %>%
    filter(SEX == sex) %>%
    mutate(
      SBP_cat     = if_else(SYSBP >= 160, "BPâ‰Ą160", "BP<160"),
      AGE_cat     = if_else(AGE >= 60, "Ageâ‰Ą60", "Age<60"),
      TOTCHOL_cat = if_else(TOTCHOL >= 240, "Cholâ‰Ą240", "Chol<240"),
      BMI_cat     = if_else(BMI >= 30, "BMIâ‰Ą30", "BMI<30")) %>%
    select(TIMESTRKorCENS_10y, STROKE_10y, all_of(names(vars))) %>%
    pivot_longer(
      cols = -c(TIMESTRKorCENS_10y, STROKE_10y),
      names_to = "variable",
      values_to = "group") %>%
    mutate(
      variable = factor(variable,
        levels = names(vars),
        labels = unname(vars)))

  fit <- survfit(
    Surv(TIMESTRKorCENS_10y, STROKE_10y == "Yes") ~ group + variable,
    data = plot_data)

  p <- ggsurvplot_facet(
    fit,
    data = plot_data,
    facet.by = "variable",
    conf.int = FALSE,
    ylim = c(0.97, 1),
    xlab = "Time (Days)",
    ylab = "Stroke-free probability",
    title = title,
    ggtheme = theme_bw() )

  p <- p + theme(legend.position = "bottom")
  p
}
```


```{r}

vars1 <- c(
  AGE_cat  = "Age",
  SBP_cat  = "Systolic BP",
  DIABETES = "Diabetes",
  PREVCHD  = "Prior CHD")

vars2 <- c(
  TOTCHOL_cat = "Total Chol",
  BMI_cat     = "BMI",
  BPMEDS      = "BP Meds",
  CURSMOKE    = "Smoker")


p_women1 <- make_stroke_plot(
  data = fmhst_filt,
  sex = "Women",
  vars = vars1,
  title = "Stroke-free survival: Women")

p_women2 <- make_stroke_plot(
  data = fmhst_filt,
  sex = "Women",
  vars = vars2,
  title = "Stroke-free survival: Women")

p_men1 <- make_stroke_plot(
  data = fmhst_filt,
  sex = "Men",
  vars = vars1,
  title = "Stroke-free survival: Men")

p_men2 <- make_stroke_plot(
  data = fmhst_filt,
  sex = "Men",
  vars = vars2,
  title = "Stroke-free survival: Men")

p_women1
p_women2

p_men1
p_men2
```



### Distributions of Time to Event by Sex

```{r}
# time-to-event variables and labels
# time_vars <- c(
#   TIMEHYP  = "Days Baseline-Inc Hypertension",
#   TIMEAP   = "Days Baseline-Inc Angina",
#   TIMEMI   = "Days Baseline-Inc Hosp MI",
#   TIMEMIFC = "Days Baseline-Inc MI-Fatal CHD",
#   TIMESTRK = "Days Baseline-Inc Stroke",
#   TIMECHD  = "Days Baseline-Inc Any CHD",
#   TIMECVD  = "Days Baseline-Inc CVD",
#   TIMEDTH  = "Days Baseline-Death")

time_vars <- c(TIMESTRKorCENS_10y = "Follow-up time (days)")

# build summary table
time_tbl <- fmhst_filt %>%
  group_by(SEX) %>%
  summarise(
    `Time to Event` = "Follow-up time (days)",
    N = sum(!is.na(TIMESTRKorCENS_10y)),
    NMiss = sum(is.na(TIMESTRKorCENS_10y)),
    Mean = mean(TIMESTRKorCENS_10y, na.rm = TRUE),
    Std = sd(TIMESTRKorCENS_10y, na.rm = TRUE),
    Min = min(TIMESTRKorCENS_10y, na.rm = TRUE),
    P25 = quantile(TIMESTRKorCENS_10y, 0.25, na.rm = TRUE),
    Median = median(TIMESTRKorCENS_10y, na.rm = TRUE),
    P75 = quantile(TIMESTRKorCENS_10y, 0.75, na.rm = TRUE),
    Max = max(TIMESTRKorCENS_10y, na.rm = TRUE),
    .groups = "drop") %>%
  # indent labels within sex groups
  mutate(SEX = if_else(duplicated(SEX), "", as.character(SEX))) %>%
  gt(groupname_col = "SEX") %>%
  tab_header(title = "Distributions of Time to Event by Sex") %>%
  cols_label(
    `Time to Event` = "Time to Event",
    N = "N",
    NMiss = "NMiss",
    Mean = "Mean",
    Std = "Std",
    Min = "Min",
    P25 = "P25",
    Median = "Median",
    P75 = "P75",
    Max = "Max") %>%
  fmt_number(
    columns = c(Mean, Std, Min, P25, Median, P75, Max),
    decimals = 0) %>%
  cols_align(
    align = "left",
    columns = c(`Time to Event`)) %>%
  cols_align(
    align = "center",
    columns = c(N, NMiss, Mean, Std, Min, P25, Median, P75, Max)) %>%
  tab_options(
    table.font.size = px(16),
    heading.align = "left",
    data_row.padding = px(6),
    row_group.font.weight = "normal",
    column_labels.font.weight = "normal",
    table.border.top.width = px(1),
    table.border.bottom.width = px(1))

time_tbl
```


## Primary Analysis

### Variable Selection

Proportional assumption - low p-values indicates variable has time dependency
"the proportionality assumption does not check linearity - the Cox PH model is 
semi-parametric and thus makes no assumption as to the form of the hazard. 
The proportionality assumption is that the hazard rate of an individual is 
relatively constant in time"
  * semi-parametric = don't have to choose prob dist'n for surv time

```{r}
# stratify datae in women and men
vars <- c("AGE", "SYSBP", "DIABETES", "TOTCHOL",
          "BMI", "CURSMOKE", "BPMEDS", "PREVCHD",
          "TIMESTRKorCENS_10y", "STROKE_bin")

men_dat <- subset(fmhst_filt, SEX == "Men", select = vars)
men_dat <- na.omit(men_dat)

women_dat <- subset(fmhst_filt, SEX == "Women", select = vars)
women_dat <- na.omit(women_dat)

```

#### Fit model (Men)
```{r}
cox.men.model <- coxph(
  Surv(TIMESTRKorCENS_10y, STROKE_bin) ~ AGE + SYSBP + DIABETES + 
    # variables to be determined
    TOTCHOL + BMI + CURSMOKE + BPMEDS + PREVCHD, 
  data = men_dat)

summary(cox.men.model)
```

```{r}
cox.zph(cox.men.model)
plot(cox.zph(cox.men.model))
```

StepReg:

```{r}
form <- Surv(TIMESTRKorCENS_10y, STROKE_bin) ~ AGE + SYSBP + DIABETES + 
    TOTCHOL + BMI + CURSMOKE + BPMEDS + PREVCHD

step.fit.men <- stepwise(formula = form,
                data = men_dat,
                type = "cox",
                include = c("AGE", "DIABETES", "SYSBP"),
                strategy = "backward",
                metric = c("AICc", "SBC", "HQ"))

summary(step.fit.men$backward$AICc)
summary(step.fit.men$backward$SBC)
summary(step.fit.men$backward$HQ)
performance(step.fit.men)
```

```{r}
summary(step.fit.men$backward$AICc)
summary(step.fit.men$backward$SL)
```

```{r}
plot(step.fit.men, strategy = "backward", process = "overview")
plot(step.fit.men, strategy = "backward", process = "detail")
```

Plot details of stepwise
```{r}
plot_list <- setNames(
  lapply(c("backward"),function(i){
    setNames(
      lapply(c("detail","overview"),function(j){
        plot(step.fit.men, strategy=i, process=j)
    }),
    c("detail","overview"))
  }),
  c("backward"))

cowplot::plot_grid(plotlist = plot_list$backward, ncol = 1, 
                   rel_heights = c(2, 1))
```


```{r}
step.men <- step(cox.men.model, 
     direction = "backward", 
     scope = list(lower = ~ AGE + SYSBP + DIABETES))
summary(step.men)
```

#### Fit model (Women)
```{r}
cox.women.model <- coxph(
  Surv(TIMESTRKorCENS_10y, STROKE_bin) ~ AGE + SYSBP + DIABETES + 
    # variables to be determined
    TOTCHOL + BMI + CURSMOKE + BPMEDS + PREVCHD, 
  data = women_dat)

summary(cox.women.model)
```

```{r}
step.fit.women <- stepwise(formula = form,
                data = women_dat,
                type = "cox",
                include = c("AGE", "DIABETES", "SYSBP"),
                strategy = "backward",
                metric = c("AICc", "SBC", "HQ"))

summary(step.fit.women$backward$AICc)
summary(step.fit.women$backward$SBC)
summary(step.fit.women$backward$HQ)
performance(step.fit.women)
```

```{r}
plot_list2 <- setNames(
  lapply(c("backward"), function(i){
    setNames(
      lapply(c("detail","overview"), function(j){
        plot(step.fit.women, strategy=i, process=j)
    }),
    c("detail","overview"))
  }),
  c("backward"))

cowplot::plot_grid(plotlist = plot_list2$backward, ncol = 2, 
                   rel_heights = c(2, 1))
```

```{r}
step.women <- step(cox.women.model, 
     direction = "backward", 
     scope = list(lower = ~ AGE + SYSBP + DIABETES))
summary(step.women)
```

```{r}
#amount of women on BP meds
fmhst_filt %>% 
  filter(SEX=="Women") %>% 
  group_by(BPMEDS) %>% 
  summarize(n = n())
```



## Secondary Analysis

"how much change there is in the risk factors over in the 3 periods of data collection"
```{r}
# # change across periods
# rf_change <- fmhst_filt %>%
#   group_by(SEX) %>%
#   summarise(
#     # SYSBP
#     SYSBP_p1_med = median(SYSBP_PERIOD1, na.rm = TRUE),
#     SYSBP_p2_med = median(SYSBP_PERIOD2, na.rm = TRUE),
#     SYSBP_p3_med = median(SYSBP_PERIOD3, na.rm = TRUE),
#     SYSBP_change_abs = SYSBP_p3_med - SYSBP_p1_med,
#     SYSBP_change_pct = (SYSBP_change_abs / SYSBP_p1_med) * 100,
#     
#     # BMI
#   )
```













