---
title: 'BIOS 6624: Project 3'
author: "Ryan Summers"
date: "2026-04-01"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(magrittr)
library(dplyr)
library(purrr)
library(tibble)
library(tidyr)
library(broom)
library(gt)
library(gtsummary)
library(survival)
library(survminer)
library(olsrr)
library(ncvreg)
library(StepReg)
library(ggcorrplot)
library(rcompanion)
```

## Load Data

```{r}
fmhst_raw <- read_csv("/Users/ryan_summers/Library/Mobile Documents/com~apple~CloudDocs/GitHub/BIOS6624-Class/Project3/DataRaw/frmgham2.csv")
```

## Data Management & Cleanup

Remove subjects who had event (stroke) at baseline and Filter to 10yrs cutoff.
Calculate censored indicator if event doesn't happen within 10yrs and calculate
time to censoring variable.

```{r}
cutoff_days <- 10*365.25

fmhst_filt <- fmhst_raw %>% 
  # remove subjects who had event prior to baseline assessment
  filter(PERIOD == 1, PREVSTRK == 0) %>%   # baseline cohort
  mutate(
    # time to event or being censored truncated at 10 years
    TIMESTRKorCENS_10y = pmin(TIMESTRK, cutoff_days),

    # event indicator within 10 years
    STROKE_10y = if_else(STROKE == 1 & TIMESTRK <= cutoff_days, 1, 0),
    
    ## all censored subjects
    ## **TIMESTRK includes censored participants if STROKE==0**
    CENSORED_10y = if_else(STROKE_10y == 0, 1, 0),
    
    # convert variables to factors
    SEX = factor(SEX, levels = c(1,2), labels = c("Men", "Women")),
    DIABETES = factor(DIABETES, levels = c(0,1), labels = c("No", "Yes")),
    PREVCHD = factor(PREVCHD, levels = c(0,1), labels = c("No", "Yes")),
    BPMEDS = factor(BPMEDS, levels = c(0,1), labels = c("No", "Yes")),
    CURSMOKE = factor(CURSMOKE, levels = c(0,1), labels = c("No", "Yes")),
    STROKE_10y = factor(STROKE_10y, levels = c(0,1), labels = c("No-Censored", "Yes")),
    STROKE_bin = ifelse(STROKE_10y == "Yes", 1, 0),
    CENSORED_10y = factor(CENSORED_10y, levels = c(0, 1), labels = c("No", "Yes")),
    TIMESTRKorCENS_10y_yrs = TIMESTRKorCENS_10y/365.25,
    ) %>% 
  mutate(
    SYSBP_10 = SYSBP / 10,
    TOTCHOL_10 = TOTCHOL / 10)
```

We restricted the primary analysis to baseline risk factors measured at period 1 and 
modeled incident stroke over the first 10 years of follow-up. This approach aligns 
predictor measurement more closely with the risk window of interest. We additionally 
examined how major risk factors changed across the three examination periods to 
assess whether a time-varying covariate analysis might be informative. If risk 
factors change materially over time,  a baseline-only model may understate or 
misclassify the relationship between current risk-factor status and stroke hazard. 
However, time-varying covariate analyses in observational data may also be harder
to interpret because updated covariates can be influenced by treatment changes and 
evolving disease status.

## Exploratory Data Analysis (EDA)

### Table1

```{r label = "Used in final report as Table 1"}
tbl1 <- fmhst_filt %>%
  tbl_summary(
    by = SEX,
    include = c(
      AGE, SYSBP, DIABETES, TOTCHOL, PREVCHD, BPMEDS, CURSMOKE,
      BMI, STROKE_10y),
    type = list(
      all_continuous() ~ "continuous2",
      all_categorical() ~ "categorical"),
    statistic = list(
      all_continuous() ~ c(
        "{mean} ({sd})",
        "{median} ({p25}, {p75})",
        "{min} - {max}"),
      all_categorical() ~ "{n} ({p}%)"),
    digits = list(AGE ~ 0, SYSBP ~ 0),
    missing = "ifany",
    missing_text = "NA",
    missing_stat = "{N_miss} ({p_miss}%)",
    label = list(
      AGE ~ "Age",
      SYSBP ~ "Systolic BP",
      DIABETES ~ "Diabetes",
      TOTCHOL ~ "Total Cholesterol (mg/dL)",
      PREVCHD ~ "Prevalent Coronary Heart Disease",
      BPMEDS ~ "BP Meds",
      CURSMOKE ~ "Smoker",
      STROKE_10y ~ "Event (Stroke)")) %>%
  add_overall() %>% bold_labels() %>% italicize_levels() %>% add_n() %>%
  as_gt() %>%
  gt::fmt_markdown(columns = everything()) %>%
  tab_header(
    title = "Characteristics of the Framingham Heart Study Cohort") %>% 
  tab_options(
    table.width = pct(70),
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

tbl1
```

Compact version (Mean vs Median??)

```{r}
tbl1 <- fmhst_filt %>%
  tbl_summary(
    by = SEX,
    include = c(
      AGE, SYSBP, DIABETES, TOTCHOL, PREVCHD, BPMEDS, CURSMOKE, BMI),
    type = list(
      all_continuous() ~ "continuous",
      all_categorical() ~ "categorical"),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"),
    digits = list(
      AGE ~ 0,
      SYSBP ~ 0,
      TOTCHOL ~ 0,
      BMI ~ 1),
    missing = "ifany",
    missing_text = "Missing",
    missing_stat = "{N_miss} ({p_miss}%)",
    label = list(
      AGE ~ "Age",
      SYSBP ~ "Systolic BP",
      DIABETES ~ "Diabetes",
      TOTCHOL ~ "Total Cholesterol (mg/dL)",
      PREVCHD ~ "Cardiovascular Disease",
      BPMEDS ~ "BP Meds",
      CURSMOKE ~ "Smoker",
      BMI ~ "BMI")) %>%
  add_overall() %>%
  bold_labels() %>%
  italicize_levels() %>%
  add_n() %>%
  as_gt() %>%
  gt::fmt_markdown(columns = everything()) %>%
  tab_header(
    title = "Characteristics of the Framingham Heart Study Cohort") %>%
  tab_options(
    table.width = pct(70),
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

tbl1
```


```{r}
tbl1 <- fmhst_filt %>%
  tbl_summary(
    by = SEX,
    include = c(
      AGE, SYSBP, DIABETES, TOTCHOL, PREVCHD, BPMEDS, CURSMOKE, BMI
    ),
    type = list(
      all_continuous() ~ "continuous",
      all_categorical() ~ "categorical"
    ),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(
      AGE ~ 0,
      SYSBP ~ 0,
      TOTCHOL ~ 0,
      BMI ~ 1
    ),
    missing = "no",
    label = list(
      AGE ~ "Age",
      SYSBP ~ "Systolic BP",
      DIABETES ~ "Diabetes",
      TOTCHOL ~ "Total Cholesterol (mg/dL)",
      PREVCHD ~ "Prevalent CHD",
      BPMEDS ~ "BP Meds",
      CURSMOKE ~ "Smoker",
      BMI ~ "BMI")) %>%
  add_overall() %>%
  add_n() %>%
  add_stat(
    fns = everything() ~ function(data, variable, ...) {
      x <- data[[variable]]
      paste0(sum(is.na(x)), " (", round(mean(is.na(x)) * 100, 1), "%)")
    },
    location = everything() ~ "label") %>%
  modify_header(add_stat_1 ~ "Missing") %>%
  modify_table_body(
    ~ dplyr::relocate(.x, add_stat_1, .after = n)) %>%
  bold_labels() %>%
  italicize_levels() %>%
  as_gt() %>% 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(columns = add_stat_1)) %>% 
  gt::fmt_markdown(columns = everything()) %>%
  tab_header(
    title = "Characteristics of the Framingham Heart Study Cohort") %>%
  tab_options(
    table.width = pct(70),
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3))

tbl1
```



```{r}
# number of participants
cat("Number of participants in entire cohort", 
   length(unique(fmhst_raw$RANDID)), "\n\n")

cat("Number of participants in 10yr subset cohort", 
   length(unique(fmhst_filt$RANDID)), "\n\n")
```


```{r}
hist(fmhst_filt$TOTCHOL)
```



```{r}
# crude incidence
fmhst_filt %>%
  group_by(SEX) %>%
  summarise(
    events = sum(STROKE_10y == "Yes"),
    total = n(),
    risk = round( (events / total)*100,1 ))
```

### Correlations & Associations
```{r}
corr_vars <- c("AGE", "SYSBP", "TOTCHOL", "BMI")

cor_matrix <- cor(
  fmhst_filt[, corr_vars],
  use = "pairwise.complete.obs",
  method = "spearman")

ggcorrplot(cor_matrix,
           type = "lower",
           lab = T)

tab <- table(fmhst_filt$DIABETES, fmhst_filt$PREVCHD)
cramerV(tab)
```

Test association of BP and BP medication use btw Women & Men
```{r}
fmhst_complete <- fmhst_filt %>% filter(!is.na(SYSBP), !is.na(BPMEDS))
t.test(log(SYSBP) ~ BPMEDS, data = subset(fmhst_complete, SEX=="Women"))
t.test(log(SYSBP) ~ BPMEDS, data = subset(fmhst_complete, SEX=="Men"))
ggplot(fmhst_complete, aes(x = BPMEDS, y = log(SYSBP), fill = BPMEDS)) +
  facet_wrap(~SEX) +
  geom_boxplot() + theme_bw()
```


### Kaplan-Meier Curve

```{r}
make_stroke_plot <- function(data, sex, vars, title = NULL) {
  plot_data <- data %>%
    filter(SEX == sex) %>%
    mutate(
      SBP_cat     = if_else(SYSBP >= 160, "BP≥160", "BP<160"),
      AGE_cat     = if_else(AGE >= 60, "Age≥60", "Age<60"),
      TOTCHOL_cat = if_else(TOTCHOL >= 240, "Chol≥240", "Chol<240"),
      BMI_cat     = if_else(BMI >= 30, "BMI≥30", "BMI<30")) %>%
    select(TIMESTRKorCENS_10y, STROKE_10y, all_of(names(vars))) %>%
    pivot_longer(
      cols = -c(TIMESTRKorCENS_10y, STROKE_10y),
      names_to = "variable",
      values_to = "group") %>%
    mutate(
      variable = factor(variable,
        levels = names(vars),
        labels = unname(vars)))

  fit <- survfit(
    Surv(TIMESTRKorCENS_10y, STROKE_10y == "Yes") ~ group + variable,
    data = plot_data)

  p <- ggsurvplot_facet(
    fit,
    data = plot_data,
    facet.by = "variable",
    conf.int = F,
    # pval = T,
    # pval.method = T,
    ylim = c(0.97, 1),
    xlab = "Time (Days)",
    ylab = "Stroke-free probability",
    title = title,
    ggtheme = theme_bw() )

  p <- p + theme(legend.position = "bottom")
  p
}
```


```{r}

vars1 <- c(
  AGE_cat  = "Age",
  SBP_cat  = "Systolic BP",
  DIABETES = "Diabetes",
  PREVCHD  = "Prior CHD")

vars2 <- c(
  TOTCHOL_cat = "Total Chol",
  BMI_cat     = "BMI",
  BPMEDS      = "BP Meds",
  CURSMOKE    = "Smoker")


p_women1 <- make_stroke_plot(
  data = fmhst_filt,
  sex = "Women",
  vars = vars1,
  title = "Stroke-free survival: Women")

p_women2 <- make_stroke_plot(
  data = fmhst_filt,
  sex = "Women",
  vars = vars2,
  title = "Stroke-free survival: Women")

p_men1 <- make_stroke_plot(
  data = fmhst_filt,
  sex = "Men",
  vars = vars1,
  title = "Stroke-free survival: Men")

p_men2 <- make_stroke_plot(
  data = fmhst_filt,
  sex = "Men",
  vars = vars2,
  title = "Stroke-free survival: Men")

p_women1
p_women2

p_men1
p_men2
```



### Distributions of Time to Event by Sex

```{r}

time_vars <- c(TIMESTRKorCENS_10y = "Follow-up time (days)")

# build summary table
time_tbl <- fmhst_filt %>%
  group_by(SEX) %>%
  summarise(
    `Time to Event` = "Follow-up time (days)",
    N = sum(!is.na(TIMESTRKorCENS_10y)),
    NMiss = sum(is.na(TIMESTRKorCENS_10y)),
    Mean = mean(TIMESTRKorCENS_10y, na.rm = TRUE),
    Std = sd(TIMESTRKorCENS_10y, na.rm = TRUE),
    Min = min(TIMESTRKorCENS_10y, na.rm = TRUE),
    P25 = quantile(TIMESTRKorCENS_10y, 0.25, na.rm = TRUE),
    Median = median(TIMESTRKorCENS_10y, na.rm = TRUE),
    P75 = quantile(TIMESTRKorCENS_10y, 0.75, na.rm = TRUE),
    Max = max(TIMESTRKorCENS_10y, na.rm = TRUE),
    .groups = "drop") %>%
  # indent labels within sex groups
  mutate(SEX = if_else(duplicated(SEX), "", as.character(SEX))) %>%
  gt(groupname_col = "SEX") %>%
  tab_header(title = "Distributions of Time to Event by Sex") %>%
  cols_label(
    `Time to Event` = "Time to Event",
    N = "N",
    NMiss = "NMiss",
    Mean = "Mean",
    Std = "Std",
    Min = "Min",
    P25 = "P25",
    Median = "Median",
    P75 = "P75",
    Max = "Max") %>%
  fmt_number(
    columns = c(Mean, Std, Min, P25, Median, P75, Max),
    decimals = 0) %>%
  cols_align(
    align = "left",
    columns = c(`Time to Event`)) %>%
  cols_align(
    align = "center",
    columns = c(N, NMiss, Mean, Std, Min, P25, Median, P75, Max)) %>%
  tab_options(
    table.font.size = px(16),
    heading.align = "left",
    data_row.padding = px(6),
    row_group.font.weight = "normal",
    column_labels.font.weight = "normal",
    table.border.top.width = px(1),
    table.border.bottom.width = px(1))

time_tbl
```


## Primary Analysis

**Risk Factors**

### Variable Selection

Proportional assumption - low p-values indicates variable has time dependency
"the proportionality assumption does not check linearity - the Cox PH model is 
semi-parametric and thus makes no assumption as to the form of the hazard. 
The proportionality assumption is that the hazard rate of an individual is 
relatively constant in time"
  * semi-parametric = don't have to choose prob dist'n for surv time

```{r}
# stratify datae in women and men
vars <- c("AGE", "SYSBP_10", "DIABETES", "TOTCHOL_10",
          "BMI", "CURSMOKE", "BPMEDS", "PREVCHD",
          "TIMESTRKorCENS_10y", "STROKE_bin")

men_dat <- subset(fmhst_filt, SEX == "Men", select = vars)
men_dat <- na.omit(men_dat)

women_dat <- subset(fmhst_filt, SEX == "Women", select = vars)
women_dat <- na.omit(women_dat)

```

#### Full model (Men)
```{r}
cox.men.model <- coxph(
  Surv(TIMESTRKorCENS_10y, STROKE_bin) ~ AGE + SYSBP_10*BPMEDS + DIABETES + 
    # variables to be determined
    TOTCHOL_10 + BMI + CURSMOKE + PREVCHD, 
  data = men_dat)

summary(cox.men.model)
```


StepReg:

```{r}
form <- Surv(TIMESTRKorCENS_10y, STROKE_bin) ~ AGE + SYSBP_10*BPMEDS + DIABETES + 
    TOTCHOL_10 + BMI + CURSMOKE + PREVCHD

step.fit.men <- stepwise(formula = form,
                data = men_dat,
                type = "cox",
                include = c("AGE", "DIABETES", "SYSBP_10"),
                strategy = "backward",
                metric = c("AICc", "SBC", "HQ"))

summary(step.fit.men$backward$AICc)
summary(step.fit.men$backward$SBC)
summary(step.fit.men$backward$HQ)
performance(step.fit.men)
```

```{r}
summary(step.fit.men$backward$AICc)
summary(step.fit.men$backward$SL)
```

```{r}
plot(step.fit.men, strategy = "backward", process = "overview")
plot(step.fit.men, strategy = "backward", process = "detail")
```

Plot details of stepwise
```{r}
plot_list <- setNames(
  lapply(c("backward"),function(i){
    setNames(
      lapply(c("detail","overview"),function(j){
        plot(step.fit.men, strategy=i, process=j)
    }),
    c("detail","overview"))
  }),
  c("backward"))

cowplot::plot_grid(plotlist = plot_list$backward, ncol = 1, 
                   rel_heights = c(2, 1))
```


```{r}
step.men <- step(cox.men.model, 
     direction = "backward", 
     scope = list(lower = ~ AGE + SYSBP_10 + DIABETES))
summary(step.men)
```

#### Full model (Women)
```{r}
cox.women.model <- coxph(
  Surv(TIMESTRKorCENS_10y, STROKE_bin) ~ AGE + SYSBP_10*BPMEDS + DIABETES + 
    # variables to be determined
    TOTCHOL_10 + BMI + CURSMOKE + PREVCHD, 
  data = women_dat)

summary(cox.women.model)
```

```{r}
step.fit.women <- stepwise(formula = form,
                data = women_dat,
                type = "cox",
                include = c("AGE", "DIABETES", "SYSBP_10"),
                strategy = "backward",
                metric = c("AICc", "SBC", "HQ"))

summary(step.fit.women$backward$AICc)
summary(step.fit.women$backward$SBC)
summary(step.fit.women$backward$HQ)
performance(step.fit.women)
```

```{r}
plot_list2 <- setNames(
  lapply(c("backward"), function(i){
    setNames(
      lapply(c("detail","overview"), function(j){
        plot(step.fit.women, strategy=i, process=j)
    }),
    c("detail","overview"))
  }),
  c("backward"))

cowplot::plot_grid(plotlist = plot_list2$backward, ncol = 2, 
                   rel_heights = c(2, 1))
```

```{r}
step.women <- step(cox.women.model, 
     direction = "backward", 
     scope = list(lower = ~ AGE + SYSBP_10 + DIABETES))
summary(step.women)
```

```{r}
#amount of women on BP meds
fmhst_filt %>% 
  filter(SEX=="Women") %>% 
  group_by(BPMEDS) %>% 
  summarize(n = n())
```

### Final Models



```{r}
cox.women.final <- coxph(
  Surv(TIMESTRKorCENS_10y, STROKE_bin) ~ AGE + SYSBP_10*BPMEDS + DIABETES + 
    CURSMOKE + PREVCHD, 
  data = women_dat)

summary(cox.women.final)

cox.men.final <- coxph(
  Surv(TIMESTRKorCENS_10y, STROKE_bin) ~ AGE + SYSBP_10*BPMEDS + DIABETES + 
    CURSMOKE + PREVCHD, 
  data = men_dat)

summary(cox.men.final)
```
Sample interpretation:
Systolic BP was positively associated with stroke risk in women, with each 1 mmHg increase corresponding to a 2.5% increase in hazard (HR = 1.025; 95% CI: 1.015–1.035). This corresponds to approximately a 28% ($1.025^{10} ≈ 1.28$) increase in hazard per 10 mmHg increase.

Systolic BP was positively associated with stroke risk in men, with each 1 mmHg increase corresponding to a 3.5% increase in hazard (HR = 1.0352; 95% CI: 1.023–1.048). This corresponds to approximately a 41.3% ($1.035^{10} ≈ 1.413$) increase in hazard per 10 mmHg increase.

#### Proportionality Check
```{r}
ph_women <- cox.zph(cox.women.final)
ph_women
par(mfrow = c(3, 3), mar = c(4, 4, 2, 2)) # c(bottom, left, top, right)
plot(ph_women)
```

```{r}
ph_men <- cox.zph(cox.men.final)
ph_men
par(mfrow = c(3, 3), mar = c(4, 4, 2, 2)) # c(bottom, left, top, right)
plot(ph_men)
```


#### Risk Factors Table

```{r}

broom::tidy(cox.men.final, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(
    pct_change = (estimate - 1) * 100,
    pct_label = paste0(
      ifelse(pct_change >= 0, "+", ""),
      round(pct_change, 1), "%"))
```

```{r}

# helper function to generate results table
make_cox_table <- function(model, sex_label) {
  tidy(model, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    mutate(
      # rename terms
      term = recode(
        term,
        "AGE" = "Age (per 1 year)",
        "SYSBP_10" = "Systolic BP (per 10 mmHg)",
        "DIABETESYes" = "Diabetes (Yes vs No)",
        "TOTCHOL" = "Total cholesterol (per 1 mg/dL)",
        "BMI" = "BMI (per 1 kg/m²)",
        "CURSMOKEYes" = "Current smoker (Yes vs No)",
        "BPMEDSYes" = "BP meds (Yes vs No)",
        "PREVCHDYes" = "Cardiovascular Disease (Yes vs No)",
        "SYSBP_10:BPMEDSYes" = "SBP × BP meds interaction"),

      # percent change in hazard
      pct_change = (estimate - 1) * 100,

      # formatted columns
      hr_ci = sprintf("%.2f (%.2f, %.2f; p=%s)",estimate,conf.low,conf.high,
                      format.pval(p.value, digits = 2, eps = 0.01)),
      pct_label = sprintf(
        "%s%.1f%%",
        ifelse(pct_change >= 0, "+", ""),
        pct_change)) %>%
    select(term, hr_ci, pct_label) %>%
    rename(
      !!paste0(sex_label, "_HR_CI") := hr_ci,
      !!paste0(sex_label, "_Pct") := pct_label)
  }

tbl_women <- make_cox_table(cox.women.final, "Women")
tbl_men   <- make_cox_table(cox.men.final, "Men")

# merge 
tbl_final <- full_join(tbl_women, tbl_men, by = "term") %>%
  rename(`Risk factor` = term)

tbl_final
```


```{r label = "Used in final report as Table 2"}
gt_tbl <- tbl_final %>%
  mutate(
    Men_HR_CI = if_else(
      `Risk factor` == "BP meds (Yes vs No)", "\u2020", Men_HR_CI),
    Men_Pct = if_else(`Risk factor`=="BP meds (Yes vs No)","\u2020",Men_Pct)) %>% 
  gt() %>%
  tab_header(
    title = "Hazard Ratios and Percent Change in Stroke Hazard by Sex") %>%
  cols_label(
    `Risk factor` = "Risk factor",
    Men_HR_CI = "HR (95% CI;P-val)",
    Men_Pct = "%Δ in Hazard",
    Women_HR_CI = "HR (95% CI;P-val)",
    Women_Pct = "%Δ in Hazard") %>%
  tab_spanner(label = "Men", columns = c(Men_HR_CI, Men_Pct)) %>%
  tab_spanner(label = "Women", columns = c(Women_HR_CI, Women_Pct)) %>%
  cols_align(align = "left", columns = `Risk factor`) %>%
  cols_align(align = "center",
    columns = c(Men_HR_CI, Men_Pct, Women_HR_CI, Women_Pct)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_spanners()) %>% 
  # by risk factor name
  # tab_style(
  #   style = cell_text(weight = "bold"),
  #   locations = cells_body(columns = `Risk factor`)) %>% 
  tab_footnote(
    footnote = md("\u2020 Because the BP \u00d7 BP medication 
                  interaction was significant in men, the effect of BP meds 
                  depends on BP and is not summarized as a single hazard ratio."),
    locations = cells_body(columns = c(Men_HR_CI, Men_Pct),
      rows = `Risk factor` %in% c("BP meds (Yes vs No)", 
                                  "SBP × BP meds interaction"))) %>% 
  tab_options(
    table.width = pct(88),
    table.font.size = px(13),
    heading.align = "left",
    data_row.padding = px(5),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2)) 

gt_tbl 
```

**10yr probability of stroke based on different risk profiles**

```{r}

risk_profiles <- tibble(
  scenario = c(
    "Baseline",
    "High BP",
    "Diabetes",
    "Diabetes + High BP",
    "Smoker + Diabetes + High BP"),
  AGE = c(40, 40, 40, 40, 40),   # placeholder, replaced below
  SYSBP_10 = c(12, 16, 12, 16, 16),   # 120 and 160 mmHg
  DIABETES = factor(c("No", "No", "Yes", "Yes", "Yes"),
                    levels = c("No", "Yes")),
  CURSMOKE = factor(c("No", "No", "No", "No", "Yes"),
                    levels = c("No", "Yes")),
  PREVCHD = factor(c("No", "No", "No", "No", "No"),
                   levels = c("No", "Yes")),
  BPMEDS = factor(c("No", "No", "No", "No", "No"),
                  levels = c("No", "Yes")))

pred_data <- bind_rows(
  risk_profiles %>% mutate(AGE = 40),
  risk_profiles %>% mutate(AGE = 50),
  risk_profiles %>% mutate(AGE = 60)) %>%
  mutate(age_group = AGE)
```


```{r}
make_preds <- function(model, sex_label, ages = c(40, 50, 60)) {
  out <- lapply(ages, function(a) {
    nd <- risk_profiles %>%
      mutate(AGE = a)

    sf <- survfit(model, newdata = nd)
    surv_10y <- as.numeric(summary(sf, times = 10*365.25, extend = TRUE)$surv)

    nd %>%
      transmute(
        sex = sex_label,
        scenario,
        age = paste("Age", a),
        risk_10y = round(100 * (1 - surv_10y), 2)) #100%
    })
  bind_rows(out)
  }

women_results <- make_preds(cox.women.final, "Women")
men_results   <- make_preds(cox.men.final, "Men")

risk_table <- bind_rows(women_results, men_results) %>%
  pivot_wider(names_from = age, values_from = risk_10y)

risk_table
```

Check that function is behaving correctly
```{r}
women_50_base <- tibble(
  AGE = 50,
  SYSBP_10 = 12,   # 120 mmHg
  DIABETES = factor("No", levels = c("No", "Yes")),
  CURSMOKE = factor("No", levels = c("No", "Yes")),
  PREVCHD = factor("No", levels = c("No", "Yes")),
  BPMEDS = factor("No", levels = c("No", "Yes")))

sf_check <- survfit(cox.women.final, newdata = women_50_base)
1-summary(sf_check, times = 10*365.25, extend = TRUE)$surv
```

#### Risk Profiles Table

```{r}

risk_gt_df <- risk_table %>%
  pivot_longer(cols = starts_with("Age"), names_to = "age",
               values_to = "risk") %>%
  pivot_wider(names_from = sex, values_from = risk) %>%
  rename(Scenario = scenario) %>%
  mutate(
    Scenario = factor(
      Scenario,
      levels = c(
        "Baseline",
        "High BP",
        "Diabetes",
        "Diabetes + High BP",
        "Smoker + Diabetes + High BP")),
    age = factor(age, levels = c("Age 40", "Age 50", "Age 60"))) %>%
  arrange(Scenario, age) %>%
  pivot_wider(
    names_from = age,
    values_from = c(Women, Men),
    names_vary = "slowest")

risk_gt_df
```
```{r label = "Used in final report as Table 3"}
risk_gt <- risk_gt_df %>%
  gt() %>%
  tab_header(
    title = "Predicted 10-Year Probability of Stroke by Risk Profile and Sex") %>%
  cols_label(
    Scenario = "Risk profile",
    `Women_Age 40` = "Women",
    `Men_Age 40` = "Men",
    `Women_Age 50` = "Women",
    `Men_Age 50` = "Men",
    `Women_Age 60` = "Women",
    `Men_Age 60` = "Men") %>%
  tab_spanner(label = "Age 40",columns = c(`Women_Age 40`, `Men_Age 40`)) %>%
  tab_spanner(label = "Age 50",columns = c(`Women_Age 50`, `Men_Age 50`)) %>%
  tab_spanner(label = "Age 60",columns = c(`Women_Age 60`, `Men_Age 60`)) %>%
  fmt_number(columns = -Scenario, decimals = 1) %>%
  cols_align(align = "left", columns = Scenario) %>%
  cols_align( align = "center", columns = -Scenario) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_spanners()) %>% 
  tab_options(
    heading.align = "left",
    table.font.size = px(14),
    data_row.padding = px(5),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

risk_gt
```

Plot the 10yr probs
```{r}

plot_df <- risk_table %>%
  pivot_longer(
    cols = starts_with("Age"),
    names_to = "age",
    values_to = "risk") %>%
  mutate(
    scenario = factor(
      scenario,
      levels = c("Baseline", "High BP", "Diabetes", 
                 "Diabetes + High BP", "Smoker")),
    age = factor(age, levels = c("Age 40", "Age 50", "Age 60")))

ggplot(plot_df, aes(x = scenario, y = risk, fill = age)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  facet_wrap(~ sex) +
    scale_fill_manual(values = c(
    "Age 40" = "#0072B2",
    "Age 50" = "#E69F00",
    "Age 60" = "#009E73")) +
  labs(
    x = NULL,
    y = "Predicted 10-year stroke risk (%)",
    fill = "Age") +
  theme_bw() +
  theme(
    legend.position = "top",
    strip.background = element_rect(fill = "white"),
    axis.text.x = element_text(angle = 20, hjust = 1))
```


## Secondary Analysis

"how much change there is in the risk factors over in the 3 periods of data collection"
```{r}
# # change across periods
# rf_change <- fmhst_filt %>%
#   group_by(SEX) %>%
#   summarise(
#     # SYSBP
#     SYSBP_p1_med = median(SYSBP_PERIOD1, na.rm = TRUE),
#     SYSBP_p2_med = median(SYSBP_PERIOD2, na.rm = TRUE),
#     SYSBP_p3_med = median(SYSBP_PERIOD3, na.rm = TRUE),
#     SYSBP_change_abs = SYSBP_p3_med - SYSBP_p1_med,
#     SYSBP_change_pct = (SYSBP_change_abs / SYSBP_p1_med) * 100,
#     
#     # BMI
#   )
```













