---
title: 'BIOS 6624 : Project 1'
author: "Ryan Summers"
date: "2026-02-04"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
library(BiocManager)
library(knitr)
library(ggplot2)
library(tidyverse)
library(magrittr)
library(ggfortify)
library(olsrr)
library(kableExtra)
library(doBy)
library(MASS)
library(dplyr)
library(table1)
library(gtsummary)
library(gt)
library(stringdist)
library(ggrepel)
library(impute)
library(limma)
library(qvalue)
library(gtools)
library(multtest)
library(GEOquery)
library(Biobase)
library(minfi)
library(R.utils)
library(stringr)
library(qs)
library(qs2)
library(lme4)
library(patchwork)
library(lme4)
library(lmerTest)
library(data.table)
library(emmeans)
library(lubridate)
library(naniar)
library(ggpubr)
library(boot)
library(mediation)
library(tidybayes)
library(priorsense)
library(bayesplot)
library(posterior)
library(brms)
library(bayestestR)
library(purrr)
library(loo)
library(report)
library(bayestestR)
library(tidybayes)

# load scripts
source("/Users/ryan_summers/GitHub/BIOS6624-Class/Project1/Code/group_counts_ftn.R")
```

## Read in the csv

```{r}
HIV_raw <- read_csv("/Users/ryan_summers/GitHub/BIOS6624-Class/Project1/DataRaw/hiv_6624_final.csv",
                    col_select = -1) # col_select = -1 removes first column
```

## Data Management & Cleanup

Convert specific variables to factors and make non-drug users the reference
group for variable of interest
```{r}
var_fct <- c("newid", "SMOKE", "RACE", "EDUCBAS", "hard_drugs")

HIV_raw <- HIV_raw %>% 
  mutate(
    newid = factor(newid),
    SMOKE = factor(SMOKE, levels = c(1,2,3), 
                   labels = c("Never", "Former", "Current")),
    RACE = factor(RACE, 
                  labels = c("White non-Hisp", "White Hisp", "Black non-Hisp",
                             "Black Hisp", "Other", "Other Hisp")),
    EDUCBAS = factor(EDUCBAS, 
                     labels = c("<=8th grade", "9,10,11th grade", "12th grade",
                             ">=1yr college", "4yr college degree", "Some grad", 
                             "Post-grad degree")),
    # non-users will be the reference for modelling
    hard_drugs = factor(hard_drugs, levels = c(0,1),
                        labels = c("No", "Yes")),
    ADH = factor(ADH, 
                 levels = c(1,2,3,4),
                 labels = c("100%", "95-99%", "75-94%", "<75%")))
    
```


Create a new dataset with only baseline and year 2 observations (715 subjects)
```{r}

base <- HIV_raw %>%
  filter(years == 0) %>%
  transmute(
    newid,
    hard0 = hard_drugs,
    age0  = age,
    bmi0  = BMI,
    smoke0 = SMOKE,
    educ0  = EDUCBAS,
    race0  = RACE,
    vload0 = VLOAD,
    cd40   = LEU3N,
    physical0  = AGG_PHYS,
    mental0  = AGG_MENT,
    cesd0 = CESD,
    hbp0 = HBP,
    diab0 = DIAB,
    liv34_0 = LIV34,
    kid0 = KID)


y2 <- HIV_raw %>%
  filter(years == 2) %>%
  transmute(
    newid,
    vload2 = VLOAD,
    cd42   = LEU3N,
    physical2  = AGG_PHYS,
    mental2  = AGG_MENT,
    adh2 = ADH,
    cesd2 = CESD,
    hbp2 = HBP,
    diab2 = DIAB,
    liv34_2 = LIV34,
    kid2 = KID)

ana <- base %>%
  left_join(y2, by="newid") %>% 
    mutate(
    log_vload0 = log10(vload0), # **based on literature log10 is typically used**
    log_vload2 = log10(vload2),
    log_cd40 = log10(cd40),
    log_cd42 = log10(cd42),
    d_log_vload = log_vload2 - log_vload0,
    d_vload = vload2 - vload0,  
    d_cd4 = cd42 - cd40,
    d_phys = physical2 - physical0,
    d_ment = mental2 - mental0)

```


Collapse levels of comorbidities into a simple yes or no category with 9s being converted to NAs. Additionally, for the depression scale (CESD), convert -1 to NAs

```{r label="Collapsing categorical levels"}
# using baseline values
ana2 <- ana %>%
  mutate(
    diab0 = case_when(
      diab0 %in% c(1, 3) ~ "No",
      diab0 %in% c(2, 4) ~ "Yes",
      diab0 == 9 ~ NA_character_,
      TRUE ~ NA_character_) %>% 
      factor(levels = c("No", "Yes")),
    hbp0 = case_when(
      hbp0 %in% c(1, 3) ~ "No",
      hbp0 %in% c(2, 4) ~ "Yes",
      hbp0 == 9 ~ NA_character_,
      TRUE ~ NA_character_) %>% 
      factor(levels = c("No", "Yes")),
    liv34_0 = case_when(
      liv34_0 %in% c(1, 3) ~ "No",
      liv34_0 %in% c(2, 4) ~ "Yes",
      liv34_0 == 9 ~ NA_character_,
      TRUE ~ NA_character_) %>% 
      factor(levels = c("No", "Yes")),
    kid0 = case_when(
      kid0 %in% c(1, 3) ~ "No",
      kid0 %in% c(2, 4) ~ "Yes",
      kid0 == 9 ~ NA_character_,
      TRUE ~ NA_character_) %>% 
      factor(levels = c("No", "Yes")),
    educ0 = case_when(
      educ0 %in% c("<=8th grade", "9,10,11th grade", "12th grade", ">=1yr college") 
      ~ "Didn't finish college or lower",
      educ0 %in% c("4yr college degree", "Some grad", "Post-grad degree") 
      ~ "College degree or higher",
      TRUE ~ educ0) %>%
      factor(),
    race0 = case_when(
      race0 %in% c("White Hisp", "Black Hisp", "Other Hisp", "Other") ~ "Hispanic",
      TRUE ~ race0) %>% 
      factor(),
    smoke0 = case_when(
      smoke0 %in% c("Never", "Former") ~ "Former",
      smoke0 %in% c("Current") ~ "Current",
      TRUE ~ NA_character_) %>% 
      factor(levels = c("Former", "Current")),
    # convert adherence into a binary indicator
      adh_bin = case_when(
      adh2 %in% c("100%","95-99%") ~ 1,  # ≥95%
      adh2 %in% c("75-94%", "<75%") ~ 0,  # <95%
      TRUE ~ NA_real_),
    adh_bin = factor(adh_bin, levels = c(0,1),
                     labels = c("<95%", "≥95%")),
    # insufficient data and implausible values for some BMIs
    bmi0 = ifelse(bmi0 %in% c(-1, 999) | bmi0>100, NA, bmi0),
    cesd0 = ifelse(cesd0==-1, NA_character_, cesd0),
    cesd0 = as.numeric(cesd0))
```


We also need a complete case analysis or we will keep getting "number of observations do not match between mediator and outcome models" error. Also, to be consistent across analyses its probably best we keep the same number of subjects across all models.

**This will be the finalized dataset used for all analyses**
* 463 subjects with complete cases for baseline and year 2 measurements
```{r}
# complete case analysis
ana2_cc <- ana2 %>%
  select(log_vload2, log_vload0, cd42, cd40, hard0, adh2, physical0, physical2,
         mental0, mental2, age0, bmi0, smoke0, educ0, race0, adh_bin) %>%
  na.omit() %>%
  droplevels()

```

## Exploratory Data Analysis (EDA)

Explain how you start with 715 subjects to final N for analyses 
Clinically meaning changes
* .5 log10 VL
* 2pts on QOL
* 50 CD4

Test function
```{r}
# load from starting R code chunk
count_grp_obs(ana, vars = c("hard0", "smoke0"), mode = "interaction")
```




Missing value analysis:
```{r}
# number of NAs
# lapply(HIV_raw, function(x) sum(is.na(x)))
```

```{r}
# missing value percentages
HIV_raw %>%
  ungroup() %>% 
  miss_var_summary()

```

```{r}
# number of subjects
length(unique(HIV_raw$newid))
```

Calculate the number of insufficient data labels for the comorbidities
```{r}
HIV_raw %>%
  filter(years == 0) %>%
  summarise(
    across(
      c(DIAB, HBP, KID, LIV34, FRP, FP),
      ~ sum(. == 9, na.rm = TRUE),
      .names = "n9_{.col}"))
```

Look at distribution of measurements
```{r}
# measurement stats
HIV_raw %>% 
  group_by(newid) %>%
  summarise(n_measurements = n()) -> mnts_per_id 

# summary stats for number of measurements per subject
mnts_per_id %>%   
  summarise(avg = mean(n_measurements),
            min = min(n_measurements),
            med = median(n_measurements),
            max = max(n_measurements))

# distribution of measurements
hist(mnts_per_id$n_measurements, breaks=11, freq = FALSE)
```

```{r}
# percentage distribution of measurements
mnts_per_id %>%
  dplyr::count(n_measurements) %>%
  mutate(percent = round(100 * n / sum(n) , 1))

# cumulative percentages (e.g. % more than 3 measurements=70.1%)
mnts_per_id %>%
  reframe(
    threshold = 1:9,
    percent_gt = sapply(threshold, function(k) {
      mean(n_measurements > k) * 100
      }))

```

How many observations per race category?
```{r}
HIV_raw %>% 
  group_by(RACE) %>% 
  summarize(n = n())
```

Problems arise when race is:
* Interpreted biologically
* Used as a proxy for genetic ancestry
* Over-interpreted or over-stratified without power

We are not interpreting or estimating race effects, so we should be good to collapse

Potential explanation in final report:
Self-reported race/ethnicity was included as a covariate to account for potential confounding by social determinants of health and healthcare access disparities. Race/ethnicity categories were collected as [describe original categories]. For analyses with limited statistical power due to sample size, we combined categories as follows: [describe any collapsing]. We acknowledge that race is a social construct and that observed differences likely reflect structural inequities rather than biological differences.


Look at distribution of CD4 T cell count, log counts, and boxplots between drug user group
```{r}
hist(ana2$physical0)
summary(ana2$physical0)
hist(ana2$mental0)
summary(ana2$mental0)

# check distribution of QOL measures to identify odd values
hist(ana2$cd40)
hist(ana2$log_cd40)
boxplot(ana2$log_cd40~ana2$hard0)
boxplot(ana2$cd40~ana2$hard0)

```

```{r}
# number missing for hard drug group
ana2 %>% 
  group_by(hard0) %>% 
  summarize(n = n())
```

### Visualizations

```{r}
# convert to long format
long_outcomes <- HIV_raw %>%
  filter(years %in% c(0,2) & !is.na(hard_drugs)) %>%
  mutate(log_vload = log10(VLOAD)) %>% 
           
  pivot_longer(cols = c(log_vload, LEU3N, AGG_PHYS, AGG_MENT),
               names_to = "outcome",
               values_to = "value") %>% 
    
    mutate(
      outcome = recode(outcome,
                       log_vload = "Viral Load (log10 copies/mL)",
                       LEU3N = "CD4+ T-Cell Count",
                       AGG_PHYS = "Physical QOL (0 to 100)",
                       AGG_MENT = "Mental QOL (0 to 100)"),
      years_f = factor(years, levels = c(0, 1, 2), labels = c("0", "1", "2")))
```

Trajectories
```{r}
ggplot(long_outcomes, aes(years, value, group = newid, color = hard_drugs)) +
  geom_line(alpha = 0.05) +
  stat_summary(aes(group = hard_drugs), fun = mean, 
               geom = "line", linewidth = 0.6) +
  facet_wrap(~ outcome, scales = "free_y") +
  scale_color_manual(values = c("steelblue", "firebrick")) +
  labs(title = "Observed Trajectories of Treatment Response",
       x = "Years",
       y="",
       color = "Hard Drugs") +
  theme_bw() +
  theme(legend.position = "bottom") 

```

Boxplot 
```{r label = "Boxplot used in final report (Figure 1)"}
ggplot(long_outcomes, aes(x = years_f, y = value, fill = hard_drugs)) +
  geom_boxplot(outlier.shape = NA, width = 0.7, position = position_dodge(width = 0.75)) +
  geom_point(
    aes(color = hard_drugs),
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
    alpha = 0.08, size = 0.6) +
  facet_wrap(~ outcome, scales = "free_y") +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_color_manual(values = c("steelblue", "firebrick")) +
  labs(
    title = "Treatment Response Distributions by Baseline & Year 2 Hard Drug Use",
    x = "Year since HAART initiation",
    y = NULL,
    fill = "Hard Drugs",
    color = "Hard Drugs") +
  theme_bw() +
  theme(legend.position = "bottom")
```


### Table1 Demographics

```{r label = "Used in final report as Table 1"}
tbl1 <- ana2_cc %>%
  
  tbl_summary(
    by = hard0,
    include = c(mental0, physical0, log_vload0, cd40, adh_bin, age0, bmi0, smoke0, 
                educ0, race0), # add ADH or no?
    type = all_continuous() ~ "continuous2",
    statistic = list(
      all_continuous() ~ c(
        "{mean} ({sd})",
        "{median} ({p25}, {p75})",
        "{min} - {max}"),
      all_categorical() ~ "{n} ({p}%)"),
    
    digits = all_continuous() ~ 1,
    missing = "ifany",
    missing_text = "NA",
    missing_stat = "{N_miss} ({p_miss}%)",
    
    # change variable names
    label = list(
      mental0 ~ "Mental QOL",
      physical0 ~ "Physical QOL",
      log_vload0  ~  "Viral Load (log<sub>10</sub> copies/mL)",
      cd40  ~ "CD4 T-cell Count",
      adh_bin ~ "Adherence",
      age0 ~ "Age",
      bmi0 ~ "BMI",
      smoke0 ~ "Smoke",
      educ0 ~ "Education",
      race0 ~ "Race")) %>% 
  
  add_overall() %>%
  bold_labels() %>% 
  italicize_levels() %>% 
  #add_p(pvalue_fun = label_style_pvalue(digits = 2)) %>% 
  add_n() %>% 
  modify_spanning_header(
    c("stat_1", "stat_2") ~ "**Hard Druge Use**") %>%
  modify_footnote_header(
    footnote = "Table includes only subjects that have both baseline and year 2 measurements",
    columns = all_stat_cols(),
    replace = FALSE) %>% 
  
  as_gt() %>%
  gt::fmt_markdown(columns = everything()) %>% 
  gt::fmt_number(
    columns = dplyr::starts_with("p.value"),
    decimals = 2) %>% 
  tab_options(
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))


tbl1

```

log10 interpretations 
* 10^n = where n is number of places after 1
* 10^8 ≈ in hundreds of millions
* 10^5 ≈ hundreds of thousands
* 10^4 ≈ tens of thousands
* 10^0.5 ≈ 3.16 - in single digits - remember exponent of 0 is 1

Create a table 2 to look at univariate relationships between hard drug use and comorbidities

```{r}
tbl2 <- ana2 %>%
  
  tbl_summary(
    by = hard0,
    include = c(hbp0, diab0, liv34_0, kid0),
    type = all_continuous() ~ "continuous2",
    statistic = list(
      all_continuous() ~ c(
        "{mean} ({sd})",
        "{median} ({p25}, {p75})",
        "{min} - {max}"),
      all_categorical() ~ "{p}%"),
    
    digits = all_continuous() ~ 1,
    missing = "ifany",
    missing_text = "NA",
    missing_stat = "{p_miss}%",
    
    # change variable names
    label = list(
      hbp0 ~ "High Bloood Pressure",
      diab0 ~ "Diabetes",
      liv34_0  ~  "Liver Disease (Stage 3/4)",
      kid0  ~ "Kidney Disease")) %>%
  
  add_overall() %>%
  bold_labels() %>% 
  italicize_levels() %>% 
  add_p(pvalue_fun = label_style_pvalue(digits = 2)) %>% 
  modify_spanning_header(
    c("stat_1", "stat_2") ~ "**Hard Druge Use**") %>%
  modify_footnote_header(
    footnote = "Table includes all subjects from cohort at baseline (years=0)",
    columns = all_stat_cols(),
    replace = FALSE) %>% 
  
  as_gt() %>%
  gt::fmt_markdown(columns = everything()) %>% 
  gt::fmt_number(
    columns = dplyr::starts_with("p.value"),
    decimals = 2) %>% 
  tab_options(
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

tbl2
```

* Rates of liver and kidney disease are higher in drug users (3.2% vs 9.7% and 16% vs 34%, respectively). No difference in diabetes and higher blood pressure is more prevalent in
the non-drug use group (27% vs 15%). Also, there is pretty substantial missing values
for the kidney (71%), diabetes (57%), and liver (34%) groups. 


## Primary Analysis

* How treatment response differs between hard drug users and non-users

* Collider bias? - do these comorbidities occur after exposure and outcome?
 - The exposure and outcome can both independently create a collider variable. In contrast to the case of confounding, where the confounder is associated with both the exposure and outcome, the collider is created by both the exposure and outcome.
 - Comorbidities can be downstream effects of higher viral load/cd4 and hard drug use,
 independently. Therefore, these comorbidities could very well be colliders
  - ***Per Dr. Moore's comments on Data Analysis plan, we do not need to worry about these***  
 

Streamline method utilizing functions to fit all models at once
```{r}
outcomes <- list(
  cd4   = list(y = "cd42", baseline = "cd40"),
  vload = list(y = "log_vload2", baseline = "log_vload0"),
  phys  = list(y = "physical2", baseline = "physical0"),
  ment  = list(y = "mental2", baseline = "mental0"))
```

```{r label="Function to perform all primary models at once"}
fit_models <- function(outcome, baseline=NULL, data) {
  # model form - consistent across all models
  full <- as.formula(
    paste(outcome, "~ hard0 +", baseline,
          "+ age0 + bmi0 + smoke0 + educ0 + race0"))
  # with hard drug coeffecient removed
  reduced <-  as.formula(
    paste(outcome, "~", baseline,
          "+ age0 + bmi0 + smoke0 + educ0 + race0"))
  
  # Frequentist model
  lm_fit <- lm(full, data = data)
  
  # Bayes model
  options(warn = -1)
  brm_fit <- brm(
    full,
    data = data,
    family = gaussian(),
    prior = c(
      # ***the priors need to be more uniform or more robust to different models**
      prior(normal(0, 10000), class = "Intercept"), # less interpretable so make var wider
      prior(normal(0, 10000), class = "b"),
      prior(student_t(3, 0, 100^2), class = "sigma")), # half-t - standard
    chains = 4, cores = 6, iter = 2000,
    refresh = 0)
  
  # WAIC
  log_lik_fit <- log_lik(brm_fit)
  
  # Bayes reduced model for testing hard drugs significance
    brm_red <- brm(
    reduced,
    data = data,
    family = gaussian(),
    prior = c(
      prior(normal(0, 10000), class = "Intercept"), 
      prior(normal(0, 10000), class = "b"),
      prior(student_t(3, 0, 1000), class = "sigma")), 
    chains = 4, cores = 6, iter = 2000,
    refresh = 0)
    options(warn = 0)
    
    # WAIC
    log_lik_red <- log_lik(brm_red)

    # leave-one-out cross-validation for reduced model
    loo_res <- suppressWarnings(brms::loo_compare(
      brms::add_criterion(brm_fit, "loo"),
      brms::add_criterion(brm_red, "loo"),
      model_names = c("Full Model", "Reduced Model")))
    
    waic_fit <- waic(log_lik_fit)
    waic_red <- waic(log_lik_red)
    waic_diff <- loo_compare(waic_fit, waic_red)
  
  # store model results into a list
  list(lm = lm_fit, brm = brm_fit, brm_red = brm_red,
      loo_res=loo_res, ll_fit=log_lik_fit, ll_red=log_lik_red,
       waic_fit=waic_fit, waic_red=waic_red, waic_diff=waic_diff)
}
```

Run all models at once
```{r, eval=F}
t0 <- Sys.time()

model.fits <- imap(outcomes, ~ fit_models(.x$y, .x$baseline, ana2_cc))

t1 <- Sys.time()
message("Runtime: ",
        round(as.numeric(difftime(t1, t0, units = "mins")), 2),
        " minutes")
```

* Macbook Pro M1 (6-cores) Runtime: ≈5.78min


Save model using qs (serialization) for quicker knitting and loading between sessions
```{r, eval=F}
# save model results
qsave(model.fits, "/Users/ryan_summers/GitHub/BIOS6624-Class/Project1/DataProcessed/model.fits.qs",
      nthreads = 6)
```


```{r}
# load model results
model.fits <- qread("/Users/ryan_summers/GitHub/BIOS6624-Class/Project1/DataProcessed/model.fits.qs",
      nthreads = 6)
```


```{r}
# extract info from function list using CD4 outcome as example
model.fits$cd4$lm
model.fits$cd4$brm
model.fits$cd4$loo_res
report(model.fits$cd4$loo_res)
model.fits$cd4$waic_fit
model.fits$cd4$waic_red
model.fits$cd4$waic_diff

# loo results
elpd_diff <- model.fits$cd4$loo_res["Reduced Model", "elpd_diff"]
se_diff <- model.fits$cd4$loo_res["Reduced Model", "se_diff"]

# confirm report() function p-value (wald-normal approx)
z_score <- abs(elpd_diff) / se_diff
(p_value <- 2 * (1 - pnorm(z_score)))

```

"The rule of thumb is that the models are "very similar" if |elpd_diff| (the absolute value of elpd_diff) is less than 4 (Sivula, Magnusson and Vehtari, 2020). If superior to 4, then one can use the SE to obtain a standardized difference (Z-diff) and interpret it as such, assuming that the difference is normally distributed. The corresponding p-value is then calculated as `2 * pnorm(-abs(Z-diff))`. However, note that if the raw ELPD difference is small (less than 4), it doesn't make much sense to rely on its standardized value: it is not very useful to conclude that a model is much better than another if both models make very similar predictions."

Extract key effects for table output function - must specify variable directly for now
```{r label="Function for extracting key elements from models"}
extract_effects <- function(lm_fit, brm_fit, outcome_name, waic_fit, waic_red) {

  # Frequentist
  freq <- broom::tidy(lm_fit, conf.int = TRUE) %>%
    filter(term == "hard0Yes") %>%
    transmute(
      outcome = outcome_name,
      freq_est = estimate,
      freq_lo = conf.low,
      freq_hi = conf.high,
      freq_p = p.value)

  # Bayesian
  draws <- posterior::as_draws_df(brm_fit)
  beta <- draws$b_hard0Yes
  # Get HDI values
  hdi <- hdi(brm_fit, parameters = "b_hard0Yes") 

  bay <- tibble(
    outcome = outcome_name,
    # median
    bay_est_med = median(beta),
    # mean
    bay_est_mean = mean(beta),
    # credible intervals
    bay_CrI_lo = quantile(beta, 0.025),
    bay_CrI_hi = quantile(beta, 0.975),
    # HDIs
    bay_HDI_lo = hdi$CI_low,
    bay_HDI_hi = hdi$CI_high)
    
  # Extract WAIC values 
  elpd_full <- waic_fit$estimates["elpd_waic", "Estimate"]
  elpd_full_se <- waic_fit$estimates["elpd_waic", "SE"]
  waic_full <- waic_fit$estimates["waic", "Estimate"]
  waic_full_se <- waic_fit$estimates["waic", "SE"]

  elpd_red  <- waic_red$estimates["elpd_waic", "Estimate"]
  elpd_red_se <- waic_red$estimates["elpd_waic", "SE"]
  waic_reduced<- waic_red$estimates["waic", "Estimate"]
  waic_reduced_se <- waic_red$estimates["waic", "SE"]

  waic_comp <- tibble::tibble(
    outcome = outcome_name,
    elpd_full = elpd_full,
    elpd_full_se = elpd_full_se,
    waic_full = waic_full,
    waic_full_se = waic_full_se, 
    
    elpd_red = elpd_red,
    elpd_red_se = elpd_red_se,
    waic_reduced = waic_reduced,
    waic_reduced_se = waic_reduced_se,
    
    elpd_diff = elpd_full - elpd_red,
    waic_diff = waic_full - waic_reduced)
  
  # Create a list column with beta draws
  beta_list <- list(beta)


  # combine all tables
  result <- freq %>%
    dplyr::left_join(bay, by = "outcome") %>%
    dplyr::left_join(waic_comp, by = "outcome") %>% 
    dplyr::mutate(beta_draws = beta_list)   # Add as list column

}
```

Comparison table
```{r}
results <- imap_dfr(model.fits, ~ extract_effects(
  lm_fit = .x$lm, 
  brm_fit = .x$brm, 
  outcome_name = .y, 
  waic_fit = .x$waic_fit,
  waic_red = .x$waic_red))
```


```{r}
# extract draws if needed
cd4_beta <- results$beta_draws[results$outcome == "cd4"][[1]]

prob_either <- round(mean(abs(cd4_beta) > 50),3)

# Display results
cat("P(|CD4| > 50):", prob_either, "\n")

vl_beta <- results$beta_draws[results$outcome == "vload"][[1]]
prob_either <- round(mean(abs(vl_beta) > 0.5),3)

# Display results
cat("P(|VL| > 0.5):", prob_either, "\n")

phys_beta <- results$beta_draws[results$outcome == "phys"][[1]]

prob_either <- round(mean(abs(phys_beta) > 2),3)

# Display results
cat("P(|phys| > 2):", prob_either, "\n")

ment_beta <- results$beta_draws[results$outcome == "ment"][[1]]

prob_either <- round(mean(abs(ment_beta) > 2),3)

# Display results
cat("P(|ment| > 50):", prob_either, "\n")

```


```{r}
# add sample sizes from each model and back-transform log10
results2 <- results %>%
  mutate(
    n_lm  = map_int(names(model.fits), 
                    ~ nobs(model.fits[[.x]]$lm)),
    n_brm = map_int(names(model.fits), 
                    ~ nrow(model.frame(model.fits[[.x]]$brm))),
    # # back-transform log10 viral load and make it more interpretable for PI
    # across(
    #   c(freq_est, freq_lo, freq_hi, bay_est_med, bay_HDI_lo, bay_HDI_hi),
    #   ~ if_else(outcome == "vload", -(1-10^.x)*100, .x)),
    outcome = recode(outcome,
                     #<b>CD4</b><br><span style='font-weight:normal'>Mean Cell Diff</span>
                     cd4   = "CD4<br><span style='font-style:italic; 
                     font-weight:normal'>Mean Diff (cells/mm³)</span>",
                     
                     vload = "log<sub>10</sub> VL<br><span style='font-style:italic;
                     font-weight:normal'>Mean Diff (copies/mL)</span>",
                     
                     phys  = "Physical QOL<br><span style='font-style:italic;
                     font-weight:normal'>Mean Diff (0-100 scale)</span>",
                     
                     ment  = "Mental QOL<br><span style='font-style:italic;
                     font-weight:normal'>Mean Diff (0-100 scale)</span>"))
```

* HDI (Highest Density Interval): Highlights the most credible values (highest probability density). If your posterior is skewed, an 89% or 95% HDI represents the shortest range containing that probability mass, directly reflecting the most likely parameter values.
* Equal-Tailed Credible Interval (ETI/Percentile Method/CrI): Based on quantiles (e.g., 2.5% to 97.5%). It is easier to compute and interpret for symmetric distributions but may include less likely values in the tails if the distribution is highly skewed.

Therefore, there should be little difference between these two due to the simplicity of this model
* check model diagnostics to be sure

### Model Results Table

```{r label="Table used in final report (Table 2)"}
tbl_compare <- results2 %>%
  # combine LL and UL CIs
  mutate(
    freq_ci = sprintf("(%.1f, %.1f)", freq_lo, freq_hi),
    bay_ci  = sprintf("(%.1f, %.1f)", bay_HDI_lo, bay_HDI_hi)) %>%
  select(outcome, n_brm, n_lm,
         bay_est_med, bay_ci, waic_diff,
         freq_est, freq_ci, freq_p) %>%
  
  gt(rowname_col = "outcome") %>%
  cols_label(
    n_lm = "N",
    n_brm = "N",
    freq_est = "Users vs Non",
    freq_ci = "95% CI",
    freq_p   = "P-value",
    bay_est_med  = "Users vs Non", # median
    bay_ci = "95% HDI",
    waic_diff = md("WAIC Diff<br>(Full-Red)")) %>%
  # header for freq
  tab_spanner(
    label = "Frequentist",
    columns = c(n_lm, freq_est, freq_ci, freq_p)) %>%
  # header for Bayes
  tab_spanner(
    label = "Bayesian",
    columns = c(n_brm, bay_est_med, bay_ci, waic_diff)) %>%
  # format numbers 
  fmt_number(columns = c(freq_est, freq_ci, bay_est_med, bay_ci), decimals = 1) %>%
  fmt_number(columns = c(waic_diff), decimals = 1,
             use_seps = F) %>%
  fmt_number(columns = freq_p, decimals = 3) %>%
  
  # change p-value reporting style
  text_transform(
    locations = cells_body(columns = freq_p),
    fn = function(x) ifelse(as.numeric(x) < 0.001, "<0.001",
                            format(round(as.numeric(x), 3), nsmall = 3))) %>% 
  # bold specific text
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_stub(rows = everything())) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_spanners(
      spanners = c("Frequentist", "Bayesian"))) %>% 
  # center column labels
  tab_style(
    style = cell_text(align = "center"),
    locations = cells_column_labels(everything())) %>% 
  # add a footnotes
  tab_footnote(
    footnote = "For hard drug users versus non-users, estimates represent 
    adjusted differences (Users − Non-users). Negative values indicate lower 
    outcome values among users.",
    locations = cells_column_labels(columns = c(bay_est_med, freq_est))) %>%
  tab_footnote(
    footnote = "Bayesian intervals are 95% high density intervals; 
    Frequentist intervals are 95% confidence intervals.",
    locations = cells_column_labels(columns = c(freq_ci, bay_ci))) %>% 
  tab_footnote(
    footnote = "For WAIC, lower values indicate better model fit. 
    WAIC difference is the full model minus model w/out exposure (hard drug use).",
    locations = cells_column_labels(columns = c(waic_diff))) %>% 
  fmt_markdown(columns = outcome) %>% 
  # makes for a more condensed table
  tab_options(
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(2), #3 appears to be best
    # reduce table width
    table.width = pct(85),
    # thickness of the horizontal borders
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

tbl_compare
```

```{r}
summary(model.fits$cd4$brm)
hdi(model.fits$cd4$brm)
```


Viral load: After adjusting for baseline viral load and covariates, hard-drug users had approximately 14% lower viral as compared to non-users at Year 2.

```{r}
nobs(model.fits$cd4$lm)
nrow(model.frame(model.fits$cd4$brm))  # or nrow(model.frame(cd4_brm))
report(model.fits$cd4$brm)
sexit(model.fits$cd4$brm)

report(model.fits$phys$brm)
sexit(model.fits$phys$brm)
```

After adjusting for baseline baseline CD4+ T cell counts, age, BMI, smoking, education, and race, baseline hard-drug users had an estimated lower CD4 count of approximately 172.2 at Year 2 compared with non-users [95% CI: (-235 to -109.4); p<0.001].




Look at distribution of CD4 T cell count, log counts, and boxplots between drug user group
```{r}
hist(ana2$cd40)
hist(ana2$log_cd40)
boxplot(ana2$log_cd40~ana2$hard0)
boxplot(ana2$cd40~ana2$hard0)
```

Check that residuals look good using raw CD4 counts
```{r}
# check residuals to determine if skewness is too extreme for this model approach
plot(model.fits$ment$lm)   # residuals vs fitted
plot(model.fits$phys$lm)
plot(model.fits$cd4$lm)

```

* some outliers but overall residuals look good with raw CD4 values


## Mediation Analysis

Custom function for calculating mediation coefficients
```{r}
# boot_fun <- function(data, idx) {
#   d <- data[idx, ]
#   m <- glm(adh_bin ~ hard0 + age0 + bmi0 + smoke0 + educ0 + race0, 
#            family = binomial,
#            data = d)
#   y <- lm(cd42 ~ hard0 + adh_bin + cd40 +
#             age0 + bmi0 + smoke0 + educ0 + race0, data = d)
# 
#   a <- coef(m)["hard0"]     
#   b <- coef(y)["adh2"]
#   cprime <- coef(y)["hard0"]
# 
#   c(indirect = a*b, direct = cprime, total = cprime + a*b)
# }
# 
# set.seed(1)
# bt <- boot(ana2_cc, statistic = boot_fun, R = 2000)
# 
# boot.ci(bt, type = "perc", index = 1)  # indirect
# boot.ci(bt, type = "perc", index = 2)  # direct
# boot.ci(bt, type = "perc", index = 3)  # total
```


Pre-lim mediation analysis w/out Covs
```{r}
# Fit our three models
mod2_crude <- lm(cd42 ~ hard0, data=ana2_cc) # Y~X
mod2_adjusted <- lm(cd42 ~ hard0 + adh_bin, data=ana2_cc) # Y~ X+M
mod2_covariate <- glm(adh_bin ~ hard0, 
                     family = binomial,
                     data=ana2_cc) # M~X

summary(mod2_covariate)

# Extract coefficients
coef(mod2_crude)
coef(mod2_adjusted)
coef(mod2_covariate)
```


DAG
```{tikz, tikz-med, fig.ext = 'png', cache=TRUE, echo=F, fig.align="center", fig.width=4}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=4cm, auto,>=latex, thick, scale = 0.5]
\node (X) [align=center] {Hard Drug Use \\ ($X$)};
\node (M) [align=center,above of=X,xshift=2cm,yshift=-2cm] {Adherence \\ ($M$)};
\node (Y) [align=center,right of=X] {Year 2 CD4 \\ ($Y$)};
\draw[->] (X) to node [above left] {$\hat{\gamma}_{X}=1.59$} (M);
\draw[->] (M) to node [above right] {$\hat{\beta}_{M}=98.03$} (Y);
\draw[->] (X) to node [align=center,swap] {$\hat{\beta}_{adj}=-202.55$ \\ $\hat{\beta}_{crude}=-193.91$} (Y);
\end{tikzpicture}
```

Preliminary models using mediation package
```{r}
# Mediator model: adherence as outcome
m_mod <- glm(adh_bin ~ hard0 + age0 + bmi0 + smoke0 + educ0 + race0, 
             family = binomial,
            data = ana2_cc)

# Outcome model: Year-2 viral load, adjusting for baseline viral load + adherence
y_mod <- lm(cd42 ~ hard0 + adh_bin + cd40 +
              age0 + bmi0 + smoke0 + educ0 + race0, 
            data = ana2_cc)

# Mediation estimation (bootstrap for inference)
med_vl <- mediate(m_mod, y_mod, treat = "hard0", mediator = "adh_bin",
                  boot = F, sims = 2000)

summary(med_vl)

```


```{r}
adh_outcomes <- list(
  cd4   = list(y = "cd42", baseline = "cd40", adh="adh_bin"),
  vload = list(y = "log_vload2", baseline = "log_vload0",adh="adh_bin"),
  phys  = list(y = "physical2", baseline = "physical0",adh="adh_bin"),
  ment  = list(y = "mental2", baseline = "mental0",adh="adh_bin"))
```

Similar to previous model function 
* could have modified original function to handle both but given time constraints
  this will have to do
```{r label="Function to perform all mediation models at once"}
fit_med_models <- function(outcome, baseline=NULL, adh, data) {
  # model form - consistent across all models
  full <- as.formula(
    paste(outcome, "~ hard0 +", baseline, "+", adh,
          "+ age0 + bmi0 + smoke0 + educ0 + race0"))
  # with adh coeffecient removed
  reduced <-  as.formula(
    paste(outcome, "~ hard0 +",baseline,
          "+ age0 + bmi0 + smoke0 + educ0 + race0"))
  
  # Frequentist model
  lm_fit <- lm(full, data = data)
  lm_red <- lm(reduced, data = data)
  
  # Bayes model
  options(warn = -1)
  brm_fit <- brm(
    full,
    data = data,
    family = gaussian(),
    prior = c(
      # ***the priors need to be more uniform or more robust to different models**
      prior(normal(0, 10000), class = "Intercept"), # less interpretable so make var wider
      prior(normal(0, 10000), class = "b"),
      prior(student_t(3, 0, 100^2), class = "sigma")), # half-t - standard
    chains = 4, cores = 6, iter = 2000,
    refresh = 0)
  
  # WAIC
  log_lik_fit <- log_lik(brm_fit)
  
  # Bayes reduced model for testing hard drugs significance
  brm_red <- brm(
    reduced,
    data = data,
    family = gaussian(),
    prior = c(
      prior(normal(0, 10000), class = "Intercept"), 
      prior(normal(0, 10000), class = "b"),
      prior(student_t(3, 0, 1000), class = "sigma")), 
    chains = 4, cores = 6, iter = 2000,
    refresh = 0)
    options(warn = 0)
    
    # WAIC
    log_lik_red <- log_lik(brm_red)

    # leave-one-out cross-validation for reduced model
    loo_res <- suppressWarnings(brms::loo_compare(
      brms::add_criterion(brm_fit, "loo"),
      brms::add_criterion(brm_red, "loo"),
      model_names = c("Full Model", "Reduced Model")))
    
    waic_fit <- waic(log_lik_fit)
    waic_red <- waic(log_lik_red)
    waic_diff <- loo_compare(waic_fit, waic_red)
  
  # store model results into a list
  list(lm_fit = lm_fit, lm_red=lm_red, brm = brm_fit, brm_red = brm_red,
      loo_res=loo_res, ll_fit=log_lik_fit, ll_red=log_lik_red,
       waic_fit=waic_fit, waic_red=waic_red, waic_diff=waic_diff)
}
```

Run all models at once
```{r, eval=F}
t0 <- Sys.time()

model.med.fits <- imap(adh_outcomes, ~ fit_med_models(.x$y, .x$baseline, .x$adh,
                                              ana2_cc))

t1 <- Sys.time()
message("Runtime: ",
        round(as.numeric(difftime(t1, t0, units = "mins")), 2),
        " minutes")
```

Runtime: 3.76min

```{r, eval=F}
# save model results
qsave(model.med.fits, "/Users/ryan_summers/GitHub/BIOS6624-Class/Project1/DataProcessed/model.med.fits.qs",
      nthreads = 6)
```


```{r}
# load model results
model.med.fits <- qread("/Users/ryan_summers/GitHub/BIOS6624-Class/Project1/DataProcessed/model.med.fits.qs",
      nthreads = 6)
```

```{r}
extract_med_effects <- function(lm_fit, lm_red=NULL, 
                            brm_fit, brm_red=NULL, 
                            outcome_name, 
                            waic_fit=NULL, waic_red=NULL) 
  {
  # Frequentist
  freq_full <- broom::tidy(lm_fit, conf.int = TRUE) %>%
    filter(term == "hard0Yes") %>% # needs to be more robust - revisit later
    transmute(
      outcome = outcome_name,
      freq_full_est = estimate,
      freq_full_lo = conf.low,
      freq_full_hi = conf.high,
      freq_full_p = p.value)
  
  freq_red <- broom::tidy(lm_red, conf.int = TRUE) %>%
    filter(term == "hard0Yes") %>% # needs to be more robust - revisit later
    transmute(
      outcome = outcome_name,
      freq_red_est = estimate,
      freq_red_lo = conf.low,
      freq_red_hi = conf.high,
      freq_red_p = p.value)

  # Bayesian
  draws_full <- posterior::as_draws_df(brm_fit)
  beta_full <- draws_full$b_hard0Yes
  # Get HDI values
  hdi_full <- hdi(brm_fit, parameters = "b_hard0Yes") # needs to be more robust - revisit later

  # full Bayes model
  bay_full <- tibble(
    outcome = outcome_name,
    # median
    bay_full_est_med = median(beta_full),
    # mean
    bay_full_est_mean = mean(beta_full),
    # credible intervals
    bay_full_CrI_lo = quantile(beta_full, 0.025),
    bay_full_CrI_hi = quantile(beta_full, 0.975),
    # HDIs
    bay_full_HDI_lo = hdi_full$CI_low,
    bay_full_HDI_hi = hdi_full$CI_high)

  # reduced Bayes model
  draws_red <- posterior::as_draws_df(brm_red)
  beta_red <- draws_red$b_hard0Yes
  # Get HDI values
  hdi_red <- hdi(brm_red, parameters = "b_hard0Yes") 
  
  bay_red <- tibble(
    outcome = outcome_name,
    # median
    bay_red_est_med = median(beta_red),
    # mean
    bay_red_est_mean = mean(beta_red),
    # credible intervals
    bay_red_CrI_lo = quantile(beta_red, 0.025),
    bay_red_CrI_hi = quantile(beta_red, 0.975),
    # HDIs
    bay_red_HDI_lo = hdi_red$CI_low,
    bay_red_HDI_hi = hdi_red$CI_high)
    
  # Extract WAIC values 
  elpd_full <- waic_fit$estimates["elpd_waic", "Estimate"]
  elpd_full_se <- waic_fit$estimates["elpd_waic", "SE"]
  waic_full <- waic_fit$estimates["waic", "Estimate"]
  waic_full_se <- waic_fit$estimates["waic", "SE"]

  elpd_red  <- waic_red$estimates["elpd_waic", "Estimate"]
  elpd_red_se <- waic_red$estimates["elpd_waic", "SE"]
  waic_reduced<- waic_red$estimates["waic", "Estimate"]
  waic_reduced_se <- waic_red$estimates["waic", "SE"]

  waic_comp <- tibble::tibble(
    outcome = outcome_name,
    elpd_full = elpd_full,
    elpd_full_se = elpd_full_se,
    waic_full = waic_full,
    waic_full_se = waic_full_se, 
    
    elpd_red = elpd_red,
    elpd_red_se = elpd_red_se,
    waic_reduced = waic_reduced,
    waic_reduced_se = waic_reduced_se,
    
    elpd_diff = elpd_full - elpd_red,
    waic_diff = waic_full - waic_reduced)

  # Create a list column with beta draws
  beta_list_full <- list(beta_full)
  beta_list_red <- list(beta_red)
  
  # combine all tables
  result <- freq_full %>%
    dplyr::left_join(freq_red, by = "outcome") %>%
    dplyr::left_join(bay_full, by = "outcome") %>%
    dplyr::left_join(bay_red, by = "outcome") %>%
    dplyr::left_join(waic_comp, by = "outcome") %>%
    dplyr::mutate(beta_draws_full = beta_list_full) %>%   # Add as list column
    dplyr::mutate(beta_draws_red = beta_list_red)
  
  return(result)
}
```


```{r}
med_results <- imap_dfr(
  model.med.fits, ~ extract_med_effects(
    lm_fit = .x$lm_fit,
    lm_red = .x$lm_red,
    brm_fit = .x$brm,
    brm_red = .x$brm_red,
    outcome_name = .y))
```

```{r}
# need to convert log10 VL back to OG units because of instability in ratio 
# calculations when near 0
med_results_bt <- med_results %>%
  mutate(
    # numeric columns
    across(
      c(freq_full_est, freq_full_lo, freq_full_hi,
        bay_full_est_med, bay_full_est_mean, bay_full_HDI_lo, bay_full_HDI_hi,
        freq_red_est,  freq_red_lo,  freq_red_hi,
        bay_red_est_med, bay_red_est_mean, bay_red_HDI_lo, bay_red_HDI_hi),
      ~ if_else(outcome == "vload", 10^.x, .x)),

    # list-columns of draws
    beta_draws_full = if_else(
      outcome == "vload",
      map(beta_draws_full, ~ 10^.x),
      beta_draws_full),
    beta_draws_red = if_else(
      outcome == "vload",
      map(beta_draws_red, ~ 10^.x),
      beta_draws_red))
```


Calculate proportion mediated for each outcome 
```{r}
summ_draws <- function(x, probs = c(.025, .5, .975)) {
  qs <- quantile(x, probs = probs, na.rm = TRUE)
  tibble(
    mean = mean(x, na.rm = TRUE),
    median = unname(qs[2]),
    lo = unname(qs[1]),
    hi = unname(qs[3]),
    p_gt0 = mean(x > 0, na.rm = TRUE))
}
```

```{r}
bay_med_calc <- med_results_bt %>%
  mutate(
    # Frequentist TE and DE
    freq_TE = freq_red_est,
    freq_DE = freq_full_est,
    freq_ind = freq_TE - freq_DE,
    freq_prop = if_else(abs(freq_TE) > 0.01, freq_ind / freq_TE, NA_real_),

    # Bayesian TE and DE draws
    bayes_prop = if_else(abs(bay_red_est_med) > 0.01, 
                         (bay_red_est_med-bay_full_est_med) / bay_red_est_med, NA_real_),
    beta_TE = beta_draws_red, # Total Effect : w/out ADH in model
    beta_DE = beta_draws_full, # Direct Effect : w/ ADH in model
    ind_draws = map2(beta_TE, beta_DE, ~ .x - .y), # indirect draws TE-DE
    prop_draws = map2(beta_TE, ind_draws, ~ .y / .x),
    prop_draws = map2(beta_TE, prop_draws, 
                      ~ ifelse(abs(.x) > 0.01, .y, NA_real_))) %>%
  transmute(
    outcome,
    # Frequentist (point estimates)
    freq_TE, freq_DE, freq_ind, freq_prop, bayes_prop,
    # indirect (TE-DE) - should be positive for proper mediation - exposure effect should decrease
    ind_summ = map(ind_draws, summ_draws),
    # proportion mediated
    prop_summ = map(prop_draws, summ_draws),
    prop_valid = map_int(prop_draws, ~ sum(is.finite(.x), na.rm = TRUE)),
    prop_total = map_int(beta_draws_red, length)) %>%
  unnest_wider(ind_summ, names_sep = "_bay_ind") %>%
  unnest_wider(prop_summ, names_sep = "_bay_prop")
```


```{r}
med_results_total <- med_results_bt %>%
  left_join(bay_med_calc, by = "outcome") 
```



```{r}
# add sample sizes from each model and back-transform log10
med_results_total2 <- med_results_total %>%
  mutate(
    n_lm  = map_int(names(model.med.fits), 
                    ~ nobs(model.med.fits[[.x]]$lm_fit)),
    n_brm = map_int(names(model.med.fits), 
                    ~ nrow(model.frame(model.med.fits[[.x]]$brm))),
    # # back-transform log10 viral load and make it more interpretable for PI
    # across(
    #   c(freq_est, freq_lo, freq_hi, bay_est_med, bay_HDI_lo, bay_HDI_hi),
    #   ~ if_else(outcome == "vload", -(1-10^.x)*100, .x)),
    outcome = recode(outcome,
                     #<b>CD4</b><br><span style='font-weight:normal'>Mean Cell Diff</span>
                     cd4   = "CD4<br><span style='font-style:italic; 
                     font-weight:normal'>Mean Diff (cells/mm³)</span>",
                     
                     vload = "log<sub>10</sub> VL<br><span style='font-style:italic;
                     font-weight:normal'>Mean Diff (copies/mL)</span>",
                     
                     phys  = "Physical QOL<br><span style='font-style:italic;
                     font-weight:normal'>Mean Diff (0-100 scale)</span>",
                     
                     ment  = "Mental QOL<br><span style='font-style:italic;
                     font-weight:normal'>Mean Diff (0-100 scale)</span>"))
```



### Mediation Model Results Table

```{r label="Table used in final report (Table 3)"}
tbl_compare2 <- med_results_total2 %>%
  # combine LL and UL CIs
  mutate(
    freq_full_ci = sprintf("(%.1f, %.1f)", freq_full_lo, freq_full_hi),
    freq_red_ci = sprintf("(%.1f, %.1f)", freq_red_lo, freq_red_hi),
    bay_full_ci  = sprintf("(%.1f, %.1f)", bay_full_HDI_lo, bay_full_HDI_hi),
    bay_red_ci  = sprintf("(%.1f, %.1f)", bay_red_HDI_lo, bay_red_HDI_hi),
    med_ci = sprintf("(%.1f, %.1f)", prop_summ_bay_proplo, prop_summ_bay_prophi),
    ind_summ_bay_indp_lt0 = 1-ind_summ_bay_indp_gt0,
    chng_in_coeff = bay_red_est_med-bay_full_est_med,
    ) %>%
  select(outcome, n_brm, n_lm, bay_red_est_med,
         bay_full_est_med, chng_in_coeff,
         bayes_prop, med_ci,
         freq_red_est, freq_full_est, freq_prop) %>%
  
  gt(rowname_col = "outcome") %>%
  cols_label(
    n_lm = "N",
    n_brm = "N",
    # Freq
    freq_red_est = "W/out ADH",
    freq_full_est = "W/ADH",
    freq_prop = "Prop Med",
    # Bayes
    bay_red_est_med  = "W/out ADH",
    bay_full_est_med  = "W/ADH",
    chng_in_coeff = "Δ",
    # mediation cacls
    bayes_prop = "Prop Med",
    med_ci = "95% HDI") %>% 
  # header for freq
  tab_spanner(
    label = "Frequentist",
    columns = c(n_lm, freq_red_est, freq_full_est, freq_prop)) %>%
  # header for Bayes
  tab_spanner(
    label = "Bayesian",
    columns = c(n_brm, bay_red_est_med, bay_full_est_med, chng_in_coeff,
                bayes_prop, med_ci)) %>%
  # format numbers 
  fmt_number(columns = c(freq_full_est, bay_full_est_med,chng_in_coeff,
                         freq_red_est, bay_red_est_med), 
             decimals = 1) %>%
  fmt_number(columns = c(bayes_prop, freq_prop,
                         med_ci), 
             decimals = 3) %>% 
  
  # bold specific text
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_stub(rows = everything())) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_spanners(
      spanners = c("Frequentist", "Bayesian"))) %>% 
  # center column labels
  tab_style(
    style = cell_text(align = "center"),
    locations = cells_column_labels(everything())) %>% 
  # add a footnotes
  tab_footnote(
    footnote = "For hard drug users versus non-users, estimates represent 
    adjusted differences (Users − Non-users). Negative values indicate lower 
    outcome values among users.",
    locations = cells_column_labels(
      columns = c(bay_full_est_med, freq_full_est))) %>%
  tab_footnote(
    footnote = "Represents the change in the hard-drug coefficient after 
    adjusting for adherence; Not a formal mediation effect.",
    locations = cells_column_labels(
      columns = c(chng_in_coeff))) %>%
  tab_footnote(
    footnote = "Proportion Mediated = (Model with ADH - Model without ADH) / Model without ADH or
    (TE-DE)/TE",
    locations = cells_column_labels(
      columns = bayes_prop)) %>%
  tab_footnote(
    footnote = "95% high density intervals for proportion mediated",
    locations = cells_column_labels(
      columns = med_ci)) %>%

  fmt_markdown(columns = outcome) %>% 
  # makes for a more condensed table
  tab_options(
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(2), #3 appears to be best
    # reduce table width
    table.width = pct(85),
    # thickness of the horizontal borders
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

tbl_compare2
```




### Diagnsotics for Bayesian Mediation Model

```{r, eval=F}
# Mediator model
med_model <- bf(adh_bin ~ hard0 + cd40 + age0 + bmi0 + smoke0 + educ0 + race0,
                family = bernoulli(link = "logit"))

# Outcome model
out_model <- bf(log_vload2 ~ hard0 + adh_bin + log_vload0 +
                age0 + bmi0 + smoke0 + educ0 + race0,
                family = gaussian())

med.brm.fit <- brm(med_model + out_model + set_rescor(FALSE),
                   chains = 4, cores = 6, iter = 2000,
           data = ana2_cc)
```

```{r, eval=F}
table(ana2_cc$adh_bin, useNA="ifany")
```

```{r, eval=F}
summary(med.brm.fit)
```


```{r, eval=F}
## X = drug use & M = adherence

## We are using complete-case analysis so use ana2_cc dataset

# make sure hard0 has the right reference 
ana2_cc$hard0 <- relevel(factor(ana2_cc$hard0), ref = "No")

# mediator levels - current reference is <95%
levM <- levels(ana2_cc$adh_bin)
m0 <- levels(ana2_cc$adh_bin)[1] # reference
m1 <- levels(ana2_cc$adh_bin)[2]      

# create new dataset with X set to No and Yes
hd_NO <- ana2_cc
hd_NO$hard0 <- factor("No", levels = levels(ana2_cc$hard0))

hd_YES <- ana2_cc
hd_YES$hard0 <- factor("Yes", levels = levels(ana2_cc$hard0))

# Posterior expected adherence probabilities P(M=1|X,C)
# For bern, posterior_epred gives probability of success (i.e., P(M == m1))
pM0 <- posterior_epred(med.brm.fit, resp = "adhbin", newdata = hd_NO)  
pM1 <- posterior_epred(med.brm.fit, resp = "adhbin", newdata = hd_YES)  

# Posterior expected CD4 means for each X and each fixed mediator level
# Build newdata that sets adh_bin to m0 or m1
hd_NO_lt95 <- hd_NO; hd_NO_lt95$adh_bin <- factor(m0, levels = levM)
hd_NO_gt95 <- hd_NO; hd_NO_gt95$adh_bin <- factor(m1, levels = levM)

hd_YES_lt95 <- hd_YES; hd_YES_lt95$adh_bin <- factor(m0, levels = levM)
hd_YES_gt95 <- hd_YES; hd_YES_gt95$adh_bin <- factor(m1, levels = levM)

# expected CD4 if No hard drugs w/ <95% ADH + Covs
mu0_m0 <- posterior_epred(med.brm.fit, resp = "logvload2", newdata = hd_NO_lt95)  
# expected CD4 if No hard drugs w/ ≥95% ADH + Covs
mu0_m1 <- posterior_epred(med.brm.fit, resp = "logvload2", newdata = hd_NO_gt95)  
# expected CD4 if Yes hard drugs w/ <95% ADH + Covs
mu1_m0 <- posterior_epred(med.brm.fit, resp = "logvload2", newdata = hd_YES_lt95)  
# expected CD4 if Yes hard drugs w/ ≥95% ADH + Covs
mu1_m1 <- posterior_epred(med.brm.fit, resp = "logvload2", newdata = hd_YES_gt95)  

## Standardize over the mediator distribution:
 # 1 - pM0 = prob of low ADH ; pM0 = prob of high ADH | No hard drug use
 # 1 - pM1 = prob of low ADH ; pM1 = prob of high ADH | Yes hard drug use
EY_0M0 <- rowMeans(mu0_m0 * (1 - pM0) + mu0_m1 * pM0) # pop-level E[CD4|X=0]
EY_1M1 <- rowMeans(mu1_m0 * (1 - pM1) + mu1_m1 * pM1) # pop-level E[CD4|X=1]

# expectation of CD4 if a hard drug user but ADH behaved as if they were NOT
# a drug user
EY_1M0 <- rowMeans(mu1_m0 * (1 - pM0) + mu1_m1 * pM0)

## CD4 depends on adherence.
## ADH depends on drug use.
## So to understand mediation, we must avg CD4 preds over how often people are ADH

# calculate the effects
TE  <- EY_1M1 - EY_0M0
NIE <- EY_1M1 - EY_1M0
NDE <- EY_1M0 - EY_0M0

# summarize
summ <- function(x) {
  c(
    mean = mean(x),
    hdi_low = hdi(x, prob = 0.95)[1],
    hdi_high = hdi(x, prob = 0.95)[2],
    Pr_gt_0 = mean(x > 0))
  }

rbind(TE = summ(TE), NIE = summ(NIE), NDE = summ(NDE))
```

Total Effect (TE)
Interpretation:
Moving from No hard drug use to Yes hard drug use is associated with an average decrease of about 165 CD4 cells at Year 2.

Indirect Effect (NIE-mediation through adherence)
Interpretation:
There is strong posterior evidence that adherence mediates part of the association.

Direct Effect (NDE)
Interpretation:
Even if adherence were held constant, hard drug use would still be associated with a ≈170 cell lower CD4 count.

This suggests that most of the effect is not explained by adherence

Proportion Mediated
mean(NIE / TE) = 6/168 ≈ 3.6%


## Dropout Rate Analysis

```{r}
ana2_dr <- ana2 %>% 
  select(log_vload2, log_vload0, cd42, cd40, hard0, adh2, physical0, physical2,
         mental0, mental2, age0, bmi0, smoke0, educ0, race0) %>% 
  mutate(complete_obs = if_all(everything(), ~ !is.na(.)))

table(ana2_dr$complete_obs, ana2_dr$hard0)
t.test(cd40 ~ complete_obs, data = ana2_dr)

```


Generate a table that shows complete case analysis vs using all 715 observations
* in other words 463 vs 715
* since we are using complete case to be consistent across all analyses - I am comparing complete cases of 463 vs using all 715 subjects who has baseline or year 2. I reason this is the best approach??
```{r}
ana2_drop <- ana2_dr %>%
  mutate(
    hard0 = factor(hard0, levels = c("No", "Yes"), exclude = NULL),
    followup_status = factor(
      if_else(complete_obs == T, "Complete", "Missing"),
      levels = c("Complete", "Missing")))

tbl_dropout <- ana2_drop %>%
  tbl_summary(
    by = followup_status,
    include = c(hard0, log_vload0, cd40, physical0, mental0),
    type = list(
    all_continuous() ~ "continuous2"),
    statistic = list(
      all_continuous() ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
      all_categorical() ~ "{n} ({p}%)"),
    digits = all_continuous() ~ 1,
    missing = "ifany",
    missing_text = "Missing Both Measurements",
    missing_stat = "{N_miss} ({p_miss}%)",
    label = list(
      hard0 ~ "Hard Drug Use",
      log_vload0 ~ "Baseline viral load (log10 copies/mL)",
      cd40 ~ "Baseline CD4 T-cell Count",
      physical0 ~ "Baseline physical QOL",
      mental0 ~ "Baseline mental QOL")) %>%
  add_p(pvalue_fun = ~style_pvalue(.x, digits = 3)) %>%
  add_n() %>%
  add_overall() %>%
  bold_labels() %>%
    modify_spanning_header(
    c("stat_1", "stat_2") ~ "**Cases**") %>%
  modify_footnote_header(
    columns = all_stat_cols(),
    footnote = "Follow-up status defined based on availability of Year 2 outcome.",
    replace = F) %>%
  as_gt() %>%
  tab_options(
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

tbl_dropout
```


```{r label = "Used in final report as table 4"}
ana2_drop <- ana2_dr %>%
  mutate(
    hard0 = factor(hard0, levels = c("No", "Yes"),
                   labels = c("Non-users", "Users")),
    followup_status = factor(
      if_else(complete_obs, "Complete", "Missing"),
      levels = c("Complete", "Missing")),
        smoke0 = case_when(
      smoke0 %in% c("2") ~ "Former",
      smoke0 %in% c("3") ~ "Current",
      TRUE ~ NA_character_) %>% 
      factor(levels = c("Former", "Current")))

tbl_dropout_strata <- ana2_drop %>%
  tbl_strata(
    strata = hard0,
    .tbl_fun = ~ .x %>%
      tbl_summary(
        by = followup_status,
        include = c(log_vload0, cd40, physical0, mental0),
        type = all_continuous() ~ "continuous2",
        statistic = list(
          all_continuous() ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
          all_categorical() ~ "{n} ({p}%)"),
        digits = all_continuous() ~ 1,
        missing = "ifany",
        missing_text = "Missing Both Measurements",
        missing_stat = "{N_miss} ({p_miss}%)",
        label = list(
          log_vload0 ~ "Baseline viral load (log10 copies/mL)",
          cd40 ~ "Baseline CD4 T-cell Count",
          physical0 ~ "Baseline physical QOL",
          mental0 ~ "Baseline mental QOL")) %>%
      add_p(
        test = list(
          all_continuous() ~ "wilcox.test",
          all_categorical() ~ "fisher.test"   # good default if hard drug Yes is small
        ),
        pvalue_fun = ~ style_pvalue(.x, digits = 3)) %>%
      add_n() %>%
      modify_spanning_header(all_stat_cols() ~ "**Cases**") %>%
      modify_caption("**Baseline characteristics by follow-up status, 
                     stratified by baseline hard drug use**")) %>%
  bold_labels()

# Convert to gt (optional; tbl_strata prints fine as gtsummary too)
tbl_dropout_strata_gt <- 
  tbl_dropout_strata %>% 
  as_gt() %>% 
    tab_options(
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

tbl_dropout_strata_gt
```


Confirmation check that gtsummary() is calculating numbers correctly
```{r}
pairs <- tribble(
  ~outcome, ~base, ~yr2,
  "viral load", "log_vload0", "log_vload2",
  "CD4", "cd40", "cd42",
  "physical QOL", "physical0", "physical2",
  "mental QOL", "mental0", "mental2")

miss_tbl <- pairs %>%
  rowwise() %>%
  mutate(
    miss0_have2 = sum(is.na(ana2[[base]]) & !is.na(ana2[[yr2]])),
    have0_miss2 = sum(!is.na(ana2[[base]]) & is.na(ana2[[yr2]])),
    both = sum(!is.na(ana2[[base]]) & !is.na(ana2[[yr2]])),
    none = sum(is.na(ana2[[base]]) & is.na(ana2[[yr2]])),
    perc_miss = 1-both/715) %>%
  ungroup()

miss_tbl
```


### Logistic Dropout Modelling (N=648)
```{r, eval=FALSE}
dropout.model <- glm(
  followup_status ~ age0 + bmi0 + race0 + smoke0 + educ0 +
            log_vload0 + cd40 + physical0 + mental0 + hard0,
  family = binomial,
  data = ana2_drop)
summary(dropout.model)
nobs(dropout.model)
broom::tidy(dropout.model, exponentiate = TRUE, conf.int = TRUE)

qsave(dropout.model, "/Users/ryan_summers/GitHub/BIOS6624-Class/Project1/DataProcessed/dropout.model.qs")
```

```{r}
dropout.model <- qread("/Users/ryan_summers/GitHub/BIOS6624-Class/Project1/DataProcessed/dropout.model.qs")
```

* about 3.8% (1-0.962) lower odds of dropout per unit increase in physical QOL
    * Each one-unit increase in baseline physical QOL was associated with approximately 3.9% higher odds of remaining in the study.
    
* about 2.5% (1-0.975) lower odds of dropout per unit increase in age
    * Each one-year increase in age was associated with approximately 2.6% higher odds of remaining in follow-up.

In logistic regression models, each one-point increase in baseline physical QOL was associated with 3.8% lower odds of dropout (OR = 0.962), corresponding to approximately 3.9% higher odds of remaining in follow-up. Similarly, each additional year of age was associated with 2.5% lower odds of dropout (OR = 0.975), or 2.6% higher odds of retention.


```{r, eval=FALSE}
dropout.brm <- brm(
  followup_status ~ age0 + bmi0 + race0 + smoke0 + educ0 +
            log_vload0 + cd40 + physical0 + mental0 + hard0,
    data = ana2_drop,
    family = bernoulli(link = "logit"),
    prior = c(
      prior(normal(0, 5), class = "Intercept"), 
      prior(normal(0, 2.5), class = "b")), 
    chains = 4, cores = 6, iter = 2000,
    refresh = 0)

summary(dropout.brm)
nobs(dropout.brm)

qsave(dropout.brm, "/Users/ryan_summers/GitHub/BIOS6624-Class/Project1/DataProcessed/dropout.brm.qs")

```


```{r}
dropout.brm <- qread("/Users/ryan_summers/GitHub/BIOS6624-Class/Project1/DataProcessed/dropout.brm.qs")
```




```{r}

# Frequentist (OR + 95% CI)
freq_tbl <- broom::tidy(dropout.model, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  transmute(
    term,
    freq_or = estimate,
    freq_lo = conf.low,
    freq_hi = conf.high,
    freq_p  = p.value)

# Bayesian tidy (OR + 95% CrI)
bay_tbl <- broom.mixed::tidy(
  dropout.brm,
  effects = "fixed",
  conf.int = TRUE,
  conf.level = 0.95) %>%
  filter(term != "Intercept") %>%
  transmute(
    term,
    bay_or = exp(estimate),
    bay_lo = exp(conf.low),
    bay_hi = exp(conf.high))

#Combine
dropout_combined <- full_join(bay_tbl, freq_tbl, by = "term")
```


```{r}
label_map <- c(
  "age0" = "Age (years)",
  "physical0" = "Baseline physical QOL",
  "mental0" = "Baseline mental QOL",
  "log_vload0" = "Baseline viral load (log10)",
  "cd40" = "Baseline CD4",
  "bmi0" = "BMI")

dropout_sig2 <- dropout_combined %>%
  mutate(
    # % change in odds (per unit increase)
    bay_pct  = (bay_or  - 1) * 100,
    bay_pct_lo = (bay_lo - 1) * 100,
    bay_pct_hi = (bay_hi - 1) * 100,
    freq_pct = (freq_or - 1) * 100,
    freq_pct_lo = (freq_lo - 1) * 100,
    freq_pct_hi = (freq_hi - 1) * 100) %>% 
  mutate(
    Predictor = recode(term, !!!label_map, .default = term),
    bay_ci  = sprintf("(%.2f, %.2f)", bay_pct_lo, bay_pct_hi),
    freq_ci = sprintf("(%.2f, %.2f)", freq_pct_lo, freq_pct_hi)) %>%
  select(Predictor, bay_pct, bay_ci, freq_pct, freq_ci, freq_p)
```

```{r label = "Used in final report as table 5"}
tbl_dropout_compare <- dropout_sig2 %>%
  mutate(
    n_lm  = nobs(dropout.model),
    n_brm = nrow(model.frame(dropout.brm))) %>% 
  filter(Predictor %in% c("Age (years)", "Baseline physical QOL")) %>% 
  select(n_brm, bay_pct,bay_ci,
         n_lm, freq_pct, freq_ci,freq_p,Predictor) %>% 
  gt(rowname_col = "Predictor") %>%
  cols_label(
    n_brm = "N", 
    n_lm = "N",
    bay_pct = "%Δ in Odds",
    bay_ci = "95% HDI",
    freq_pct = "%Δ in Odds",
    freq_ci = "95% CI",
    freq_p  = "P-value") %>%
  tab_spanner(
    label = "Bayesian",
    columns = c(n_brm, bay_pct, bay_ci)) %>%
  tab_spanner(
    label = "Frequentist",
    columns = c(n_lm, freq_pct, freq_ci, freq_p)) %>%
  fmt_number(columns = c(bay_pct, freq_pct), decimals = 2) %>%
  fmt_number(columns = freq_p, decimals = 3) %>%
    # bold specific text
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_stub(rows = everything())) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_spanners(
      spanners = c("Frequentist", "Bayesian"))) %>% 
  # p-value formatting like your other tables
  text_transform(
    locations = cells_body(columns = freq_p),
    fn = function(x) ifelse(as.numeric(x) < 0.001, "<0.001",
                            format(round(as.numeric(x), 3), nsmall = 3))) %>%
  tab_options(table.font.size = "small") %>%
  tab_source_note(
    source_note = "Values represent percent change in the odds of being missing 
    (dropout) per one-unit increase in the predictor (computed as (OR − 1) × 100).
    Negative values indicate lower odds of dropout.") %>% 
  tab_options(
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3),
    table.width = pct(85),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

tbl_dropout_compare
```


## Bayesian Diagnostic Plots

```{r, eval=FALSE}
all_models <- c(model.fits, list(dropout = dropout.brm))
all_med_models <- model.med.fits   

get_brms <- function(x) {
  if (inherits(x, "brmsfit")) return(x)
  if (is.list(x) && !is.null(x$brm) && inherits(x$brm, "brmsfit")) return(x$brm)
  stop("Couldn't find a brmsfit. Is it stored as $brm or is the element itself a brmsfit?")
  }

get_pars <- function(fit, pars_regex = "b_.*hard0Yes|b_.*adh_bin|sigma") {
  all_pars <- parnames(fit)
  all_pars[grepl(pars_regex, all_pars)]
  }

trace_one <- function(fit, title = NULL) {

  pars_regex <- if (!is.null(title) && grepl("dropout", title, ignore.case = TRUE)) {
    # dropout model: trace age0 and physical0 (Bernoulli has no sigma)
    "b_.*age0$|b_.*physical0$"
  } else {
    "b_.*hard0Yes|b_.*adh_bin|sigma"
  }

  pars <- get_pars(fit, pars_regex = pars_regex)

  if (length(pars) == 0) {
    message("No matching parameters found for: ", title)
    return(invisible(NULL))
  }

  p <- bayesplot::mcmc_trace(as.array(fit), pars = pars)
  if (!is.null(title)) p <- p + ggplot2::ggtitle(title)
  p
}

# ---- SAVE TO PDF ----
out_file <- "/Users/ryan_summers/GitHub/BIOS6624-Class/Project1/Reports/traceplots_all_models.pdf"

pdf(out_file, width = 10, height = , onefile = TRUE)

# main outcome models + dropout
iwalk(imap(all_models, ~ get_brms(.x)), ~ {
  p <- trace_one(.x, title = paste("Main:", .y))
  if (!is.null(p)) print(p)
})

# mediation models
iwalk(imap(all_med_models, ~ get_brms(.x)), ~ {
  p <- trace_one(.x, title = paste("Mediation:", .y))
  if (!is.null(p)) print(p)
})

dev.off()
out_file
```

`


