---
title: 'BIOS 6624 : Project 1'
author: "Ryan Summers"
date: "2026-02-04"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(BiocManager)
library(knitr)
library(ggplot2)
library(tidyverse)
library(magrittr)
library(ggfortify)
library(olsrr)
library(kableExtra)
library(doBy)
library(MASS)
library(dplyr)
library(table1)
library(gtsummary)
library(gt)
library(stringdist)
library(ggrepel)
library(impute)
library(limma)
library(qvalue)
library(gtools)
library(multtest)
library(GEOquery)
library(Biobase)
library(minfi)
library(R.utils)
library(stringr)
library(qs)
library(qs2)
library(lme4)
library(patchwork)
library(lme4)
library(lmerTest)
library(data.table)
library(emmeans)
library(lubridate)
library(naniar)
library(ggpubr)
library(boot)
library(mediation)
library(tidybayes)
library(priorsense)
library(bayesplot)
library(posterior)
```

## Read in the csv

```{r}
HIV_raw <- read_csv("/Users/ryan_summers/GitHub/BIOS6624-Class/Project1/DataRaw/hiv_6624_final.csv",
                    col_select = -1) # col_select = -1 removes first column
```
## Data Management & Cleanup

Convert specific variables to factors
```{r}
var_fct <- c("newid", "SMOKE", "RACE", "EDUCBAS", "hard_drugs")

HIV_raw <- HIV_raw %>% 
  mutate(
    newid = factor(newid),
    SMOKE = factor(SMOKE, levels = c(1,2,3), 
                   labels = c("Never", "Former", "Current")),
    RACE = factor(RACE, 
                  labels = c("White non-Hisp", "White Hisp", "Black non-Hisp",
                             "Black Hisp", "Other", "Other Hisp")),
    EDUCBAS = factor(EDUCBAS, 
                     labels = c("<=8th grade", "9,10,11th grade", "12th grade",
                             ">=1yr college", "4yr college degree", "Some grad", 
                             "Post-grad degree")),
    hard_drugs = factor(hard_drugs, levels = c(0,1),
                        labels = c("No", "Yes")),
    ADH = factor(ADH, 
                 levels = c(1,2,3,4),
                 labels = c("100%", "95-99%", "75-94%", "<75%")))
    
```

Collapse specific factors
* education and race - we are not making inferences or estimating the effects of either


## Exploratory Data Analysis (EDA)

Missing value analysis:
```{r}
# number of NAs
lapply(HIV_raw, function(x) sum(is.na(x)))
```

```{r}
# missing value percentages
HIV_raw %>%
  ungroup() %>% 
  miss_var_summary()

```

```{r}
# number of subjects
length(unique(HIV_raw$newid))
```

Look at distribution of measurements (cumulatively)
```{r}
# measurement stats
HIV_raw %>% 
  group_by(newid) %>%
  summarise(n_measurements = n()) -> mnts_per_id 

# summary stats for number of measurements per subject
mnts_per_id %>%   
  summarise(avg = mean(n_measurements),
            min = min(n_measurements),
            med = median(n_measurements),
            max = max(n_measurements))

# distribution of measurements
hist(mnts_per_id$n_measurements, breaks=11, freq = FALSE)
```

```{r}
# percentage distribution of measurements
mnts_per_id %>%
  dplyr::count(n_measurements) %>%
  mutate(percent = round(100 * n / sum(n) , 1))

# cumulative percentages (e.g. % more than 3 measurements)
mnts_per_id %>%
  reframe(
    threshold = 1:9,
    percent_gt = sapply(threshold, function(k) {
      mean(n_measurements > k) * 100
      }))

```

How many observations per race category?
```{r}
HIV_raw %>% 
  group_by(RACE) %>% 
  summarize(n = n())
```

Problems arise when race is:
* Interpreted biologically
* Used as a proxy for genetic ancestry
* Over-interpreted or over-stratified without power

We are not interpreting or estimating race effects, so we should be good to collapse

Potential explanation in final report:
Self-reported race/ethnicity was included as a covariate to account for potential confounding by social determinants of health and healthcare access disparities. Race/ethnicity categories were collected as [describe original categories]. For analyses with limited statistical power due to sample size, we combined categories as follows: [describe any collapsing]. We acknowledge that race is a social construct and that observed differences likely reflect structural inequities rather than biological differences.

Look at distributions of outcomes

```{r}
hist(HIV_raw$VLOAD)
```

Spaghetti plots

```{r}
HIV_raw %>% 
  filter(years %in% c(0,1,2) & !is.na(hard_drugs)) %>%

  
  ggplot(aes(x = years, y = log10(VLOAD), group = newid, color = hard_drugs)) +
  geom_line(alpha = 0.25, linewidth = 0.6) +
  geom_point(alpha = 0.25, size = 1.0) +
  stat_summary(aes(group = hard_drugs), fun = mean, geom = "line",
               linewidth = 0.75, color = "firebrick") +
  facet_wrap(~ hard_drugs, ncol = 3) +
  scale_color_manual(values = c("steelblue", "rosybrown2")) +
  #geom_smooth(aes(group=hard_drugs), se=F) +
  labs(
    title = "Viral Load Trajectories From Baseline to Year 2, Stratified by Drug Users",
    subtitle = "Faint lines = subjects; bold red lines = group means",
    x = "Year", y = "Viral Load log10 copies/ml", color = "Group") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# number missing for hard drug group
HIV_raw %>% 
  group_by(hard_drugs) %>% 
  summarize(n = n())
```


```{r}
# conver to long format
long_outcomes <- HIV_raw %>%
  filter(years %in% c(0,2) & !is.na(hard_drugs)) %>%
  mutate(log_vload = log10(VLOAD)) %>% 
           
  pivot_longer(cols = c(log_vload, LEU3N, AGG_PHYS, AGG_MENT),
               names_to = "outcome",
               values_to = "value") %>% 
    
    mutate(
      outcome = recode(outcome,
                       log_vload = "Viral Load (log10 copies/mL)",
                       LEU3N = "CD4+ T-Cell Count",
                       AGG_PHYS = "Physical QOL (0 to 100)",
                       AGG_MENT = "Mental QOL (0 to 100)"),
      years_f = factor(years, levels = c(0, 1, 2), labels = c("0", "1", "2")))

ggplot(long_outcomes, aes(years, value, group = newid, color = hard_drugs)) +
  geom_line(alpha = 0.05) +
  stat_summary(aes(group = hard_drugs), fun = mean, 
               geom = "line", linewidth = 0.6) +
  facet_wrap(~ outcome, scales = "free_y") +
  scale_color_manual(values = c("steelblue", "firebrick")) +
  labs(title = "Observed Trajectories of Treatment Response",
       x = "Years",
       y="",
       color = "Hard Drugs") +
  theme_bw() +
  theme(legend.position = "bottom") 

```

```{r}
ggplot(long_outcomes, aes(x = years_f, y = value, fill = hard_drugs)) +
  geom_boxplot(outlier.shape = NA, width = 0.7, position = position_dodge(width = 0.75)) +
  geom_point(
    aes(color = hard_drugs),
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
    alpha = 0.08, size = 0.6
  ) +
  facet_wrap(~ outcome, scales = "free_y") +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_color_manual(values = c("steelblue", "firebrick")) +
  labs(
    title = "Treatment Response Distributions by Year and Baseline Hard Drug Use",
    x = "Year since HAART initiation",
    y = NULL,
    fill = "Hard Drugs",
    color = "Hard Drugs"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```


## Model Building

```{r}

base <- HIV_raw %>%
  filter(years == 0) %>%
  transmute(
    newid,
    hard0 = hard_drugs,
    age0  = age,
    bmi0  = BMI,
    smoke0 = SMOKE,
    educ0  = EDUCBAS,
    race0  = RACE,
    vload0 = VLOAD,
    cd40   = LEU3N,
    physical0  = AGG_PHYS,
    mental0  = AGG_MENT,
    cesd0 = CESD,
    hbp0 = HBP,
    diab0 = DIAB,
    liv34_0 = LIV34,
    kid0 = KID)


y2 <- HIV_raw %>%
  filter(years == 2) %>%
  transmute(
    newid,
    vload2 = VLOAD,
    cd42   = LEU3N,
    physical2  = AGG_PHYS,
    mental2  = AGG_MENT,
    adh2 = ADH,
    cesd2 = CESD,
    hbp2 = HBP,
    diab2 = DIAB,
    liv34_2 = LIV34,
    kid2 = KID)

ana <- base %>%
  left_join(y2, by="newid") %>% 
    mutate(
    log_vload0 = log10(vload0), # **based on literature log10 is typically used**
    log_vload2 = log10(vload2),
    log_cd40 = log10(cd40),
    log_cd42 = log10(cd42),
    d_log_vload = log_vload2 - log_vload0,
    d_vload = vload2 - vload0,  
    d_cd4 = cd42 - cd40,
    d_phys = physical2 - physical0,
    d_ment = mental2 - mental0)

```

Calculate the number of insufficient data labels for the comorbidities
```{r}
HIV_raw %>%
  filter(years == 0) %>%
  summarise(
    across(
      c(DIAB, HBP, KID, LIV34, FRP, FP),
      ~ sum(. == 9, na.rm = TRUE),
      .names = "n9_{.col}"))
```

Collapse levels of comorbidities into a simple yes or no category with 9s being converted to NAs. Additionally, for the depression scale (CESD), convert -1 to NAs

```{r}
# using baseline values
ana2 <- ana %>%
  mutate(
    diab0 = case_when(
      diab0 %in% c(1, 3) ~ "No",
      diab0 %in% c(2, 4) ~ "Yes",
      diab0 == 9 ~ NA_character_,
      TRUE ~ NA_character_) %>% 
      factor(levels = c("No", "Yes")),
    hbp0 = case_when(
      hbp0 %in% c(1, 3) ~ "No",
      hbp0 %in% c(2, 4) ~ "Yes",
      hbp0 == 9 ~ NA_character_,
      TRUE ~ NA_character_) %>% 
      factor(levels = c("No", "Yes")),
    liv34_0 = case_when(
      liv34_0 %in% c(1, 3) ~ "No",
      liv34_0 %in% c(2, 4) ~ "Yes",
      liv34_0 == 9 ~ NA_character_,
      TRUE ~ NA_character_) %>% 
      factor(levels = c("No", "Yes")),
    kid0 = case_when(
      kid0 %in% c(1, 3) ~ "No",
      kid0 %in% c(2, 4) ~ "Yes",
      kid0 == 9 ~ NA_character_,
      TRUE ~ NA_character_) %>% 
      factor(levels = c("No", "Yes")),
    educ0 = case_when(
      educ0 %in% c("<=8th grade", "9,10,11th grade", "12th grade") ~ "High School or less",
      educ0 %in% c(">=1yr college", "4yr college degree") ~ "College",
      educ0 %in% c("Some grad", "Post-grad degree") ~ "Post-grad",
      TRUE ~ educ0) %>%
      factor(),
    race0 = case_when(
      race0 %in% c("White Hisp", "Black Hisp", "Other Hisp", "Other") ~ "Hispanic",
      TRUE ~ race0) %>% 
      factor(),
    cesd0 = ifelse(cesd0==-1, NA_character_, cesd0),
    cesd0 = as.numeric(cesd0))
```


## Create table1

```{r}
tbl1 <- ana2 %>%
  
  tbl_summary(
    by = hard0,
    include = c(mental0, physical0, log_vload0, cd40, age0, bmi0, smoke0, 
                educ0, race0, cesd0), # add ADH or no?
    type = all_continuous() ~ "continuous2",
    statistic = list(
      all_continuous() ~ c(
        "{mean} ({sd})",
        "{median} ({p25}, {p75})",
        "{min} - {max}"),
      all_categorical() ~ "{n} ({p}%)"),
    
    digits = all_continuous() ~ 1,
    missing = "ifany",
    missing_text = "NA",
    missing_stat = "{N_miss} ({p_miss}%)",
    
    # change variable names
    label = list(
      mental0 ~ "Mental QOL",
      physical0 ~ "Physical QOL",
      log_vload0  ~  "Viral Load (log<sub>10</sub> copies/mL)",
      cd40  ~ "CD4 T-cell Count",
      age0 ~ "Age",
      bmi0 ~ "BMI",
      smoke0 ~ "Smoke",
      educ0 ~ "Education",
      race0 ~ "Race",
      cesd0 ~ "Depression Scale (>=16 is depressed)")) %>% 
  
  add_overall() %>%
  bold_labels() %>% 
  italicize_levels() %>% 
  #add_p(pvalue_fun = label_style_pvalue(digits = 2)) %>% 
  add_n() %>% 
  modify_spanning_header(
    c("stat_1", "stat_2") ~ "**Hard Druge Use**") %>%
  modify_footnote_header(
    footnote = "Table includes all subjects from cohort at baseline (years=0)",
    columns = all_stat_cols(),
    replace = FALSE) %>% 
  
  as_gt() %>%
  gt::fmt_markdown(columns = everything()) %>% 
  gt::fmt_number(
    columns = dplyr::starts_with("p.value"),
    decimals = 2) %>% 
  tab_options(
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))


tbl1

```

log10 interpretations 
* 10^n = where n is number of places after 1
* 10^8 ≈ in hundreds of millions
* 10^5 ≈ hundreds of thousands
* 10^4 ≈ tens of thousands
* 10^0.2 ≈ 1.6 - in single digits - remember exponent of 0 is 1Create a table 2 to look at univariate relationships between hard drug use and comorbidities

```{r}
tbl2 <- ana2 %>%
  
  tbl_summary(
    by = hard0,
    include = c(hbp0, diab0, liv34_0, kid0),
    type = all_continuous() ~ "continuous2",
    statistic = list(
      all_continuous() ~ c(
        "{mean} ({sd})",
        "{median} ({p25}, {p75})",
        "{min} - {max}"),
      all_categorical() ~ "{p}%"),
    
    digits = all_continuous() ~ 1,
    missing = "ifany",
    missing_text = "NA",
    missing_stat = "{p_miss}%",
    
    # change variable names
    label = list(
      hbp0 ~ "High Bloood Pressure",
      diab0 ~ "Diabetes",
      liv34_0  ~  "Liver Disease (Stage 3/4)",
      kid0  ~ "Kidney Disease")) %>%
  
  add_overall() %>%
  bold_labels() %>% 
  italicize_levels() %>% 
  add_p(pvalue_fun = label_style_pvalue(digits = 2)) %>% 
  modify_spanning_header(
    c("stat_1", "stat_2") ~ "**Hard Druge Use**") %>%
  modify_footnote_header(
    footnote = "Table includes all subjects from cohort at baseline (years=0)",
    columns = all_stat_cols(),
    replace = FALSE) %>% 
  
  as_gt() %>%
  gt::fmt_markdown(columns = everything()) %>% 
  gt::fmt_number(
    columns = dplyr::starts_with("p.value"),
    decimals = 2) %>% 
  tab_options(
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

tbl2
```

* Collider bias? - do these comorbidities occur after exposure and outcome?
 - The exposure and outcome can both independently create a collider variable. In contrast to the case of confounding, where the confounder is associated with both the exposure and outcome, the collider is created by both the exposure and outcome.
 - Comorbidities can be downstream effects of higher viral load/cd4 and hard drug use,
 independently. Therefore, these comorbidities could very well be colliders
 
* Rates of liver and kidney disease are higher in drug users (3.2% vs 9.7% and 16% vs 34%, respectively). No difference in diabetes and higher blood pressure is more prevalent in
the non-drug use group (27% vs 15%). Also, there is pretty substantial missing values
for the kidney (71%), diabetes (57%), and liver (34%) groups. 


```{r}
options(scipen = 999)

vload.model2 <- lm(log_vload2 ~ factor(hard0) + log_vload0 + age0 + bmi0 + factor(smoke0) + 
                       factor(educ0) + factor(race0),
              data=ana)

summary(vload.model2)

tidy(vload.model2, conf.int = TRUE, exponentiate = FALSE) %>% 
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = round(p.value, 3),
    conf.low = round(conf.low, 3),
    conf.high = round(conf.high, 3),
    vload_copies_ml = round(10^estimate, 3),
    vload_ml_LL = round(10^conf.low, 3),
    vload_ml_UL = round(10^conf.high, 3))
```

After adjusting for baseline viral load, age, BMI, smoking, education, and race, baseline hard-drug users had an estimated 42.9% lower viral load at Year 2 compared with non-users (1-0.571), but this difference did not reach significance (p = 0.232).

* We are comparing hard-drug users and non-users while holding the baseline viral load, age, BMI, smoking status, education, and race values constant. In other words, after other affects have been accounted for.
* The model estimates differences in Year-2 viral load between groups after removing variation attributable to baseline viral load and covariates, effectively comparing individuals at the same starting level.

Test to see if comorbidities can be used as precision variables or may be
on the casual pathway from drug use to viral load
```{r}
options(scipen = 999)

vload.model3 <- lm(log_vload2 ~ 
                       # comorbidites
                       factor(hbp0) + factor(diab0) + factor(liv34_0) + factor(kid0) +
                       # Primary
                       factor(hard0) + log_vload0 + age0 + bmi0 + 
                       factor(smoke0) + factor(educ0) + factor(race0),
              data=ana2)

summary(vload.model3)
tidy(vload.model3, conf.int = TRUE, exponentiate = FALSE) %>% 
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = round(p.value, 3),
    conf.low = round(conf.low, 3),
    conf.high = round(conf.high, 3),
    vload_copies_ml = round(10^estimate, 3),
    vload_ml_LL = round(10^conf.low, 3),
    vload_ml_UL = round(10^conf.high, 3))
```

Look at multicollinearity issues if present
```{r}
car::vif(vload.model3)
```


* hard drug use completely changes direction when including comorbidites. It now has an
estimated 12.2% higher viral load as compared to non-drug users

Compare natural log and log10 transformations. log10 is standard use in prior literature in
HIV studies
```{r}
hist(log10(ana$vload0))
hist(log(ana$vload0))
```

Build CD4 outcome model (frequentist). Did not transform as residuals looked good with
raw values.
```{r}
options(scipen = 999)

cd4.model <- lm(cd42 ~ factor(hard0) + cd40 + age0 + bmi0 + factor(smoke0) + 
                       factor(educ0) + factor(race0),
              data=ana2)

summary(cd4.model)

tidy(cd4.model, conf.int = TRUE, exponentiate = FALSE) %>% 
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = round(p.value, 3),
    conf.low = round(conf.low, 3),
    conf.high = round(conf.high, 3))
```

Indentify why I'm getting contrast error
```{r}
table(ana2$hard0, useNA = "ifany")
table(ana2$smoke0, useNA = "ifany")
table(ana2$educ0, useNA = "ifany")
table(ana2$race0, useNA = "ifany")
```


After adjusting for baseline CD4+ T cell counts age, BMI, smoking, education, and race, baseline hard-drug users had an estimated 42.9% lower viral load at Year 2 compared with non-users (1-0.571), but this difference did not reach significance (p = 0.232).

Look at distribution of CD4 T cell count, log counts, and boxplots between drug user group
```{r}
hist(ana2$cd40)
hist(ana2$log_cd40)
boxplot(ana2$log_cd40~ana2$hard0)
boxplot(ana2$cd40~ana2$hard0)
```

Check that residuals look good using raw CD4 counts
```{r}
# check residuals to determine if skewness is too extreme for this model approach
plot(cd4.model)   # residuals vs fitted

```

* some outliers but overall residuals look good with raw CD4 values


Check distribution of QOL measures to identify odd values
```{r}
hist(ana2$physical0)
summary(ana2$physical0)
hist(ana2$mental0)
summary(ana2$mental0)

```



Build Mental QOL outcome model (frequentist)
```{r}
options(scipen = 999)

mental.model <- lm(mental2 ~ factor(hard0) + mental0 + age0 + bmi0 + factor(smoke0) + 
                       factor(educ0) + factor(race0),
              data=ana2)

summary(mental.model)

tidy(mental.model, conf.int = TRUE, exponentiate = FALSE) %>% 
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = round(p.value, 3),
    conf.low = round(conf.low, 3),
    conf.high = round(conf.high, 3))
```


Build Physical QOL outcome model (frequentist)
```{r}
options(scipen = 999)

physical.model <- lm(physical2 ~ factor(hard0) + physical0 + age0 + bmi0 + factor(smoke0) + 
                       factor(educ0) + factor(race0),
              data=ana2)

summary(physical.model)

tidy(physical.model, conf.int = TRUE, exponentiate = FALSE) %>% 
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = round(p.value, 3),
    conf.low = round(conf.low, 3),
    conf.high = round(conf.high, 3))
```

## Mediation Analysis

Adherence variable is ordinal in nature and values are not evenly spaced. Therefore,
we can't use a linear approach. Let's try converting to a binary and see results

```{r}
ana2 <- ana2 %>%
  mutate(
    adh_bin = case_when(
      adh2 %in% c("100%","95-99%") ~ 1,  # ≥95%
      adh2 %in% c("75-94%", "<75%") ~ 0,  # <95%
      TRUE ~ NA_real_),
    adh_bin = factor(adh_bin, levels = c(0,1),
                     labels = c("<95%", "≥95%")))
```

We also need a complete case analysis or we will keep getting "number of observations do not match between mediator and outcome models" error

```{r}
# complete case analysis
ana2_cc <- ana2 %>%
  select(cd42, hard0, adh_bin, cd40,
         age0, bmi0, smoke0, educ0, race0) %>%
  na.omit() %>%
  droplevels()
```



Custom function for calculating mediation coefficients
```{r}
# boot_fun <- function(data, idx) {
#   d <- data[idx, ]
#   m <- glm(adh_bin ~ hard0 + age0 + bmi0 + smoke0 + educ0 + race0, 
#            family = binomial,
#            data = d)
#   y <- lm(cd42 ~ hard0 + adh_bin + cd40 +
#             age0 + bmi0 + smoke0 + educ0 + race0, data = d)
# 
#   a <- coef(m)["hard0"]     
#   b <- coef(y)["adh2"]
#   cprime <- coef(y)["hard0"]
# 
#   c(indirect = a*b, direct = cprime, total = cprime + a*b)
# }
# 
# set.seed(1)
# bt <- boot(ana2_cc, statistic = boot_fun, R = 2000)
# 
# boot.ci(bt, type = "perc", index = 1)  # indirect
# boot.ci(bt, type = "perc", index = 2)  # direct
# boot.ci(bt, type = "perc", index = 3)  # total
```


Pre-lim mediation analysis w/out Covs
```{r}
# Fit our three models
mod2_crude <- lm(cd42 ~ hard0, data=ana2_cc) # Y~X
mod2_adjusted <- lm(cd42 ~ hard0 + adh_bin, data=ana2_cc) # Y~ X+M
mod2_covariate <- glm(adh_bin ~ hard0, 
                     family = binomial,
                     data=ana2_cc) # M~X

# Extract coefficients
coef(mod2_crude)
coef(mod2_adjusted)
coef(mod2_covariate)
```


DAG
```{tikz, tikz-med, fig.ext = 'png', cache=TRUE, echo=F, fig.align="center", fig.width=4}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=4cm, auto,>=latex, thick, scale = 0.5]
\node (X) [align=center] {Hard Drug Use \\ ($X$)};
\node (M) [align=center,above of=X,xshift=2cm,yshift=-2cm] {Adherence \\ ($M$)};
\node (Y) [align=center,right of=X] {Year 2 CD4 \\ ($Y$)};
\draw[->] (X) to node [above left] {$\hat{\gamma}_{X}=1.59$} (M);
\draw[->] (M) to node [above right] {$\hat{\beta}_{M}=98.03$} (Y);
\draw[->] (X) to node [align=center,swap] {$\hat{\beta}_{adj}=-202.55$ \\ $\hat{\beta}_{crude}=-193.91$} (Y);
\end{tikzpicture}
```

Using mediation package
```{r}
# Mediator model: adherence as outcome
m_mod <- glm(adh_bin ~ hard0 + age0 + bmi0 + smoke0 + educ0 + race0, 
             family = binomial,
            data = ana2_cc)

# Outcome model: Year-2 viral load, adjusting for baseline viral load + adherence
y_mod <- lm(cd42 ~ hard0 + adh_bin + cd40 +
              age0 + bmi0 + smoke0 + educ0 + race0, 
            data = ana2_cc)

# Mediation estimation (bootstrap for inference)
med_vl <- mediate(m_mod, y_mod, treat = "hard0", mediator = "adh_bin",
                  boot = F, sims = 2000)

summary(med_vl)

```

```{r}
table(ana2_cc$race0)
table(ana2_cc$adh_bin)
```
```{r}
library(brms)

# Mediator model
med_model <- bf(adh_bin ~ hard0 + cd40 + age0 + bmi0 + smoke0 + educ0 + race0,
                family = bernoulli(link = "logit"))

# Outcome model
out_model <- bf(cd42 ~ hard0 + adh_bin + cd40 +
                age0 + bmi0 + smoke0 + educ0 + race0,
                family = gaussian())

med.brm.fit <- brm(med_model + out_model + set_rescor(FALSE),
                   chains = 4, cores = 6, iter = 2000,
           data = ana2_cc)
```

```{r}
table(ana2_cc$adh_bin, useNA="ifany")
```

```{r}
summary(med.brm.fit)
```


Extract the draws

```{r}
# extract posterior draws 
med_draws <- as_draws_df(med.brm.fit)

# get parameter names
names(med_draws)[grepl("^b_", names(med_draws))]
```

### Diagnsotics for Bayesian Mediation Model

```{r, eval=F}
# diagnostics
mcmc_areas(med_draws,prob = 0.95,
             pars = c("b_adhbin_hard0Yes"))
```

```{r, eval=F}
# Trace plots for each parameter
mcmc_trace(post_samples,
           pars = c("b_adhbin_hard0Yes", "b_cd42_hard0Yes"))

```

```{r}
levels(ana2_cc$hard0)
names(med_draws)[grepl("^b_", names(med_draws))]
```

```{r}
## X = drug use & M = adherence

## We are using complete-case analysis so use ana2_cc dataset

# make sure hard0 has the right reference 
ana2_cc$hard0 <- relevel(factor(ana2_cc$hard0), ref = "No")

# mediator levels - current reference is <95%
levM <- levels(ana2_cc$adh_bin)
m0 <- levels(ana2_cc$adh_bin)[1] # reference
m1 <- levels(ana2_cc$adh_bin)[2]      

# create new dataset with X set to No and Yes
hd_NO <- ana2_cc
hd_NO$hard0 <- factor("No", levels = levels(ana2_cc$hard0))

hd_YES <- ana2_cc
hd_YES$hard0 <- factor("Yes", levels = levels(ana2_cc$hard0))

# Posterior expected adherence probabilities P(M=1|X,C)
# For bern, posterior_epred gives probability of success (i.e., P(M == m1))
pM0 <- posterior_epred(med.brm.fit, resp = "adhbin", newdata = hd_NO)  
pM1 <- posterior_epred(med.brm.fit, resp = "adhbin", newdata = hd_YES)  

# Posterior expected CD4 means for each X and each fixed mediator level
# Build newdata that sets adh_bin to m0 or m1
hd_NO_lt95 <- hd_NO; hd_NO_lt95$adh_bin <- factor(m0, levels = levM)
hd_NO_gt95 <- hd_NO; hd_NO_gt95$adh_bin <- factor(m1, levels = levM)

hd_YES_lt95 <- hd_YES; hd_YES_lt95$adh_bin <- factor(m0, levels = levM)
hd_YES_gt95 <- hd_YES; hd_YES_gt95$adh_bin <- factor(m1, levels = levM)

# expected CD4 if No hard drugs w/ <95% ADH + Covs
mu0_m0 <- posterior_epred(med.brm.fit, resp = "cd42", newdata = hd_NO_lt95)  
# expected CD4 if No hard drugs w/ ≥95% ADH + Covs
mu0_m1 <- posterior_epred(med.brm.fit, resp = "cd42", newdata = hd_NO_gt95)  
# expected CD4 if Yes hard drugs w/ <95% ADH + Covs
mu1_m0 <- posterior_epred(med.brm.fit, resp = "cd42", newdata = hd_YES_lt95)  
# expected CD4 if Yes hard drugs w/ ≥95% ADH + Covs
mu1_m1 <- posterior_epred(med.brm.fit, resp = "cd42", newdata = hd_YES_gt95)  

## Standardize over the mediator distribution:
 # 1 - pM0 = prob of low ADH ; pM0 = prob of high ADH | No hard drug use
 # 1 - pM1 = prob of low ADH ; pM1 = prob of high ADH | Yes hard drug use
EY_0M0 <- rowMeans(mu0_m0 * (1 - pM0) + mu0_m1 * pM0) # pop-level E[CD4|X=0]
EY_1M1 <- rowMeans(mu1_m0 * (1 - pM1) + mu1_m1 * pM1) # pop-level E[CD4|X=1]

# expectation of CD4 if a hard drug user but ADH behaved as if they were NOT
# a drug user
EY_1M0 <- rowMeans(mu1_m0 * (1 - pM0) + mu1_m1 * pM0)

## CD4 depends on adherence.
## ADH depends on drug use.
## So to understand mediation, we must avg CD4 preds over how often people are ADH

# calculate the effects
TE  <- EY_1M1 - EY_0M0
NIE <- EY_1M1 - EY_1M0
NDE <- EY_1M0 - EY_0M0

# summarize
summ <- function(x) {
  c(
    mean = mean(x),
    hdi_low = hdi(x, prob = 0.95)[1],
    hdi_high = hdi(x, prob = 0.95)[2],
    Pr_gt_0 = mean(x > 0))
  }

rbind(TE = summ(TE), NIE = summ(NIE), NDE = summ(NDE))
```

Total Effect (TE)
Interpretation:
Moving from No hard drug use to Yes hard drug use is associated with an average decrease of about 168 CD4 cells at Year 2.

Indirect Effect (NIE-mediation through adherence)
Interpretation:
There is strong posterior evidence that adherence mediates part of the association.

Direct Effect (NDE)
Interpretation:
Even if adherence were held constant, hard drug use would still be associated with a ~174 cell lower CD4 count.

This suggests that most of the effect is not explained by adherence

Proportion Mediated
mean(NIE / TE) = 6/168 ≈ 3.6%


## Dropout Rates

```{r}
ana2$complete_cd4 <- ifelse(!is.na(ana2$cd42), 1, 0)

table(ana2$complete_cd4, ana2$hard0)
t.test(cd40 ~ complete_cd4, data = ana2)
```

```{r}
ana2_drop <- ana2 %>%
  mutate(
    hard0 = factor(hard0, levels = c("No", "Yes"), exclude = NULL),
    followup_status = factor(
      if_else(complete == 1, "Complete (baseline + Year 2) CD4", "Missing Year 2 CD4"),
      levels = c("Complete (baseline + Year 2)", "Missing Year 2")))

tbl_dropout <- ana2_drop %>%
  tbl_summary(
    by = followup_status,
    include = c(
      hard0, age0, bmi0, smoke0, educ0, race0,
      log_vload0, cd40, physical0, mental0, cesd0),
    type = list(
      hard0 ~ "categorical",
    all_continuous() ~ "continuous2"),
    statistic = list(
      all_continuous() ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
      all_categorical() ~ "{n} ({p}%)"),
    digits = all_continuous() ~ 1,
    missing = "ifany",
    missing_text = "NA",
    missing_stat = "{N_miss} ({p_miss}%)",
    label = list(
      hard0 ~ "Baseline hard drug use",
      age0 ~ "Age",
      bmi0 ~ "BMI",
      smoke0 ~ "Smoking status",
      educ0 ~ "Education",
      race0 ~ "Race/ethnicity",
      log_vload0 ~ "Baseline viral load (log10 copies/mL)",
      cd40 ~ "Baseline CD4 T-cell Count",
      physical0 ~ "Baseline physical QOL",
      mental0 ~ "Baseline mental QOL",
      cesd0 ~ "Baseline CESD score")) %>%
  add_p(pvalue_fun = ~style_pvalue(.x, digits = 3)) %>%
  add_n() %>% 
  add_overall() %>%
  bold_labels() %>% 
  as_gt() %>%
    modify_footnote_header(
    footnote = "Follow-up status defined based on availability of Year 2 CD4 measurement.",
    columns = all_stat_cols(),
    replace = FALSE)
  tab_options(
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

tbl_dropout
```

Missing data under ana2 format where each row is one subject showing baseline and year 2
```{r}
ana2_drop %>%
  ungroup() %>% 
  miss_var_summary()
```


