---
title: 'BIOS 6624 : Project 1'
author: "Ryan Summers"
date: "2026-02-04"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(BiocManager)
library(knitr)
library(ggplot2)
library(tidyverse)
library(magrittr)
library(ggfortify)
library(olsrr)
library(kableExtra)
library(doBy)
library(MASS)
library(dplyr)
library(table1)
library(gtsummary)
library(gt)
library(stringdist)
library(ggrepel)
library(impute)
library(limma)
library(qvalue)
library(gtools)
library(multtest)
library(GEOquery)
library(Biobase)
library(minfi)
library(R.utils)
library(stringr)
library(qs)
library(qs2)
library(lme4)
library(patchwork)
library(lme4)
library(lmerTest)
library(data.table)
library(emmeans)
library(lubridate)
library(naniar)
library(ggpubr)
library(boot)
library(mediation)
library(tidybayes)
library(priorsense)
library(bayesplot)
library(posterior)
library(brms)
library(bayestestR)
library(purrr)
library(loo)
library(report)

# load scripts
source("/Users/ryan_summers/GitHub/BIOS6624-Class/Project1/Code/group_counts_ftn.R")
```

## Read in the csv

```{r}
HIV_raw <- read_csv("/Users/ryan_summers/GitHub/BIOS6624-Class/Project1/DataRaw/hiv_6624_final.csv",
                    col_select = -1) # col_select = -1 removes first column
```

## Data Management & Cleanup

Convert specific variables to factors and make non-drug users the reference
group for variable of interest
```{r}
var_fct <- c("newid", "SMOKE", "RACE", "EDUCBAS", "hard_drugs")

HIV_raw <- HIV_raw %>% 
  mutate(
    newid = factor(newid),
    SMOKE = factor(SMOKE, levels = c(1,2,3), 
                   labels = c("Never", "Former", "Current")),
    RACE = factor(RACE, 
                  labels = c("White non-Hisp", "White Hisp", "Black non-Hisp",
                             "Black Hisp", "Other", "Other Hisp")),
    EDUCBAS = factor(EDUCBAS, 
                     labels = c("<=8th grade", "9,10,11th grade", "12th grade",
                             ">=1yr college", "4yr college degree", "Some grad", 
                             "Post-grad degree")),
    # non-users will be the reference for modelling
    hard_drugs = factor(hard_drugs, levels = c(0,1),
                        labels = c("No", "Yes")),
    ADH = factor(ADH, 
                 levels = c(1,2,3,4),
                 labels = c("100%", "95-99%", "75-94%", "<75%")))
    
```


Create a new dataset with only baseline and year 2 observations (715 subjects)
```{r}

base <- HIV_raw %>%
  filter(years == 0) %>%
  transmute(
    newid,
    hard0 = hard_drugs,
    age0  = age,
    bmi0  = BMI,
    smoke0 = SMOKE,
    educ0  = EDUCBAS,
    race0  = RACE,
    vload0 = VLOAD,
    cd40   = LEU3N,
    physical0  = AGG_PHYS,
    mental0  = AGG_MENT,
    cesd0 = CESD,
    hbp0 = HBP,
    diab0 = DIAB,
    liv34_0 = LIV34,
    kid0 = KID)


y2 <- HIV_raw %>%
  filter(years == 2) %>%
  transmute(
    newid,
    vload2 = VLOAD,
    cd42   = LEU3N,
    physical2  = AGG_PHYS,
    mental2  = AGG_MENT,
    adh2 = ADH,
    cesd2 = CESD,
    hbp2 = HBP,
    diab2 = DIAB,
    liv34_2 = LIV34,
    kid2 = KID)

ana <- base %>%
  left_join(y2, by="newid") %>% 
    mutate(
    log_vload0 = log10(vload0), # **based on literature log10 is typically used**
    log_vload2 = log10(vload2),
    log_cd40 = log10(cd40),
    log_cd42 = log10(cd42),
    d_log_vload = log_vload2 - log_vload0,
    d_vload = vload2 - vload0,  
    d_cd4 = cd42 - cd40,
    d_phys = physical2 - physical0,
    d_ment = mental2 - mental0)

```


Collapse levels of comorbidities into a simple yes or no category with 9s being converted to NAs. Additionally, for the depression scale (CESD), convert -1 to NAs

```{r label="Collapsing categorical levels"}
# using baseline values
ana2 <- ana %>%
  mutate(
    diab0 = case_when(
      diab0 %in% c(1, 3) ~ "No",
      diab0 %in% c(2, 4) ~ "Yes",
      diab0 == 9 ~ NA_character_,
      TRUE ~ NA_character_) %>% 
      factor(levels = c("No", "Yes")),
    hbp0 = case_when(
      hbp0 %in% c(1, 3) ~ "No",
      hbp0 %in% c(2, 4) ~ "Yes",
      hbp0 == 9 ~ NA_character_,
      TRUE ~ NA_character_) %>% 
      factor(levels = c("No", "Yes")),
    liv34_0 = case_when(
      liv34_0 %in% c(1, 3) ~ "No",
      liv34_0 %in% c(2, 4) ~ "Yes",
      liv34_0 == 9 ~ NA_character_,
      TRUE ~ NA_character_) %>% 
      factor(levels = c("No", "Yes")),
    kid0 = case_when(
      kid0 %in% c(1, 3) ~ "No",
      kid0 %in% c(2, 4) ~ "Yes",
      kid0 == 9 ~ NA_character_,
      TRUE ~ NA_character_) %>% 
      factor(levels = c("No", "Yes")),
    educ0 = case_when(
      educ0 %in% c("<=8th grade", "9,10,11th grade", "12th grade", ">=1yr college") 
      ~ "Didn't finish college or lower",
      educ0 %in% c("4yr college degree", "Some grad", "Post-grad degree") 
      ~ "College degree or higher",
      TRUE ~ educ0) %>%
      factor(),
    race0 = case_when(
      race0 %in% c("White Hisp", "Black Hisp", "Other Hisp", "Other") ~ "Hispanic",
      TRUE ~ race0) %>% 
      factor(),
    smoke0 = ifelse(smoke0=="Never", "Former", smoke0) %>% factor(),
    # convert adherence into a binary indicator
      adh_bin = case_when(
      adh2 %in% c("100%","95-99%") ~ 1,  # ≥95%
      adh2 %in% c("75-94%", "<75%") ~ 0,  # <95%
      TRUE ~ NA_real_),
    adh_bin = factor(adh_bin, levels = c(0,1),
                     labels = c("<95%", "≥95%")),
    # insufficient data and implausible values for some BMIs
    bmi0 = ifelse(bmi0 %in% c(-1, 999), NA, bmi0),
    cesd0 = ifelse(cesd0==-1, NA_character_, cesd0),
    cesd0 = as.numeric(cesd0))
```


We also need a complete case analysis or we will keep getting "number of observations do not match between mediator and outcome models" error. Also, to be consistent across analyses its probably best we keep the same number of subjects across all models.

**This will be the finalized dataset used for all analyses**
* 464 subjects with complete cases for baseline and year 2 measurements
```{r}
# complete case analysis
ana2_cc <- ana2 %>%
  select(log_vload2, log_vload0, cd42, cd40, hard0, adh2, physical0, physical2,
         mental0, mental2, age0, bmi0, smoke0, educ0, race0, adh_bin) %>%
  na.omit() %>%
  droplevels()
```

## Exploratory Data Analysis (EDA)

Explain how you start with 715 subjects to final N for analyses 
Clinically meaning changes
* .5 log10 VL
* 2pts on QOL
* 50 CD4

***Remove BMI values that are negative 1 and extreme***
***Explain why we choose a specific modeling strategy in Methods***
***If you remove observations, state if they were removed from all analyses***
***Keep track of sample sizes for each model ran***
***Show table1 with all subjects that will be used in analyses***
***Don't show regression coefficients - be more informative to a non-stats person***
***Label the plots in code that are used in final report***
***Add frequentist and Bayesian results in one table***
***reference worksheet 5 for method section write-up***
***collapse smoker category***
***Descriptive for adherence and fit hard drug and hard drug + ADH***

Test function
```{r}
# load from starting R code chunk
count_grp_obs(ana, vars = c("hard0", "smoke0"), mode = "interaction")
```




Missing value analysis:
```{r}
# number of NAs
# lapply(HIV_raw, function(x) sum(is.na(x)))
```

```{r}
# missing value percentages
HIV_raw %>%
  ungroup() %>% 
  miss_var_summary()

```

```{r}
# number of subjects
length(unique(HIV_raw$newid))
```

Calculate the number of insufficient data labels for the comorbidities
```{r}
HIV_raw %>%
  filter(years == 0) %>%
  summarise(
    across(
      c(DIAB, HBP, KID, LIV34, FRP, FP),
      ~ sum(. == 9, na.rm = TRUE),
      .names = "n9_{.col}"))
```

Look at distribution of measurements
```{r}
# measurement stats
HIV_raw %>% 
  group_by(newid) %>%
  summarise(n_measurements = n()) -> mnts_per_id 

# summary stats for number of measurements per subject
mnts_per_id %>%   
  summarise(avg = mean(n_measurements),
            min = min(n_measurements),
            med = median(n_measurements),
            max = max(n_measurements))

# distribution of measurements
hist(mnts_per_id$n_measurements, breaks=11, freq = FALSE)
```

```{r}
# percentage distribution of measurements
mnts_per_id %>%
  dplyr::count(n_measurements) %>%
  mutate(percent = round(100 * n / sum(n) , 1))

# cumulative percentages (e.g. % more than 3 measurements=70.1%)
mnts_per_id %>%
  reframe(
    threshold = 1:9,
    percent_gt = sapply(threshold, function(k) {
      mean(n_measurements > k) * 100
      }))

```

How many observations per race category?
```{r}
HIV_raw %>% 
  group_by(RACE) %>% 
  summarize(n = n())
```

Problems arise when race is:
* Interpreted biologically
* Used as a proxy for genetic ancestry
* Over-interpreted or over-stratified without power

We are not interpreting or estimating race effects, so we should be good to collapse

Potential explanation in final report:
Self-reported race/ethnicity was included as a covariate to account for potential confounding by social determinants of health and healthcare access disparities. Race/ethnicity categories were collected as [describe original categories]. For analyses with limited statistical power due to sample size, we combined categories as follows: [describe any collapsing]. We acknowledge that race is a social construct and that observed differences likely reflect structural inequities rather than biological differences.

Look at distributions of outcomes

```{r}
hist(HIV_raw$VLOAD)
```

```{r}
# number missing for hard drug group
HIV_raw %>% 
  group_by(hard_drugs) %>% 
  summarize(n = n())
```

Spaghetti plots
```{r}
# convert to long format
long_outcomes <- HIV_raw %>%
  filter(years %in% c(0,1,2) & !is.na(hard_drugs)) %>%
  mutate(log_vload = log10(VLOAD)) %>% 
           
  pivot_longer(cols = c(log_vload, LEU3N, AGG_PHYS, AGG_MENT),
               names_to = "outcome",
               values_to = "value") %>% 
    
    mutate(
      outcome = recode(outcome,
                       log_vload = "Viral Load (log10 copies/mL)",
                       LEU3N = "CD4+ T-Cell Count",
                       AGG_PHYS = "Physical QOL (0 to 100)",
                       AGG_MENT = "Mental QOL (0 to 100)"),
      years_f = factor(years, levels = c(0, 1, 2), labels = c("0", "1", "2")))

ggplot(long_outcomes, aes(years, value, group = newid, color = hard_drugs)) +
  geom_line(alpha = 0.05) +
  stat_summary(aes(group = hard_drugs), fun = mean, 
               geom = "line", linewidth = 0.6) +
  facet_wrap(~ outcome, scales = "free_y") +
  scale_color_manual(values = c("steelblue", "firebrick")) +
  labs(title = "Observed Trajectories of Treatment Response",
       x = "Years",
       y="",
       color = "Hard Drugs") +
  theme_bw() +
  theme(legend.position = "bottom") 

```

```{r}
ggplot(long_outcomes, aes(x = years_f, y = value, fill = hard_drugs)) +
  geom_boxplot(outlier.shape = NA, width = 0.7, position = position_dodge(width = 0.75)) +
  geom_point(
    aes(color = hard_drugs),
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
    alpha = 0.08, size = 0.6
  ) +
  facet_wrap(~ outcome, scales = "free_y") +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_color_manual(values = c("steelblue", "firebrick")) +
  labs(
    title = "Treatment Response Distributions by Year and Baseline Hard Drug Use",
    x = "Year since HAART initiation",
    y = NULL,
    fill = "Hard Drugs",
    color = "Hard Drugs"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```




### Table1 Demographics

```{r}
tbl1 <- ana2_cc %>%
  
  tbl_summary(
    by = hard0,
    include = c(mental0, physical0, log_vload0, cd40, age0, bmi0, smoke0, 
                educ0, race0), # add ADH or no?
    type = all_continuous() ~ "continuous2",
    statistic = list(
      all_continuous() ~ c(
        "{mean} ({sd})",
        "{median} ({p25}, {p75})",
        "{min} - {max}"),
      all_categorical() ~ "{n} ({p}%)"),
    
    digits = all_continuous() ~ 1,
    missing = "ifany",
    missing_text = "NA",
    missing_stat = "{N_miss} ({p_miss}%)",
    
    # change variable names
    label = list(
      mental0 ~ "Mental QOL",
      physical0 ~ "Physical QOL",
      log_vload0  ~  "Viral Load (log<sub>10</sub> copies/mL)",
      cd40  ~ "CD4 T-cell Count",
      age0 ~ "Age",
      bmi0 ~ "BMI",
      smoke0 ~ "Smoke",
      educ0 ~ "Education",
      race0 ~ "Race")) %>% 
  
  add_overall() %>%
  bold_labels() %>% 
  italicize_levels() %>% 
  #add_p(pvalue_fun = label_style_pvalue(digits = 2)) %>% 
  add_n() %>% 
  modify_spanning_header(
    c("stat_1", "stat_2") ~ "**Hard Druge Use**") %>%
  modify_footnote_header(
    footnote = "Table includes all subjects from cohort at baseline (years=0)",
    columns = all_stat_cols(),
    replace = FALSE) %>% 
  
  as_gt() %>%
  gt::fmt_markdown(columns = everything()) %>% 
  gt::fmt_number(
    columns = dplyr::starts_with("p.value"),
    decimals = 2) %>% 
  tab_options(
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))


tbl1

```

log10 interpretations 
* 10^n = where n is number of places after 1
* 10^8 ≈ in hundreds of millions
* 10^5 ≈ hundreds of thousands
* 10^4 ≈ tens of thousands
* 10^0.5 ≈ 3.16 - in single digits - remember exponent of 0 is 1

Create a table 2 to look at univariate relationships between hard drug use and comorbidities

```{r}
tbl2 <- ana2 %>%
  
  tbl_summary(
    by = hard0,
    include = c(hbp0, diab0, liv34_0, kid0),
    type = all_continuous() ~ "continuous2",
    statistic = list(
      all_continuous() ~ c(
        "{mean} ({sd})",
        "{median} ({p25}, {p75})",
        "{min} - {max}"),
      all_categorical() ~ "{p}%"),
    
    digits = all_continuous() ~ 1,
    missing = "ifany",
    missing_text = "NA",
    missing_stat = "{p_miss}%",
    
    # change variable names
    label = list(
      hbp0 ~ "High Bloood Pressure",
      diab0 ~ "Diabetes",
      liv34_0  ~  "Liver Disease (Stage 3/4)",
      kid0  ~ "Kidney Disease")) %>%
  
  add_overall() %>%
  bold_labels() %>% 
  italicize_levels() %>% 
  add_p(pvalue_fun = label_style_pvalue(digits = 2)) %>% 
  modify_spanning_header(
    c("stat_1", "stat_2") ~ "**Hard Druge Use**") %>%
  modify_footnote_header(
    footnote = "Table includes all subjects from cohort at baseline (years=0)",
    columns = all_stat_cols(),
    replace = FALSE) %>% 
  
  as_gt() %>%
  gt::fmt_markdown(columns = everything()) %>% 
  gt::fmt_number(
    columns = dplyr::starts_with("p.value"),
    decimals = 2) %>% 
  tab_options(
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

tbl2
```

* Rates of liver and kidney disease are higher in drug users (3.2% vs 9.7% and 16% vs 34%, respectively). No difference in diabetes and higher blood pressure is more prevalent in
the non-drug use group (27% vs 15%). Also, there is pretty substantial missing values
for the kidney (71%), diabetes (57%), and liver (34%) groups. 


## Primary Analysis
* How treatment response differs between hard drug users and non-users

* Collider bias? - do these comorbidities occur after exposure and outcome?
 - The exposure and outcome can both independently create a collider variable. In contrast to the case of confounding, where the confounder is associated with both the exposure and outcome, the collider is created by both the exposure and outcome.
 - Comorbidities can be downstream effects of higher viral load/cd4 and hard drug use,
 independently. Therefore, these comorbidities could very well be colliders
 

Streamline method utilizing functions to fit all models at once
```{r}
outcomes <- list(
  cd4   = list(y = "cd42", baseline = "cd40"),
  vload = list(y = "log_vload2", baseline = "log_vload0"),
  phys  = list(y = "physical2", baseline = "physical0"),
  ment  = list(y = "mental2", baseline = "mental0"))
```

```{r label="Function to perform all primary models at once"}
fit_models <- function(outcome, baseline, data) {
  # model form - consistent across all models
  full <- as.formula(
    paste(outcome, "~ hard0 +", baseline,
          "+ age0 + bmi0 + smoke0 + educ0 + race0"))
  # with hard drug coeffecient removed
  reduced <-  as.formula(
    paste(outcome, "~", baseline,
          "+ age0 + bmi0 + smoke0 + educ0 + race0"))
  
  # Frequentist model
  lm_fit <- lm(full, data = data)
  
  # Bayes model
  options(warn = -1)
  brm_fit <- brm(
    full,
    data = data,
    family = gaussian(),
    prior = c(
      # ***the priors need to be more uniform or more robust to different models**
      prior(normal(0, 10000), class = "Intercept"), # less interpretable so make var wider
      prior(normal(0, 10000), class = "b"),
      prior(student_t(3, 0, 100^2), class = "sigma")), # half-t - standard
    chains = 4, cores = 6, iter = 2000,
    refresh = 0)
  
  # WAIC
  log_lik_fit <- log_lik(brm_fit)
  
  # Bayes reduced model for testing hard drugs significance
    brm_red <- brm(
    reduced,
    data = data,
    family = gaussian(),
    prior = c(
      prior(normal(0, 10000), class = "Intercept"), 
      prior(normal(0, 10000), class = "b"),
      prior(student_t(3, 0, 1000), class = "sigma")), 
    chains = 4, cores = 6, iter = 2000,
    refresh = 0)
    options(warn = 0)
    
    # WAIC
    log_lik_red <- log_lik(brm_red)

    # leave-one-out cross-validation for reduced model
    loo_res <- suppressWarnings(brms::loo_compare(
      brms::add_criterion(brm_fit, "loo"),
      brms::add_criterion(brm_red, "loo"),
      model_names = c("Full Model", "Reduced Model")))
    
    waic_fit <- waic(log_lik_fit)
    waic_red <- waic(log_lik_red)
    waic_diff <- loo_compare(waic_fit, waic_red)
  
  # store model results into a list
  list(lm = lm_fit, brm = brm_fit, brm_red = brm_red,
      loo_res=loo_res, ll_fit=log_lik_fit, ll_red=log_lik_red,
       waic_fit=waic_fit, waic_red=waic_red, waic_diff=waic_diff)
}
```

Run all models at once
```{r}
t0 <- Sys.time()

model.fits <- imap(outcomes, ~ fit_models(.x$y, .x$baseline, ana2_cc))

t1 <- Sys.time()
message("Runtime: ",
        round(as.numeric(difftime(t1, t0, units = "mins")), 2),
        " minutes")
```

* Macbook Pro M1 (6-cores) Runtime: 5.58min

```{r}
# extract info from function list using CD4 outcome as example
model.fits$cd4$lm
model.fits$cd4$brm
model.fits$cd4$loo_res
report(model.fits$cd4$loo_res)
model.fits$cd4$waic_fit
model.fits$cd4$waic_red
model.fits$cd4$waic_diff

# loo results
elpd_diff <- model.fits$cd4$loo_res["Reduced Model", "elpd_diff"]
se_diff <- model.fits$cd4$loo_res["Reduced Model", "se_diff"]

# confirm report() function p-value (wald-normal approx)
z_score <- abs(elpd_diff) / se_diff
(p_value <- 2 * (1 - pnorm(z_score)))



```

"The rule of thumb is that the models are "very similar" if |elpd_diff| (the absolute value of elpd_diff) is less than 4 (Sivula, Magnusson and Vehtari, 2020). If superior to 4, then one can use the SE to obtain a standardized difference (Z-diff) and interpret it as such, assuming that the difference is normally distributed. The corresponding p-value is then calculated as `2 * pnorm(-abs(Z-diff))`. However, note that if the raw ELPD difference is small (less than 4), it doesn't make much sense to rely on its standardized value: it is not very useful to conclude that a model is much better than another if both models make very similar predictions."

Extract key effects for table output function
```{r label="Function for extracting key elements from models"}
extract_effects <- function(lm_fit, brm_fit, outcome_name, waic_fit, waic_red) {

  # Frequentist
  freq <- broom::tidy(lm_fit, conf.int = TRUE) %>%
    filter(term == "hard0Yes") %>%
    transmute(
      outcome = outcome_name,
      freq_est = estimate,
      freq_lo = conf.low,
      freq_hi = conf.high,
      freq_p = p.value)

  # Bayesian
  draws <- posterior::as_draws_df(brm_fit)
  beta <- draws$b_hard0Yes

  bay <- tibble(
    outcome = outcome_name,
    bay_est = mean(beta),
    bay_lo = quantile(beta, 0.025),
    bay_hi = quantile(beta, 0.975))
    
  # Extract WAIC values 
  elpd_full <- waic_fit$estimates["elpd_waic", "Estimate"]
  elpd_full_se <- waic_fit$estimates["elpd_waic", "SE"]
  waic_full <- waic_fit$estimates["waic", "Estimate"]
  waic_full_se <- waic_fit$estimates["waic", "SE"]

  elpd_red  <- waic_red$estimates["elpd_waic", "Estimate"]
  elpd_red_se <- waic_red$estimates["elpd_waic", "SE"]
  waic_reduced<- waic_red$estimates["waic", "Estimate"]
  waic_reduced_se <- waic_red$estimates["waic", "SE"]

  waic_comp <- tibble::tibble(
    outcome = outcome_name,
    elpd_full = elpd_full,
    elpd_full_se = elpd_full_se,
    waic_full = waic_full,
    waic_full_se = waic_full_se, 
    
    elpd_red = elpd_red,
    elpd_red_se = elpd_red_se,
    waic_reduced = waic_reduced,
    waic_reduced_se = waic_reduced_se,
    
    elpd_diff = elpd_full - elpd_red,
    waic_diff = waic_full - waic_reduced
  )

  # combine all tables
  result <- freq %>%
    dplyr::left_join(bay, by = "outcome") %>%
    dplyr::left_join(waic_comp, by = "outcome")

}
```

Comparison table
```{r}
results <- imap_dfr(model.fits, ~ extract_effects(
  lm_fit = .x$lm, 
  brm_fit = .x$brm, 
  outcome_name = .y, 
  waic_fit = .x$waic_fit,
  waic_red = .x$waic_red))


results %>% 
  filter(outcome = vload)
```

```{r}
# add sample sizes from each model and back-transform log10
results2 <- results %>%
  mutate(
    n_lm  = map_int(names(model.fits), 
                    ~ nobs(model.fits[[.x]]$lm)),
    n_brm = map_int(names(model.fits), 
                    ~ nrow(model.frame(model.fits[[.x]]$brm))),
    across(
      c(freq_est, freq_lo, freq_hi, bay_est, bay_lo, bay_hi),
      ~ if_else(outcome == "vload", 1-10^.x, .x)),
    outcome = recode(outcome,
                     cd4   = "CD4",
                     vload = "Viral load (log10)",
                     phys  = "Physical QOL",
                     ment  = "Mental QOL"))
```


### Results Table

```{r}
tbl_compare <- results2 %>%
  # combine LL and UL CIs
  mutate(
    freq_ci = sprintf("(%.2f, %.2f)", freq_lo, freq_hi),
    bay_ci  = sprintf("(%.2f, %.2f)", bay_lo, bay_hi)) %>%
  select(outcome, n_lm, n_brm,
         freq_est, freq_ci, freq_p,
         bay_est, bay_ci, waic_full, waic_reduced, waic_diff) %>%
  
  gt(rowname_col = "outcome") %>%
  cols_label(
    n_lm = "N",
    n_brm = "N",
    freq_est = "Estimate",
    freq_ci = "95% CI",
    freq_p   = "P-value",
    bay_est  = "Mean",
    bay_ci = "95% HDI",
    waic_full = "WAIC Full",
    waic_reduced = "WAIC Red",
    waic_diff = "WAIC Diff"
    ) %>%
  # header for freq
  tab_spanner(
    label = "Frequentist",
    columns = c(n_lm, freq_est, freq_ci, freq_p)) %>%
  # header for Bayes
  tab_spanner(
    label = "Bayesian",
    columns = c(n_brm, bay_est, bay_ci, waic_full,waic_reduced,waic_diff)) %>%
  # format numbers 
  fmt_number(columns = c(freq_est, freq_ci, bay_est, bay_ci), decimals = 2) %>%
  fmt_number(columns = c(waic_full, waic_reduced, waic_diff), decimals = 1,
             use_seps = F) %>%
  fmt_number(columns = freq_p, decimals = 3) %>%
  
  # change p-value reporting style
  text_transform(
    locations = cells_body(columns = freq_p),
    fn = function(x) ifelse(as.numeric(x) < 0.001, "<0.001",
                            format(round(as.numeric(x), 3), nsmall = 3))) %>% 
  # bold specific text
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_stub(rows = everything())) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_spanners(
      spanners = c("Frequentist", "Bayesian"))) %>% 
  # center column labels
  tab_style(
  style = cell_text(align = "center"),
  locations = cells_column_labels(everything())) %>% 
  # size of font
  tab_options(
    table.font.size = "small") %>% 
  # add a footnotes
  tab_footnote(
    footnote = "Frequentist intervals are 95% confidence intervals; Bayesian intervals are 95% credible intervals.",
    locations = cells_column_labels(columns = c(freq_ci, bay_ci))) %>% 
  tab_footnote(
    footnote = "For WAIC, lower values indicate better model fit.",
    locations = cells_column_labels(columns = c(waic_full, waic_reduced, waic_diff)))

tbl_compare
```

After adjusting for baseline viral load and covariates, hard-drug users have about 0.37× the viral load of non-users at Year 2 (≈63% lower).

```{r}
nobs(model.fits$cd4$lm)
nrow(model.frame(model.fits$cd4$brm))  # or nrow(model.frame(cd4_brm))
```


 

### Viral Load as outcome (N=464):
```{r label = "frequentist model"}
#options(scipen = 999)

vload.model <- lm(log_vload2 ~ factor(hard0) + log_vload0 + age0 + bmi0 + factor(smoke0) + 
                       factor(educ0) + factor(race0),
              data=ana2_cc)

summary(vload.model)
nobs(vload.model)

tidy(vload.model, conf.int = TRUE, exponentiate = FALSE) %>% 
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = round(p.value, 3),
    conf.low = round(conf.low, 3),
    conf.high = round(conf.high, 3),
    vload_copies_ml = round(10^estimate, 3),
    vload_ml_LL = round(10^conf.low, 3),
    vload_ml_UL = round(10^conf.high, 3))
```

After adjusting for baseline viral load, age, BMI, smoking, education, and race, baseline hard-drug users had an estimated 8.2% lower viral load at Year 2 compared with non-users (1-0.918), but this difference did not reach significance (p = 0.86).

* We are comparing hard-drug users and non-users while holding the baseline viral load, age, BMI, smoking status, education, and race values constant. In other words, after other affects have been accounted for.
* The model estimates differences in Year-2 viral load between groups after removing variation attributable to baseline viral load and covariates, effectively comparing individuals at the same starting level.

Test to see if comorbidities can be used as precision variables or may be
on the casual pathway from drug use to viral load - **ignore for primary analysis**
```{r}
options(scipen = 999)

vload.model2 <- lm(log_vload2 ~ 
                       # comorbidites
                       factor(hbp0) + factor(diab0) + factor(liv34_0) + factor(kid0) +
                       # Primary
                       factor(hard0) + log_vload0 + age0 + bmi0 + 
                       factor(smoke0) + factor(educ0) + factor(race0),
              data=ana2)

summary(vload.model2)
nobs(vload.model2)

tidy(vload.model2, conf.int = TRUE, exponentiate = FALSE) %>% 
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = round(p.value, 3),
    conf.low = round(conf.low, 3),
    conf.high = round(conf.high, 3),
    vload_copies_ml = round(10^estimate, 3),
    vload_ml_LL = round(10^conf.low, 3),
    vload_ml_UL = round(10^conf.high, 3))
```

Look at multicollinearity issues if present
```{r}
car::vif(vload.model2)
```

* hard drug use completely changes direction when including comorbidites. It now has an
estimated 122.3% higher viral load as compared to non-drug users

Bayesian (N=464)

```{r label = "Bayes model"}
vload.brm <- brm(
  log_vload2 ~ hard0 + log_vload0 + age0 + bmi0 + smoke0 + educ0 + race0,
  data = ana2_cc,
  family = gaussian(),
  prior = c(
    # slope of 1 on log10 is 10-fold difference in viral load (weakly informative)
    prior(normal(0, 1), class = "b"), 
    prior(normal(0, 5), class = "Intercept"),  # less interpretable so make var wider
    prior(student_t(3, 0, 2.5), class = "sigma")), # residual SD (half-t)
  chains = 4, cores = 6, iter = 4000)

summary(vload.brm)

```

```{r}
vload_post1 <- posterior_summary(vload.brm) %>%
  as.data.frame() %>%
  rownames_to_column(var = "parameter")

vload_post1

vload_post2 <- describe_posterior(vload.brm, ci_method = "HDI", effects = "all")
vload_post2

vload_draws <- posterior::as_draws_df(vload.brm)
vload_beta  <- vload_draws$b_hard0Yes

# Directional evidence (hard drug users have higher log10 viral load)
mean(vload_beta > 0)

## limited evidence from labs in vitro and animal studies show hard drugs inhibits 
## immune response and increases HIV replication. PIs believe TRT should lower viral load
## and increase CD4 counts

# alternative for viral load
mean(vload_beta < 0)

# 95% credible interval
quantile(vload_beta, c(0.025, 0.975))
```
* The posterior mean difference in Year-2 log10 viral load for hard drug users vs non-users was -0.009 (95% CrI: -0.41, 0.39), with posterior probability $P(\beta>0)=0.97$


### CD4 outcome model (N=464):
* Did not transform as residuals looked good with raw values.
```{r}
options(scipen = 999)

cd4.model <- lm(cd42 ~ factor(hard0) + cd40 + age0 + bmi0 + factor(smoke0) + 
                       factor(educ0) + factor(race0),
              data=ana2_cc)

summary(cd4.model)
nobs(cd4.model)

tidy(cd4.model, conf.int = TRUE, exponentiate = FALSE) %>% 
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = round(p.value, 3),
    conf.low = round(conf.low, 3),
    conf.high = round(conf.high, 3))
```

After adjusting for baseline baseline CD4+ T cell counts, age, BMI, smoking, education, and race, baseline hard-drug users had an estimated lower CD4 count of approximately 172.2 at Year 2 compared with non-users [95% CI: (-235 to -109.4); p<0.001].

```{r}
cd4.brm <- brm(
  cd42 ~ hard0 + log_vload0 + age0 + bmi0 + smoke0 + educ0 + race0,
  data = ana2_cc,
  family = gaussian(),
  prior = c(
    # slope of 1 on log10 is 10-fold difference in viral load (weakly informative)
    prior(normal(0, 1), class = "b"), 
    prior(normal(0, 5), class = "Intercept"),  # less interpretable so make var wider
    prior(student_t(3, 0, 2.5), class = "sigma")), # residual SD (half-t)
  chains = 4, cores = 6, iter = 4000)

summary(cd4.brm)
```



Identify why I'm getting contrast error
```{r}
table(ana2$hard0, useNA = "ifany")
table(ana2$smoke0, useNA = "ifany")
table(ana2$educ0, useNA = "ifany")
table(ana2$race0, useNA = "ifany")
```


Look at distribution of CD4 T cell count, log counts, and boxplots between drug user group
```{r}
hist(ana2$cd40)
hist(ana2$log_cd40)
boxplot(ana2$log_cd40~ana2$hard0)
boxplot(ana2$cd40~ana2$hard0)
```

Check that residuals look good using raw CD4 counts
```{r}
# check residuals to determine if skewness is too extreme for this model approach
plot(cd4.model)   # residuals vs fitted

```

* some outliers but overall residuals look good with raw CD4 values


Check distribution of QOL measures to identify odd values
```{r}
hist(ana2$physical0)
summary(ana2$physical0)
hist(ana2$mental0)
summary(ana2$mental0)

```



Build Mental QOL outcome model (frequentist)
```{r}
options(scipen = 999)

mental.model <- lm(mental2 ~ factor(hard0) + mental0 + age0 + bmi0 + factor(smoke0) + 
                       factor(educ0) + factor(race0),
              data=ana2)

summary(mental.model)

tidy(mental.model, conf.int = TRUE, exponentiate = FALSE) %>% 
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = round(p.value, 3),
    conf.low = round(conf.low, 3),
    conf.high = round(conf.high, 3))
```


Build Physical QOL outcome model (frequentist)
```{r}
options(scipen = 999)

physical.model <- lm(physical2 ~ factor(hard0) + physical0 + age0 + bmi0 + factor(smoke0) + 
                       factor(educ0) + factor(race0),
              data=ana2)

summary(physical.model)

tidy(physical.model, conf.int = TRUE, exponentiate = FALSE) %>% 
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = round(p.value, 3),
    conf.low = round(conf.low, 3),
    conf.high = round(conf.high, 3))
```

## Mediation Analysis

Custom function for calculating mediation coefficients
```{r}
# boot_fun <- function(data, idx) {
#   d <- data[idx, ]
#   m <- glm(adh_bin ~ hard0 + age0 + bmi0 + smoke0 + educ0 + race0, 
#            family = binomial,
#            data = d)
#   y <- lm(cd42 ~ hard0 + adh_bin + cd40 +
#             age0 + bmi0 + smoke0 + educ0 + race0, data = d)
# 
#   a <- coef(m)["hard0"]     
#   b <- coef(y)["adh2"]
#   cprime <- coef(y)["hard0"]
# 
#   c(indirect = a*b, direct = cprime, total = cprime + a*b)
# }
# 
# set.seed(1)
# bt <- boot(ana2_cc, statistic = boot_fun, R = 2000)
# 
# boot.ci(bt, type = "perc", index = 1)  # indirect
# boot.ci(bt, type = "perc", index = 2)  # direct
# boot.ci(bt, type = "perc", index = 3)  # total
```


Pre-lim mediation analysis w/out Covs
```{r}
# Fit our three models
mod2_crude <- lm(cd42 ~ hard0, data=ana2_cc) # Y~X
mod2_adjusted <- lm(cd42 ~ hard0 + adh_bin, data=ana2_cc) # Y~ X+M
mod2_covariate <- glm(adh_bin ~ hard0, 
                     family = binomial,
                     data=ana2_cc) # M~X

summary(mod2_covariate)

# Extract coefficients
coef(mod2_crude)
coef(mod2_adjusted)
coef(mod2_covariate)
```


DAG
```{tikz, tikz-med, fig.ext = 'png', cache=TRUE, echo=F, fig.align="center", fig.width=4}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=4cm, auto,>=latex, thick, scale = 0.5]
\node (X) [align=center] {Hard Drug Use \\ ($X$)};
\node (M) [align=center,above of=X,xshift=2cm,yshift=-2cm] {Adherence \\ ($M$)};
\node (Y) [align=center,right of=X] {Year 2 CD4 \\ ($Y$)};
\draw[->] (X) to node [above left] {$\hat{\gamma}_{X}=1.59$} (M);
\draw[->] (M) to node [above right] {$\hat{\beta}_{M}=98.03$} (Y);
\draw[->] (X) to node [align=center,swap] {$\hat{\beta}_{adj}=-202.55$ \\ $\hat{\beta}_{crude}=-193.91$} (Y);
\end{tikzpicture}
```

Using mediation package
```{r}
# Mediator model: adherence as outcome
m_mod <- glm(adh_bin ~ hard0 + age0 + bmi0 + smoke0 + educ0 + race0, 
             family = binomial,
            data = ana2_cc)

# Outcome model: Year-2 viral load, adjusting for baseline viral load + adherence
y_mod <- lm(cd42 ~ hard0 + adh_bin + cd40 +
              age0 + bmi0 + smoke0 + educ0 + race0, 
            data = ana2_cc)

# Mediation estimation (bootstrap for inference)
med_vl <- mediate(m_mod, y_mod, treat = "hard0", mediator = "adh_bin",
                  boot = F, sims = 2000)

summary(med_vl)

```


Bayesian mediation
```{r}
# Mediator model
med_model <- bf(adh_bin ~ hard0 + cd40 + age0 + bmi0 + smoke0 + educ0 + race0,
                family = bernoulli(link = "logit"))

# Outcome model
out_model <- bf(cd42 ~ hard0 + adh_bin + cd40 +
                age0 + bmi0 + smoke0 + educ0 + race0,
                family = gaussian())

med.brm.fit <- brm(med_model + out_model + set_rescor(FALSE),
                   chains = 4, cores = 6, iter = 2000,
           data = ana2_cc)
```

```{r}
table(ana2_cc$adh_bin, useNA="ifany")
```

```{r}
summary(med.brm.fit)
```


Extract the draws

```{r}
# extract posterior draws 
med_draws <- as_draws_df(med.brm.fit)

# get parameter names
names(med_draws)[grepl("^b_", names(med_draws))]
```

### Diagnsotics for Bayesian Mediation Model

```{r, eval=F}
# diagnostics
mcmc_areas(med_draws,prob = 0.95,
             pars = c("b_adhbin_hard0Yes"))
```

```{r, eval=F}
# Trace plots for each parameter
mcmc_trace(post_samples,
           pars = c("b_adhbin_hard0Yes", "b_cd42_hard0Yes"))

```

```{r}
levels(ana2_cc$hard0)
names(med_draws)[grepl("^b_", names(med_draws))]
```

```{r}
## X = drug use & M = adherence

## We are using complete-case analysis so use ana2_cc dataset

# make sure hard0 has the right reference 
ana2_cc$hard0 <- relevel(factor(ana2_cc$hard0), ref = "No")

# mediator levels - current reference is <95%
levM <- levels(ana2_cc$adh_bin)
m0 <- levels(ana2_cc$adh_bin)[1] # reference
m1 <- levels(ana2_cc$adh_bin)[2]      

# create new dataset with X set to No and Yes
hd_NO <- ana2_cc
hd_NO$hard0 <- factor("No", levels = levels(ana2_cc$hard0))

hd_YES <- ana2_cc
hd_YES$hard0 <- factor("Yes", levels = levels(ana2_cc$hard0))

# Posterior expected adherence probabilities P(M=1|X,C)
# For bern, posterior_epred gives probability of success (i.e., P(M == m1))
pM0 <- posterior_epred(med.brm.fit, resp = "adhbin", newdata = hd_NO)  
pM1 <- posterior_epred(med.brm.fit, resp = "adhbin", newdata = hd_YES)  

# Posterior expected CD4 means for each X and each fixed mediator level
# Build newdata that sets adh_bin to m0 or m1
hd_NO_lt95 <- hd_NO; hd_NO_lt95$adh_bin <- factor(m0, levels = levM)
hd_NO_gt95 <- hd_NO; hd_NO_gt95$adh_bin <- factor(m1, levels = levM)

hd_YES_lt95 <- hd_YES; hd_YES_lt95$adh_bin <- factor(m0, levels = levM)
hd_YES_gt95 <- hd_YES; hd_YES_gt95$adh_bin <- factor(m1, levels = levM)

# expected CD4 if No hard drugs w/ <95% ADH + Covs
mu0_m0 <- posterior_epred(med.brm.fit, resp = "cd42", newdata = hd_NO_lt95)  
# expected CD4 if No hard drugs w/ ≥95% ADH + Covs
mu0_m1 <- posterior_epred(med.brm.fit, resp = "cd42", newdata = hd_NO_gt95)  
# expected CD4 if Yes hard drugs w/ <95% ADH + Covs
mu1_m0 <- posterior_epred(med.brm.fit, resp = "cd42", newdata = hd_YES_lt95)  
# expected CD4 if Yes hard drugs w/ ≥95% ADH + Covs
mu1_m1 <- posterior_epred(med.brm.fit, resp = "cd42", newdata = hd_YES_gt95)  

## Standardize over the mediator distribution:
 # 1 - pM0 = prob of low ADH ; pM0 = prob of high ADH | No hard drug use
 # 1 - pM1 = prob of low ADH ; pM1 = prob of high ADH | Yes hard drug use
EY_0M0 <- rowMeans(mu0_m0 * (1 - pM0) + mu0_m1 * pM0) # pop-level E[CD4|X=0]
EY_1M1 <- rowMeans(mu1_m0 * (1 - pM1) + mu1_m1 * pM1) # pop-level E[CD4|X=1]

# expectation of CD4 if a hard drug user but ADH behaved as if they were NOT
# a drug user
EY_1M0 <- rowMeans(mu1_m0 * (1 - pM0) + mu1_m1 * pM0)

## CD4 depends on adherence.
## ADH depends on drug use.
## So to understand mediation, we must avg CD4 preds over how often people are ADH

# calculate the effects
TE  <- EY_1M1 - EY_0M0
NIE <- EY_1M1 - EY_1M0
NDE <- EY_1M0 - EY_0M0

# summarize
summ <- function(x) {
  c(
    mean = mean(x),
    hdi_low = hdi(x, prob = 0.95)[1],
    hdi_high = hdi(x, prob = 0.95)[2],
    Pr_gt_0 = mean(x > 0))
  }

rbind(TE = summ(TE), NIE = summ(NIE), NDE = summ(NDE))
```

Total Effect (TE)
Interpretation:
Moving from No hard drug use to Yes hard drug use is associated with an average decrease of about 168 CD4 cells at Year 2.

Indirect Effect (NIE-mediation through adherence)
Interpretation:
There is strong posterior evidence that adherence mediates part of the association.

Direct Effect (NDE)
Interpretation:
Even if adherence were held constant, hard drug use would still be associated with a ~174 cell lower CD4 count.

This suggests that most of the effect is not explained by adherence

Proportion Mediated
mean(NIE / TE) = 6/168 ≈ 3.6%


## Dropout Rate Analysis

```{r}
ana2_dr <- ana2 %>% 
  select(log_vload2, log_vload0, cd42, cd40, hard0, adh2, physical0, physical2,
         mental0, mental2, age0, bmi0, smoke0, educ0, race0) %>% 
  mutate(complete_obs = if_all(everything(), ~ !is.na(.)))

table(ana2_dr$complete_obs, ana2_dr$hard0)
t.test(cd40 ~ complete_obs, data = ana2_dr)

```


Generate a table that shows complete case analysis vs using all 715 observations
* in other words 464 vs 715
* since we are using complete case to be consistent across all analyses - I am comparing complete cases of 464 vs using all 715 subjects who has baseline or year 2. I reason this is the best approach??
```{r}
ana2_drop <- ana2_dr %>%
  mutate(
    hard0 = factor(hard0, levels = c("No", "Yes"), exclude = NULL),
    followup_status = factor(
      if_else(complete_obs == T, "Complete", "Missing"),
      levels = c("Complete", "Missing")))

tbl_dropout <- ana2_drop %>%
  tbl_summary(
    by = followup_status,
    include = c(log_vload0, cd40, physical0, mental0),
    type = list(
    all_continuous() ~ "continuous2"),
    statistic = list(
      all_continuous() ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
      all_categorical() ~ "{n} ({p}%)"),
    digits = all_continuous() ~ 1,
    missing = "ifany",
    missing_text = "Missing Both Measurements",
    missing_stat = "{N_miss} ({p_miss}%)",
    label = list(
      log_vload0 ~ "Baseline viral load (log10 copies/mL)",
      cd40 ~ "Baseline CD4 T-cell Count",
      physical0 ~ "Baseline physical QOL",
      mental0 ~ "Baseline mental QOL")) %>%
  add_p(pvalue_fun = ~style_pvalue(.x, digits = 3)) %>%
  add_n() %>%
  add_overall() %>%
  bold_labels() %>%
    modify_spanning_header(
    c("stat_1", "stat_2") ~ "**Cases**") %>%
  modify_footnote_header(
    columns = all_stat_cols(),
    footnote = "Follow-up status defined based on availability of Year 2 outcome.",
    replace = F) %>%
  as_gt() %>%
  tab_options(
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

tbl_dropout
```


Function to create a table for each outcome to compare summary statistics at baseline to determine if dropout may bias results. - **Needs work**

```{r label="dropout table function"}
make_dropout_tbl <- function(data, var, label_text) {
  stopifnot(is.data.frame(data))
  stopifnot(is.character(var), length(var) == 1)
  stopifnot(is.character(label_text), length(label_text) == 1)

  # header counts + %
  hdr_df <- data %>%
    dplyr::count(followup_status, name = "n") %>%
    dplyr::mutate(
      pct = n / sum(n),
      hdr = paste0(
        "**", followup_status, "**<br>N = ", n,
        " (", scales::percent(pct, accuracy = 0.1), ")"))

  hdr_list <- setNames(as.list(hdr_df$hdr), paste0("stat_", seq_len(nrow(hdr_df))))

  # label as a formula: var ~ "Label"
  lab_fml <- rlang::new_formula(rlang::sym(var), label_text)

  data %>%
    gtsummary::tbl_summary(
      by = followup_status,
      include = dplyr::all_of(var),
      type = list(gtsummary::all_continuous() ~ "continuous2"),
      statistic = list(
        gtsummary::all_continuous() ~ c("{mean} ({sd})", "{median} ({p25}, {p75})")
      ),
      digits = gtsummary::all_continuous() ~ 1,
      missing = "ifany",
      missing_text = "Missing Both Measurements",
      missing_stat = "{N_miss} ({p_miss}%)",
      label = lab_fml
    ) %>%
    gtsummary::add_p(pvalue_fun = ~gtsummary::style_pvalue(.x, digits = 3)) %>%
    gtsummary::add_n() %>%
    gtsummary::add_overall() %>%
    gtsummary::bold_labels() %>%
    gtsummary::modify_spanning_header(c("stat_1", "stat_2") ~ "**Cases**") %>%
    gtsummary::modify_header(!!!hdr_list) %>% 
    as_gt() %>%
    tab_options(
      table.font.size = "small",
      heading.align = "left",
      data_row.padding = px(3),
      table.border.top.width = px(2),
      table.border.bottom.width = px(2))
}
```

Test function
```{r}
vars <- list(
  log_vload0 = "Baseline viral load (log10 copies/mL)",
  cd40 = "Baseline CD4 T-cell Count",
  physical0 = "Baseline physical QOL",
  mental0 = "Baseline mental QOL")


tables <- purrr::imap(vars, ~ make_dropout_tbl(ana2_drop, var = .y, label_text = .x))
tables$log_vload0
```


Confirmation check that gtsummary() is calculating numbers correctly
```{r}
pairs <- tribble(
  ~outcome, ~base, ~yr2,
  "viral load", "log_vload0", "log_vload2",
  "CD4", "cd40", "cd42",
  "physical QOL", "physical0", "physical2",
  "mental QOL", "mental0", "mental2")

miss_tbl <- pairs %>%
  rowwise() %>%
  mutate(
    miss0_have2 = sum(is.na(ana2[[base]]) & !is.na(ana2[[yr2]])),
    have0_miss2 = sum(!is.na(ana2[[base]]) & is.na(ana2[[yr2]])),
    both = sum(!is.na(ana2[[base]]) & !is.na(ana2[[yr2]])),
    none = sum(is.na(ana2[[base]]) & is.na(ana2[[yr2]])),
    perc_miss = 1-both/715) %>%
  ungroup()

miss_tbl
```

Using nanier package to calculate missing based ana dataset where each row is one subject
```{r}
ana2_drop %>%
  ungroup() %>% 
  miss_var_summary()
```

### Logistic Dropout Modelling
```{r}
dropout.model <- glm(
  followup_status ~ age0 + bmi0 + race0 + smoke0 + educ0 +
            log_vload0 + cd40 + physical0 + mental0 + hard0,
  family = binomial,
  data = ana2_drop)
summary(dropout.model)
broom::tidy(dropout.model, exponentiate = TRUE, conf.int = TRUE)
```
* about 3.8% (1-0.962) lower odds of dropout per unit increase in physical QOL
* about 2.5% (1-0.975) lower odds of dropout per unit increase in age



